---
title: "Cabezon sdmTMB — Primary Analysis"
author: "Dan Portnof"
date: "2026-02-25"
output: word_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = F,
  message = F,
  warning = F
  )
```


---

# PART I — FULL DATASET ANALYSIS

---

# 0. SETUP & STAR REPORT

```{r 0. SETUP & STAR REPORT}
# ---- 0. SETUP & STAR REPORT

# Load packages
library(tidyverse)
library(r4ss)
library(here)
library(dplyr)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(maps)
library(sdmTMB)
library(fmesher)
library(segmented)

my.seed <- 666
set.seed(my.seed)

wd <- "Cab_SCS_BC_STAR" # Mac

# This reads the output files
pp <- SS_output(wd)

# Pull out derived quantities.  These are things like SSB and age-0 recruits.
derived_quants <- pp[["derived_quants"]]

# Pull out parameter estimates.  Recruitment deviations are estimated parameters.  
params <- pp[["parameters"]]

# Plot recruitment deviations for the main estimation period.
devyrs <- 1970:2018
devdf <- pp$parameters[28:76, ]
devvalue <- devdf[,3]
devsd <- devdf[,11]
devlower <- (devvalue-1.96*devsd)
devmed   <- devvalue
devupper <- (devvalue+1.96*devsd)
STAR_recdevs <- data.frame(year=devyrs, value=devmed,lo=devlower,hi=devupper)

STAR_recdevs_plot <- ggplot(data=STAR_recdevs, aes(x=year, ymin=devlower, ymax=devupper)) +
  geom_ribbon(fill="skyblue", alpha=0.5) +
  geom_line(aes(y=devmed)) +
  labs(title="Cabezon SoCal Recruitment Deviations") +
  theme_bw()

STAR_recdevs_plot

# Plot age-0 recruits
age0yrs <- 1970:2018
startrowindex <- which(rownames(derived_quants)=="Recr_1970")
endrowindex <- which(rownames(derived_quants)=="Recr_2018")
age0 <- derived_quants[174:222, "Value"]
age0sd <- derived_quants[174:222, "StdDev"]
age0lower <- (age0-1.96*age0sd)
age0upper <- (age0+1.96*age0sd)

STAR_age0 <- data.frame(year=age0yrs, value=age0,lo=age0lower,hi=age0upper)

STAR_age0_plot <- ggplot(data=STAR_age0, aes(x=year, ymin=age0lower, ymax=age0upper)) +
  geom_ribbon(fill="skyblue", alpha=0.5) +
  geom_line(aes(y=age0)) +
  labs(title="Cabezon SoCal Age 0 Recruits") +
  theme_bw()

STAR_age0_plot

# Plot SSB
ssbyrs <- 1916:2018
startrowindex <- which(rownames(derived_quants)=="SSB_1916")
endrowindex <- which(rownames(derived_quants)=="SSB_2018")
ssb <- derived_quants[3:105, "Value"]
ssbsd <- derived_quants[3:105, "StdDev"]
ssblower <- (ssb-1.96*ssbsd)
ssbupper <- (ssb+1.96*ssbsd)

STAR_SSB <- data.frame(year=ssbyrs, value=ssb,lo=ssblower,hi=ssbupper)

STAR_SSB_plot <- ggplot(data=STAR_SSB, aes(x=year, ymin=ssblower, ymax=ssbupper)) +
  geom_ribbon(fill="skyblue", alpha=0.5) +
  geom_line(aes(y=ssb)) +
  labs(title="Cabezon SoCal SSB") +
  theme_bw()

STAR_SSB_plot

```

This code from Julia gives us the STAR derived quantities against which we'll compare the novel abundance index. With the exception of a period roughly between 1981-1988, uncertainty associated with these quantities is fairly high over time. 

**FLAG!** Quantify the uncertainty where possible, and if necessary interrogate the absolute values, e.g., age0 recruits ranges from -50 - 1000, rec-devs from -1 - 1.5, and so on.

# 1. DATA QC + EXPLORATION

```{r 1. DATA QC, cache = T}
#---- DATA QC + EXPLORATION

cabezon <- read.csv("cabezon_calcofi_data.csv")

str(cabezon)

# Add factor variable for season
cabezon <- cabezon %>%
  mutate(season = as.factor(season))

# ---- 1. DATA QC + EXPLORATION

# Check for how many observations per ...

# year
cabezon %>%
  count(year, season, line.station) %>%
  filter(n > 1)
# month
cabezon %>%
  count(year, month, line.station) %>%
  filter(n > 1)
# cruise
cabezon %>%
  count(cruise, line.station) %>%
  filter(n > 1)

# Station coverage across years
cabezon %>%
  group_by(year) %>%
  summarize(n_stations = n_distinct(line.station))
table(cabezon$line.station, cabezon$year)
  # Pretty steady coverage, especially post-1989 

# Check the spatial grid for a quick visual
ggplot(cabezon, aes(longitude, latitude)) +
  geom_point(alpha = 0.3)

# How much zero inflation do we have?
cabezon %>%
  mutate(zero = larvae_100m3 == 0) %>%
  count(zero)
  # Roughly ~85% zeros

# Raw distribution visual
plot(cabezon$volume_sampled, cabezon$larvae_100m3)

```

This initial set of inspections is intended to give a rough picture of how standardized the sampling was over the period of study. For example, were the same number of cruises taken every year? Every season? If not, what were the deviations? How many stations were covered in each cruise? Is this number stable?

The grid map shows the locations of each tow, telling us that, with the exception of some light deviation, all tows are pretty spatially consistent.

**FLAG!** After the grid map, let's generate a plot that attempts to capture the uniformity (or non-uniformity) of tows over time. I'm thinking bubbles sized to how frequently a station is sampled over time, telling us roughly if the spatial arrangement of sampling is stable.

The zero inflation code tells us that we have roughly 84.5% zeroes. That's decent zero inflation, suggesting that we at least include some Delta-Gamma models.

The last scatterplot is a rough picture of the cabezon sampling histogram. It shows us what volume of cabezon larvae we picked up by the volume of water sampled. The pattern is mostly stable, with a few outliers. The pattern is also more or less what we'd expect; namely, that a vast majority of cabezon-positive samples are clustered around a low volume, with some pulling up much larger quantities here and there. In other words, it looks like a histogram.


# 2. DATA EXPLORATION & SUMMARY STATISTICS

```{r 2. DATA EXPLORATION & SUMMARY STATISTICS, cache = T}
#---- 2. DATA EXPLORATION & SUMMARY STATISTICS

# Composite station column (combines line + station)
cabezon <- cabezon %>%
  mutate(station_id = paste(line, station, sep = "_"))

# Annual mean larval observations with SE
annual_cab_summary <- cabezon %>%
  group_by(year) %>%
  summarize(
    mean_larvae = mean(larvae_100m3, na.rm = T),
    sd_larvae = sd(larvae_100m3, na.rm = T),
    n = n(),
    se = sd_larvae / sqrt(n)
  )

# Annual mean larval observations plotted
ggplot(annual_cab_summary, aes(x = year, y = mean_larvae)) +
  geom_line() + 
  geom_point() + 
  geom_ribbon(aes(ymin = mean_larvae - se,
                  ymax = mean_larvae + se),
              alpha = 0.5,
              fill = "skyblue") +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue") +
  #geom_errorbar(aes(ymin = mean_larvae - se,     # Can use error bars and/or ribbon
  #                  ymax = mean_larvae + se),
  #              width = 0.2) +
  labs(
    x = "Year",
    y = expression("Mean larvae per 100 m"^3),
    title = "Annual Mean Larval Observations",
    subtitle = expression("Ribbon shows " %+-% " 1 SE")
  ) +
  theme_bw()

# Average larvae-positive tows by ...
# ... Year
pos_tows_year <- cabezon %>%
  group_by(year) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )
print(pos_tows_year,    caption = "Positive tows by year")

ggplot(pos_tows_year, aes(x = year, y = pct_positive_tows)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue") +
  labs(
    x     = "Year",
    y     = "Positive Tows (%)",
    title = "Percentage of Larvae-Positive Tows by Year"
  ) +
  theme_bw()

# ... Month
pos_tows_month <- cabezon %>%
  group_by(month) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )
print(pos_tows_month,   caption = "Positive tows by month")

ggplot(pos_tows_month, aes(x = month, y = pct_positive_tows)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue") +
  scale_x_continuous(
    breaks = 1:12,
    labels = month.abb
  ) +
  labs(
    x     = "Month",
    y     = "Positive Tows (%)",
    title = "Percentage of Larvae-Positive Tows by Month"
  ) +
  theme_bw()

# ... Season
pos_tows_season <- cabezon %>%
  group_by(season) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )
print(pos_tows_season,  caption = "Positive tows by season")

ggplot(pos_tows_season, aes(x = season, y = pct_positive_tows, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue") +
  labs(
    x     = "Season",
    y     = "Positive Tows (%)",
    title = "Percentage of Larvae-Positive Tows by Season"
  ) +
  theme_bw()

# ... Station
pos_tows_station <- cabezon %>%
  group_by(station_id) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )
print(pos_tows_station, caption = "Positive tows by station")

# Average larvae counts (all tows and positive tows only) by ...
# ... Year
avg_larvae_year <- cabezon %>%
  group_by(year) %>%
  summarise(
    mean_all      = mean(larvae_100m3),
    mean_positive = mean(larvae_100m3[larvae_100m3 > 0])
  )
print(avg_larvae_year,    digits = 3, caption = "Mean larvae by year")

# ... Month
avg_larvae_month <- cabezon %>%
  group_by(month) %>%
  summarise(
    mean_all      = mean(larvae_100m3),
    mean_positive = mean(larvae_100m3[larvae_100m3 > 0])
  )
print(avg_larvae_month,   digits = 3, caption = "Mean larvae by month")

# ... Season
avg_larvae_season <- cabezon %>%
  group_by(season) %>%
  summarise(
    mean_all      = mean(larvae_100m3),
    mean_positive = mean(larvae_100m3[larvae_100m3 > 0])
  )
print(avg_larvae_season,  digits = 3, caption = "Mean larvae by season")

# ... Station
avg_larvae_station <- cabezon %>%
  group_by(station_id) %>%
  summarise(
    mean_all      = mean(larvae_100m3),
    mean_positive = mean(larvae_100m3[larvae_100m3 > 0])
  )
print(avg_larvae_station, digits = 3, caption = "Mean larvae by station")

plot_data <- bind_rows(
  pos_tows_year   %>% mutate(group_var = "Year",   level = as.character(year)),
  pos_tows_month  %>% mutate(group_var = "Month",  level = as.character(month)),
  pos_tows_season %>% mutate(group_var = "Season", level = as.character(season))
) %>%
  mutate(negative_tows = total_tows - positive_tows) %>%
  pivot_longer(
    cols      = c(positive_tows, negative_tows),
    names_to  = "tow_type",
    values_to = "count"
  )

plot_data <- plot_data %>%
  mutate(level = case_when(
    group_var == "Month" ~ factor(level, levels = as.character(1:12)),
    group_var == "Season" ~ factor(level, levels = c("winter", "spring", "summer", "fall")),
    T ~ factor(level)
  ))

```

This section shows how tows break down numerically over time to continue diagnosing the stability and uniformity of sampling.

A few key takeaways from this section:

- It's perhaps worth noting that the percentage of cabezon-positive tows seems fairly stochastic over time. In other words, there is some very rough trending (perhaps decadal?), but overall the percentage of positive samples varies widely between close to zero and approaching 30%.

- The trend is far more steady for months and seasons, as we'd expect. **FLAG!** check biological life history of cabezon, but based on these patterns I'm assuming that spawning season is roughly Oct-Mar, peaking in the winter months.

- It's more difficult to intuitively describe what's happening across space with these checks, so we'll explore that further in Section 4 using heatmaps.

- Inspecting annual values for average volume across all tows and then across all positive tows yields relatively steady results, with the caveat that some years in the late 80's and late 90's/early 00's show higher values for both. **FLAG!** again, perhaps consider a decadal pattern here?

- The monthly values for these averages are pretty much as expected, **but**, **FLAG!** **check out what's happening with December**

- Seasonal values for these averages are also expected, showing higher values during the spawning seasons and dropping off sharply for summer.


# 3. STAR COMPARISONS & INITIAL CCFs

```{r 3. STAR COMPARISONS & INITIAL CCFs, cache = T}

#---- 3. STAR COMPARISONS & INITIAL CCFs

# Join CalCOFI abundance to STAR age-0 recruits and recdevs
STAR_age0_calcofi <- left_join(STAR_age0, annual_cab_summary, by = "year") 
STAR_recdevs_calcofi <- left_join(STAR_recdevs, annual_cab_summary, by = "year")

# Scale factors for dual axis plots
scale_factor_age0 <- max(STAR_age0_calcofi$value, na.rm = T) / max(STAR_age0_calcofi$mean_larvae, na.rm = T)
scale_factor_recdevs <- max(STAR_recdevs_calcofi$value, na.rm = T) / max(STAR_recdevs_calcofi$mean_larvae, na.rm = T)

# Age-0 recruits vs. larval observations
ggplot(STAR_age0_calcofi, aes(x = year)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "skyblue", alpha = 0.5) +
  geom_line(aes(y = value), color = "black") +
  geom_line(aes(y = mean_larvae * scale_factor_age0), color = "coral") +
  scale_y_continuous(
    name = "Age-0 Recruits",
    sec.axis = sec_axis(~ . / scale_factor_age0,
                        name = expression("Mean Larvae per 100 m"^3))
  ) +
  labs(title = "SoCal Cabezon Age-0 Recruits and Larval Observations Over Time") +
  theme_bw()

# Recdevs vs. observed larval observations
ggplot(STAR_recdevs_calcofi, aes(x = year)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "skyblue", alpha = 0.5) +
  geom_line(aes(y = value), color = "black") +
  geom_line(aes(y = mean_larvae * scale_factor_recdevs), color = "coral") +
  scale_y_continuous(
    name = "Recruitment Deviations",
    sec.axis = sec_axis(~ . / scale_factor_recdevs,
                        name = expression("Mean Larvae per 100 m"^3))
  ) +
  labs(title = "SoCal Cabezon Recruitment Deviations and Larval Observations Over Time") +
  theme_bw()

# CCF: Observed larvae vs. recruitment deviations (overlapping years only)
obs_recdevs <- STAR_recdevs_calcofi %>%
  filter(year %in% annual_cab_summary$year & year %in% STAR_recdevs$year)

ccf(obs_recdevs$mean_larvae, obs_recdevs$value,
    lag.max = 10,
    main = "CCF: Larvae Observations vs. Recruitment Deviations (observed years only")

# CCF: Observed larvae vs. age-0 recruits (overlapping years only)
obs_age0 <- STAR_age0_calcofi %>%
  filter(year %in% annual_cab_summary$year & year %in% STAR_age0$year)

ccf(obs_age0$mean_larvae, obs_age0$value,
    lag.max = 10,
    main = "CCF: Larvae Observations vs. Age-0 Recruits (observed years only)")
```

This section is meant to preemptively explore patterns -- if any exist -- between STAR derived quantities and larval observations, *not* estimated larval abundance, which is derived later in our modeling. This step is strictly intended to visualize how our average larvae observations over time compare to stock assessment outputs, assuming both vary relative to some baseline.

We also run some initial CCFs to see if we can pick up any correlation between this raw observation data and the stock assessment outputs. We do **not** see any correlation.

# 4. DATA VISUALIZATION

```{r 4. DATA VISUALIZATION, cache = T}

# ---- 4. DATA VISUALIZATION

#---- 4.a. Station map for visual reference

station_locations <- cabezon %>%
  group_by(line.station) %>%
  summarise(
    latitude  = mean(latitude),
    longitude = mean(longitude),
    .groups   = "drop"
  )

coast <- map_data("world", region = c("USA", "Mexico"))

ggplot() +
  geom_polygon(data = coast, aes(x = long, y = lat, group = group),
               fill = "grey85", color = "grey50", linewidth = 0.3) +
  geom_point(data = station_locations,
             aes(x = longitude, y = latitude),
             color = "steelblue", size = 2) +
  geom_text(data = station_locations,
            aes(x = longitude, y = latitude, label = line.station),
            size = 2.5, hjust = -0.15, vjust = 0.4, color = "grey20") +
  coord_fixed(
    ratio  = 1.3,
    xlim   = c(min(station_locations$longitude) - 1, max(station_locations$longitude) + 1),
    ylim   = c(min(station_locations$latitude)  - 1, max(station_locations$latitude)  + 1)
  ) +
  labs(
    title    = "Cabezon Larvae Sampling Stations",
    subtitle = "CalCOFI survey stations; labels show line.station identifier",
    x        = "Longitude",
    y        = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey40", size = 10),
    panel.grid    = element_line(color = "grey92")
  )

# ---- 4.b. Annual average larval observations

ggplot(annual_cab_summary, aes(x = year, y = mean_larvae)) +
  geom_ribbon(aes(ymin = mean_larvae - se, ymax = mean_larvae + se),
              alpha = 0.5, fill = "skyblue") +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue") +
  labs(
    x = "Year",
    y = expression("Mean larvae per 100 m"^3),
    title = "Annual Mean Larval Observations",
    subtitle = expression("Ribbon shows " %+-% " 1 SE")
  ) +
  theme_bw()

# ---- 4.c. Stacked bar chart: positive vs. total tows by year, month, season

# Shared elements
shared_fill <- scale_fill_manual(
  values = c("negative_tows" = "grey80", "positive_tows" = "steelblue"),
  labels = c("negative_tows" = "Larvae Absent", "positive_tows" = "Larvae Present")
)

shared_bar_theme <- theme_minimal(base_size = 12) +
  theme(
    strip.text         = element_text(face = "bold", size = 13),
    axis.text.x        = element_text(angle = 45, hjust = 1),
    legend.position    = "bottom",
    panel.grid.major.x = element_blank(),
    plot.title         = element_text(face = "bold", size = 15),
    plot.subtitle      = element_text(color = "grey40", size = 10)
  )

# One plot for years; one plot for months + seasons
plot_data_year <- plot_data %>% filter(group_var == "Year")
plot_data_ms   <- plot_data %>% filter(group_var %in% c("Month", "Season"))

# Years plot
ggplot(plot_data_year, aes(x = level, y = count, fill = tow_type)) +
  geom_bar(stat = "identity", width = 0.7) +
  shared_fill +
  shared_bar_theme +
  labs(
    title    = "Cabezon Larvae: Positive vs. Total Tows",
    subtitle = "Stacked bars show larvae-present (blue) against total sampling effort (gray)",
    x        = NULL,
    y        = "Number of Tows",
    fill     = NULL
  )

# Months + Seasons plot
ggplot(plot_data_ms, aes(x = level, y = count, fill = tow_type)) +
  geom_bar(stat = "identity", width = 0.7) +
  facet_wrap(~ group_var, scales = "free_x", nrow = 1) +
  shared_fill + shared_bar_theme +
  labs(
    title = "Cabezon Larvae: Positive vs. Total Tows",
    subtitle = "Stacked bar charts show larve-present (blue) against total sampling effort (gray)",
    x        = NULL,
    y        = "Number of Tows",
    fill     = NULL
  )

#---- 4.d. Boxplots: log-larvae by year, month, season (positive tows only)

# Data binning
box_data <- bind_rows(
  cabezon %>% mutate(group_var = "Year",   level = as.character(year)),
  cabezon %>% mutate(group_var = "Month",  level = as.character(month)),
  cabezon %>% mutate(group_var = "Season", level = as.character(season))
) %>%
  filter(larvae_100m3 > 0) %>%
  mutate(
    log_larvae = log(larvae_100m3),
    group_var  = factor(group_var, levels = c("Year", "Month", "Season")),
    level = case_when(
      group_var == "Month"  ~ factor(level, levels = as.character(1:12)),
      group_var == "Season" ~ factor(level, levels = c("winter", "spring", "summer", "fall")),
      TRUE                  ~ factor(level)
    )
  )

# Shared elements
shared_box_theme <- theme_minimal(base_size = 12) +
  theme(
    strip.text         = element_text(face = "bold", size = 13),
    axis.text.x        = element_text(angle = 45, hjust = 1),
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    plot.title         = element_text(face = "bold", size = 15),
    plot.subtitle      = element_text(color = "grey40", size = 10)
  )

# Years plot
ggplot(box_data %>% filter(group_var == "Year"), 
       aes(x = level, y = log_larvae)) +
  geom_boxplot(fill = "steelblue", color = "grey30", alpha = 0.7, outlier.size = 1) +
  shared_box_theme +
  labs(
    title    = "Cabezon Larvae: Average Count by Year (Positive Tows Only)",
    subtitle = "Log-transformed larvae per 100m³; zero-catch tows excluded",
    x        = NULL,
    y        = "log(Larvae per 100m³)"
  )

# Months + Seasons plot
ggplot(box_data %>% filter(group_var %in% c("Month", "Season")),
       aes(x = level, y = log_larvae)) +
  geom_boxplot(fill = "steelblue", color = "grey30", alpha = 0.7, outlier.size = 1) +
  facet_wrap(~ group_var, scales = "free_x", nrow = 1) +
  shared_box_theme +
  labs(
    title    = "Cabezon Larvae: Average Count by Month and Season (Positive Tows Only)",
    subtitle = "Log-transformed larvae per 100m³; zero-catch tows excluded",
    x = NULL, y = "log(Larvae per 100m³)"
  )


#---- 4.e. Heatmaps: mean log-larvae by station x year, month, season

heatmap_theme <- theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 7),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey40", size = 10)
  )

heatmap_fill <- scale_fill_viridis_c(option = "mako", name = "log(Larvae\nper 100m³)", na.value = "white")

# Year plot
cabezon %>%
  filter(larvae_100m3 > 0) %>%
  group_by(line.station, year) %>%
  summarise(mean_log_larvae = mean(log(larvae_100m3)), .groups = "drop") %>%
  ggplot(aes(x = year, y = line.station, fill = mean_log_larvae)) +
  geom_tile() + heatmap_fill + scale_y_discrete(limits = rev) + heatmap_theme +
  labs(title    = "Cabezon Larval Observations by Station and Year",
       subtitle = "Positive tows only; white = no positive tows recorded",
       x = NULL, y = "South \u2190 Station \u2192 North")

# Month plot
cabezon %>%
  filter(larvae_100m3 > 0) %>%
  group_by(line.station, month) %>%
  summarise(mean_log_larvae = mean(log(larvae_100m3)), .groups = "drop") %>%
  ggplot(aes(x = month, y = line.station, fill = mean_log_larvae)) +
  geom_tile() + heatmap_fill +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_discrete(limits = rev) + heatmap_theme +
  labs(title    = "Cabezon Larval Observations by Station and Month",
       subtitle = "Positive tows only; white = no positive tows recorded",
       x = NULL, y = "South \u2190 Station \u2192 North")

# Season plot
cabezon %>%
  filter(larvae_100m3 > 0) %>%
  group_by(line.station, season) %>%
  summarise(mean_log_larvae = mean(log(larvae_100m3)), .groups = "drop") %>%
  ggplot(aes(x = season, y = line.station, fill = mean_log_larvae)) +
  geom_tile() + heatmap_fill + scale_y_discrete(limits = rev) + heatmap_theme +
  labs(title    = "Cabezon Larval Observations by Station and Season",
       subtitle = "Positive tows only; white = no positive tows recorded",
       x = NULL, y = "South \u2190 Station \u2192 North")
```

**FLAG!!** from the stock assessment, La Nina conditions were observed in 1984, 1987, and 2001.

These plots expand further on some of the numerical summaries we generated earlier in Sections 1 and 2, as visualizing them graphically is probably more intuitive.

For the sake of reference, we also generate a rudimentary map to visualize where the CalCOFI stations are located along the California coast.

The second plot shows again our annual mean larval observations with SE ribbons over the study period where CalCOFI data is available (1981 -- 2015, excepting 1982).

Next, we take a look at some bar plots to visualize the percentage of tows that picked up cabezon larvae. This is much easier to interpret visually than in numbers. Overall, the percentages do not appear to change dramatically between years, though it is worth noting some of the outliers:

- 1983 (which we know had some sampling issues), 1984, 1998, and 2015

 - *1998 high sampling rate due to historic El Nino.*

- **FLAG!** Let's do some further investigation into these outlier values. Something to keep in mind is that later, 1983 is the year that artificially pulls our CCFs tighter.

- **FLAG!** It's probably also worth generating these same plots for percent-positive tows as 100% bar charts, which might be easier to interpret if percentage is our key value here.

- **FLAG!** Again, we need to look closer at December. The percent of positive tows is quite high, but the sampling is very low. Not sure if this will throw anything off of note, but it's worth taking a closer look.

Next, we generate some box plots to inspect the relative volume of larvae captured in positive tows. In other words, the gamma component of our model, or how many larvae are present given there are larvae present. **FLAG!** it's probably worth running some kind of line over these plots (particularly the annual time-series), to smooth out the annual mean observations line and the associated patterns over time. Perhaps this might yield something interesting re: our growing suspicion that there's a (admittedly modest) decadal pattern here.

I'm a bit less certain about how we approach interpreting the month and season box plots. I suppose you could ascribe some of the results here to the vagaries of survey sampling, but it's worth noting some of the more unexpected results. **For example,** August typically pulls very few cabezon larvae despite a high volume of tows, *but*, the log volume of cabezon pulled when August *does* yield cabezon is quite high, relative to other tows. One inspection that's probably worthwhile is **what are the August values over time? Check the averages.**

The next three plots are heatmaps to visualize these patterns in another way, including some coarse approximation of the spatial arrangements.

- One thing we can see, for example, is that on the annual timescale, it appears that northern waters *generally* pull more cabezon larvae, *generally* in higher volumes than southern waters. This makes sense given cabezon habitat **FLAG!** what is cabezon habitat?

- Again, *generally speaking*, the monthly timescale shows similar patterns, namely that cabezon larvae are pulled somewhat more regularly in northern waters and that those northern waters typically have higher volume pulls of cabezon when they are pulled. 

- Lastly, the seasonal plots show much heavier cabezon pulls in the winter, fewer in the spring, fewer still in the autumn, and nearly none in the summer.


# 5. sdmTMB MODEL FITTING

```{r 5. sdmTMB MODEL FITTING, cache = T}

#---- 5. sdmTMB MODEL FITTING
set.seed(my.seed)

#---- 5.a. Build mesh
if (!all(c("X", "Y") %in% names(cabezon))) {
  cabezon <- sdmTMB::add_utm_columns(cabezon, ll_names = c("longitude", "latitude"))
}

mesh <- fm_mesh_2d(
  loc = cabezon[,c("X","Y")],
  cutoff = 20,
  max.edge = c(75, 150),
  offset = c(45, 120)
)
cab_mesh <- make_mesh(data = cabezon, c("X", "Y"), mesh = mesh)
plot(cab_mesh); title("SPDE Mesh for Cabezon CalCOFI analysis")

#---- 5.b. Fit candidate models

# Add year factor variable
cabezon$fyear <- as.factor(cabezon$year)

# NULL Delta-Gamma model

cab_null <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon,
  mesh = cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = "off",
  spatiotemporal = "off",
  offset = NULL,
  extra_time = 1982
)

sanity(cab_null)

# Delta-Gamma: fixed year effect + iid: 
cab_fit01 <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon,
  mesh = cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("iid", "iid"),
  offset = NULL,
  extra_time = 1982
)

sanity(cab_fit01)

# Delta-Gamma: fixed year effect - spatiotemporal: # Is this the best fit for a fixed year effect? Our sanity checks tell us that in models with both a fixed year effect and spatiotemporal fields cause the fields to collapse, suggesting maybe that any temporal variation is adequately captured by the fixed year effect already.
cab_fit02 <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon,
  mesh = cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("off", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(cab_fit02)

# Delta-Gamma: fixed year effect + iid delta:
cab_fit03 <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon,
  mesh = cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("iid", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(cab_fit03)

# Delta-Gamma: full AR1 spatial + spatiotemporal 
cab_fit1 <- sdmTMB(
  list(larvae_100m3 ~ 1, larvae_100m3 ~ 1),
  data = cabezon,
  mesh = cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("ar1", "ar1"),
  offset = NULL,
  extra_time = 1982
)

sanity(cab_fit1)

# Delta-Gamma: spatial only, spatiotemporal off
cab_fit1a <- sdmTMB(
  list(larvae_100m3 ~ 1, larvae_100m3 ~ 1),
  data = cabezon,
  mesh = cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("off", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(cab_fit1a)

# Delta-Gamma; Delta component AR1, Gamma spatiotemporal off
cab_fit1b <- sdmTMB(
  list(larvae_100m3 ~ 1, larvae_100m3 ~ 1),
  data = cabezon,
  mesh = cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("ar1", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(cab_fit1b)

# Delta-Gamma: Gamma component AR1, Delta spatiotemporal off
cab_fit1c <- sdmTMB(
  list(larvae_100m3 ~ 1, larvae_100m3 ~ 1),
  data = cabezon,
  mesh = cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("off", "ar1"),
  offset = NULL,
  extra_time = 1982
)

sanity(cab_fit1c)

# Tweedie: full AR1 spatial + spatiotemporal
cab_fit2 <- sdmTMB(
  formula = larvae_100m3 ~ 1,
  data = cabezon,
  mesh = cab_mesh,
  time = "year",
  family = tweedie(),
  spatial = "on",
  spatiotemporal = "ar1",
  offset = NULL,
  extra_time = 1982
)

sanity(cab_fit2)

# Tweedie: spatial only, spatiotemporal off
cab_fit2a <- sdmTMB(
  formula = larvae_100m3 ~ 1,
  data = cabezon,
  mesh = cab_mesh,
  time = "year",
  family = tweedie(),
  spatial = "on",
  spatiotemporal = "off",
  offset = NULL,
  extra_time = 1982
)

sanity(cab_fit2a)

#---- 5.c. Cross validation

cv_null <- sdmTMB_cv(cab_null$call$formula, cabezon, cab_mesh, delta_gamma(), k_folds = 5)
cv_m01 <- sdmTMB_cv(cab_fit01$call$formula, cabezon, cab_mesh, delta_gamma(), k_folds = 5)
cv_m02 <- sdmTMB_cv(cab_fit02$call$formula, cabezon, cab_mesh, delta_gamma(), k_folds = 5)
cv_m03 <- sdmTMB_cv(cab_fit03$call$formula, cabezon, cab_mesh, delta_gamma(), k_folds = 5)
cv_m1 <- sdmTMB_cv(cab_fit1$call$formula, cabezon, cab_mesh, delta_gamma(),k_folds = 5)
cv_m1a <- sdmTMB_cv(cab_fit1a$call$formula, cabezon, cab_mesh, delta_gamma(), k_folds = 5)
cv_m1b <- sdmTMB_cv(cab_fit1b$call$formula, cabezon, cab_mesh, delta_gamma(), k_folds = 5)
cv_m1c <- sdmTMB_cv(cab_fit1c$call$formula, cabezon, cab_mesh, delta_gamma(), k_folds = 5)
cv_m2 <- sdmTMB_cv(cab_fit2a$call$formula, cabezon, cab_mesh, tweedie(), k_folds = 5)
cv_m2a <- sdmTMB_cv(cab_fit2a$call$formula, cabezon, cab_mesh, tweedie(), k_folds = 5)

#---- 5.d. Construct model comparison table

cv_table <- data.frame(
  fit_name = c(
    "cab_null",
    "cab_fit01",
    "cab_fit02",
    "cab_fit03",
    "cab_fit1",
    "cab_fit1a",
    "cab_fit1b",
    "cab_fit1c",
    "cab_fit2",
    "cab_fit2a"
  ),
  model = c(
    "Delta-Gamma NULL",
    "Delta-Gamma spatial + spatiotemporal IID + fixed year effect",
    "Delta-Gamma spatial + fixed year effect - spatiotemporal",
    "Delta-Gamma: fixed year effect + iid delta",
    "Delta-Gamma spatial + spatiotemporal AR1",
    "Delta-Gamma spatial",
    "Delta-Gamma spatiotemporal Delta AR1",
    "Delta-Gamma spatiotemporal Gamma AR1",
    "Tweedie spatial + spatiotemporal AR1",
    "Tweedie spatial"
  ),
  AIC = c(
    AIC(cab_null),
    AIC(cab_fit01),
    AIC(cab_fit02),
    AIC(cab_fit03),
    AIC(cab_fit1),
    AIC(cab_fit1a),
    AIC(cab_fit1b),
    AIC(cab_fit1c),
    AIC(cab_fit2),
    AIC(cab_fit2a)
  ),
    delta_AIC = NA,
    CV = c(
      cv_null$sum_loglik,
      cv_m01$sum_loglik,
      cv_m02$sum_loglik,
      cv_m03$sum_loglik,
      cv_m1$sum_loglik,
      cv_m1a$sum_loglik,
      cv_m1b$sum_loglik,
      cv_m1c$sum_loglik,
      cv_m2$sum_loglik,
      cv_m2a$sum_loglik
    ),
    delta_CV = NA
  )

cv_table$delta_AIC <- cv_table$AIC - min(cv_table$AIC)
cv_table$delta_CV <- cv_table$CV - min(cv_table$CV)

cv_table <- cv_table[order(cv_table$CV, decreasing = T), ]

cv_table

#---- 5.e. Sanity checks and random effect parameter estimates for best models

sanity(cab_fit1)
sanity(cab_fit01)
sanity(cab_fit2)
sanity(cab_fit03)

tidy(cab_fit1, effects = "ran_pars")
tidy(cab_fit01, effects = "ran_pars")
tidy(cab_fit2, effects = "ran_pars")
tidy(cab_fit03, effects = "ran_pars")

```

**RE-RUN** models constrained to spawning season and spatially 

- *double check how the model is interpreting the spatial exactitude of station locations*



# 6. MODEL DIAGNOSTICS & ABUNDANCE INDICES

## 6.1. cab_fit1 — Full Delta-Gamma Model

```{r 6.1. cab_fit1 diagnostics, cache = T}

#---- 6.1.a. Calibration summaries

pred_obs <- predict(cab_fit1, type = "response")

# Overall mean calibration
pred_obs %>%
  summarise(
    obs_mean = mean(larvae_100m3, na.rm = TRUE),
    pred_mean = mean(est, na.rm = TRUE),
    obs_occ = mean(larvae_100m3 > 0, na.rm = TRUE),
    pred_occ = mean(est1, na.rm = TRUE),
    obs_pos_mean = mean(larvae_100m3[larvae_100m3 > 0], na.rm = TRUE),
    pred_pos_mean = mean(est2, na.rm = TRUE)
  )

# Binomial calibration; i.e. predicted presence probability vs. observed presence rate
pred_obs$presence <- pred_obs$larvae_100m3 > 0

calibration_summary <- pred_obs %>%
  mutate(prob_bin = cut(est1, breaks = seq(0, 1, by = 0.1))) %>%
  group_by(prob_bin) %>%
  summarise(
    mean_pred = mean(est1),
    observed = mean(presence),
    n = n()
  )

plot(calibration_summary$mean_pred, calibration_summary$observed,
     xlab = "Predicted presence probability",
     ylab = "Observed presence frequency")
abline(0,1,lty=2)

#---- 6.1.b. Predicted vs. observed plots

# Raw scale comparison
plot(
  x = pred_obs$est,
  y = cabezon$larvae_100m3,
  xlab = "Predicted",
  ylab = "Observed"
)
abline(0, 1, lty = 2)

# Log scale comparison
plot(log1p(pred_obs$est), log1p(cabezon$larvae_100m3),
     xlab = "log(Predicted + 1)", ylab = "log(Observed + 1)")
abline(0, 1, lty = 2)

# Gamma (positive density) component only
pos_idx <- cabezon$larvae_100m3 > 0
plot(log(pred_obs$est2[pos_idx]), log(cabezon$larvae_100m3[pos_idx]),
     xlab = "log(Predicted positive abundance)",
     ylab = "log(Observed positive abundance)")
abline(0, 1, lty = 2)

#---- 6.1.c. Randomized quantile residuals

rq_res <- residuals(cab_fit1, type = "mle-mvn")

# QQ plot
qqnorm(rq_res)
qqline(rq_res)

# Spatial residual map by year (kind of hard to see)
pred_obs$rq_resid <- rq_res

ggplot(pred_obs, aes(x = longitude, y = latitude, color = rq_resid)) +
  geom_point() +
  scale_color_gradient2() +
  facet_wrap(~ year)

#---- 6.1.d. Temporal calibration

# Are predictions tracking temporal trends?
pred_obs %>%
  group_by(year) %>%
  summarize(
    obs = mean(larvae_100m3),
    pred = mean(est)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = obs), color = "black") +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed")

#---- 6.1.e. Abundance index construction
# Note! The index uncertainty here was far too wide using a regular spatial grid, probably because interpolation over years with no CalCOFI coverage was causing inflation. We'll probably wind up using this interpolated index, as it's more in line with how we'd expect to use sdmTMB results; namely, that we're extrapolating abundance between spatial and temporal steps (i.e., beyond just the sampled stations).

pred_grid <- expand.grid(
  X = seq(min(cabezon$X), max(cabezon$X), by = 5),
  Y = seq(min(cabezon$Y), max(cabezon$Y), by = 5)
) %>%
  tidyr::crossing(year = as.integer(sort(unique(cabezon$year)))) %>%
  mutate(fyear = as.factor(year))

cab_map <- predict(cab_fit1, newdata = pred_grid, return_tmb_object = T)
cab_index <- get_index(cab_map, area = 5^2, bias_correct = T)

ggplot(cab_index, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year", title = "Abundance Index: Regular Grid (Inflated by Interpolation)")

#---- 6.1.f. Abundance index: constrained to observed station grid (probably preferred)

pred_grid_obs <- cabezon %>%
  dplyr::select(X, Y, year) %>%
  distinct()

cab_map_obs <- predict(cab_fit1,
                       newdata = pred_grid_obs,
                       return_tmb_object = T)

cab_index_obs <- get_index(cab_map_obs,
                           area = 5^2,
                           bias_correct = T)

ggplot(cab_index_obs, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year", title = "Abundance Index: Observed Station Grid")
```

## 6.2. cab_fit01 — Fixed Year Effects Delta-Gamma Model

```{r 6.2. cab_fit01 diagnostics, cache = T}

#---- 6.2.a. Calibration summaries

pred_obs_sp <- predict(cab_fit01, type = "response")

# Overall mean calibration
pred_obs_sp %>%
  summarise(
    obs_mean = mean(larvae_100m3, na.rm = TRUE),
    pred_mean = mean(est, na.rm = TRUE),
    obs_occ = mean(larvae_100m3 > 0, na.rm = TRUE),
    pred_occ = mean(est1, na.rm = TRUE),
    obs_pos_mean = mean(larvae_100m3[larvae_100m3 > 0], na.rm = TRUE),
    pred_pos_mean = mean(est2, na.rm = TRUE)
  )

# Binomial calibration; i.e. predicted presence probability vs. observed presence rate
pred_obs_sp$presence <- pred_obs_sp$larvae_100m3 > 0

calibration_summary_sp <- pred_obs_sp %>%
  mutate(prob_bin = cut(est1, breaks = seq(0, 1, by = 0.1))) %>%
  group_by(prob_bin) %>%
  summarise(
    mean_pred = mean(est1),
    observed = mean(presence),
    n = n()
  )

plot(calibration_summary_sp$mean_pred, calibration_summary_sp$observed,
     xlab = "Predicted presence probability",
     ylab = "Observed presence frequency")
abline(0,1,lty=2)

#---- 6.2.b. Predicted vs. observed plots

# Raw scale comparison
plot(
  x = pred_obs_sp$est,
  y = cabezon$larvae_100m3,
  xlab = "Predicted",
  ylab = "Observed"
)
abline(0, 1, lty = 2)

# Log scale comparison
plot(log1p(pred_obs_sp$est), log1p(cabezon$larvae_100m3),
     xlab = "log(Predicted + 1)", ylab = "log(Observed + 1)")
abline(0, 1, lty = 2)

# Gamma (positive density) component only
pos_idx_sp <- cabezon$larvae_100m3 > 0
plot(log(pred_obs_sp$est2[pos_idx_sp]), log(cabezon$larvae_100m3[pos_idx_sp]),
     xlab = "log(Predicted positive abundance)",
     ylab = "log(Observed positive abundance)")
abline(0, 1, lty = 2)

#---- 6.2.c. Randomized quantile residuals

rq_res_sp <- residuals(cab_fit01, type = "mle-mvn")

# QQ plot
qqnorm(rq_res_sp)
qqline(rq_res_sp)

# Spatial residual map by year (kind of hard to see)
pred_obs_sp$rq_resid <- rq_res_sp

ggplot(pred_obs_sp, aes(x = X, y = Y, color = rq_resid)) +
  geom_point() +
  scale_color_gradient2() +
  facet_wrap(~ year)

#---- 6.2.d. Temporal calibration

# Are predictions tracking temporal trends?
pred_obs_sp %>%
  group_by(year) %>%
  summarize(
    obs = mean(larvae_100m3),
    pred = mean(est)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = obs), color = "black") +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed")

#---- 6.2.e. Abundance index construction
# Note! The index uncertainty here was far too wide using a regular spatial grid, probably because interpolation over years with no CalCOFI coverage was causing inflation. I kept that plot in for reference, but 6.2.f. (observed station grid) is probably preferred for diagnosis.

pred_grid_sp <- expand.grid(
  X = seq(min(cabezon$X), max(cabezon$X), by = 5),
  Y = seq(min(cabezon$Y), max(cabezon$Y), by = 5)
) %>%
  tidyr::crossing(year = as.integer(sort(unique(cabezon$year)))) %>%
  mutate(fyear = as.factor(year))

cab_map_sp <- predict(cab_fit01, newdata = pred_grid_sp, return_tmb_object = T)
cab_index_sp <- get_index(cab_map_sp, area = 5^2, bias_correct = T)

ggplot(cab_index_sp, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year", title = "Abundance Index: Regular Grid (Inflated by Interpolation)")

#---- 6.2.f. Abundance index: constrained to observed station grid (probably preferred)

pred_grid_sp_obs <- cabezon %>%
  dplyr::select(X, Y, year, fyear) %>%
  distinct()

cab_map_obs_sp <- predict(cab_fit01,
                       newdata = pred_grid_sp_obs,
                       return_tmb_object = T)

cab_index_obs_sp <- get_index(cab_map_obs_sp,
                           area = 5^2,
                           bias_correct = T)

ggplot(cab_index_obs_sp, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year", title = "Abundance Index: Observed Station Grid")

```



# 7. STAR COMPARISONS & CCFs

## 7.1. cab_fit1

```{r 7.1. STAR comparisons cab_fit1 full extrapolation, cache = T}

# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 6.1.f. abundance index, assuming it's more stable.
combined <- cab_index %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
combined_obs <- combined %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf(combined_obs$larvae_index, combined_obs$age0,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Age-0 Recruits")

# CCF: Restricted larval index vs. recruitment deviations
ccf(combined_obs$larvae_index, combined_obs$rec_dev,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Recruitment Deviations")

# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf(combined_obs$larvae_index, combined_obs$ssb,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Spawning Stock Biomass (raw)")

larvae_detrended <- residuals(loess(larvae_index ~ year, data = combined_obs, span = 0.75))
ssb_detrended    <- residuals(loess(ssb ~ year,          data = combined_obs, span = 0.75))

ccf(larvae_detrended, ssb_detrended, 
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. SSB (detrended)")
```

```{r 7.1.2. STAR comparisons cab_fit1 station extrapolation}

# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 6.1.f. abundance index, assuming it's more stable.
combined <- cab_index_obs %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
combined_obs <- combined %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf(combined_obs$larvae_index, combined_obs$age0,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Age-0 Recruits")

# CCF: Restricted larval index vs. recruitment deviations
ccf(combined_obs$larvae_index, combined_obs$rec_dev,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Recruitment Deviations")

# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf(combined_obs$larvae_index, combined_obs$ssb,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Spawning Stock Biomass (raw)")

larvae_detrended <- residuals(loess(larvae_index ~ year, data = combined_obs, span = 0.75))
ssb_detrended    <- residuals(loess(ssb ~ year,          data = combined_obs, span = 0.75))

ccf(larvae_detrended, ssb_detrended, 
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. SSB (detrended)")
```


## 7.2. cab_fit01

```{r 7.2. STAR comparisons cab_fit01 full extrapolation, cache = T}

# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 6.2.f. abundance index, assuming it's more stable.
combined_sp <- cab_index_sp %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
combined_sp_obs <- combined_sp %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf(combined_sp_obs$larvae_index, combined_sp_obs$age0,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Age-0 Recruits")

# CCF: Restricted larval index vs. recruitment deviations
ccf(combined_sp_obs$larvae_index, combined_sp_obs$rec_dev,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Recruitment Deviations")

# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf(combined_sp_obs$larvae_index, combined_sp_obs$ssb,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Spawning Stock Biomass (raw)")

larvae_detrended <- residuals(lm(larvae_index ~ year, data = combined_obs))
ssb_detrended <- residuals(lm(ssb ~ year, data = combined_obs))

ccf(larvae_detrended, ssb_detrended, 
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. SSB (detrended)")
```

```{r 7.2.2. STAR comparisons cab_fit01 station extrapolation, cache = T}

# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 6.2.f. abundance index, assuming it's more stable.
combined_sp <- cab_index_obs_sp %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
combined_sp_obs <- combined_sp %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf(combined_sp_obs$larvae_index, combined_sp_obs$age0,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Age-0 Recruits")

# CCF: Restricted larval index vs. recruitment deviations
ccf(combined_sp_obs$larvae_index, combined_sp_obs$rec_dev,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Recruitment Deviations")

# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf(combined_sp_obs$larvae_index, combined_sp_obs$ssb,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Spawning Stock Biomass (raw)")

larvae_detrended <- residuals(lm(larvae_index ~ year, data = combined_obs))
ssb_detrended <- residuals(lm(ssb ~ year, data = combined_obs))

ccf(larvae_detrended, ssb_detrended, 
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. SSB (detrended)")
```


## 7.3. cab_fit1 — Correlation Analysis


```{r 7.3. cab_fit1 — Correlation Analysis, cache = T}
# RQ residuals for full AR1 spatiotemporal model
pred_obs %>%
  group_by(year) %>%
  mutate(med_resid = median(rq_resid)) %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method  = "loess", span = 0.4,
    se      = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Year",
    y     = "RQ Residual",
    title = "Annual RQ Residual Distributions — cab_fit1"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

pred_obs %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x     = "Year",
    y     = "Cumulative Sum of Median RQ Residual",
    title = "CUSUM — cab_fit1"
  ) +
  theme_minimal()

#---- Simple lagged regression at lag -1

combined_lagged <- combined_obs %>%
  arrange(year) %>%
  mutate(
    age0_lead1    = lead(age0, 1),
    rec_dev_lead1 = lead(rec_dev, 1)
  )

lm_age0   <- lm(age0_lead1    ~ larvae_index, data = combined_lagged)
lm_recdev <- lm(rec_dev_lead1 ~ larvae_index, data = combined_lagged)

summary(lm_age0)
summary(lm_recdev)

combined_lagged %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

# Sensitivity: remove 1983
combined_lagged_no83 <- combined_lagged %>% filter(year != 1983)
lm_age0_no83 <- lm(age0_lead1 ~ larvae_index, data = combined_lagged_no83)
summary(lm_age0_no83)

# Detrended regressions
lm_age0_detrend   <- lm(age0_lead1    ~ larvae_index + year, data = combined_lagged)
lm_recdev_detrend <- lm(rec_dev_lead1 ~ larvae_index + year, data = combined_lagged)
summary(lm_age0_detrend)
summary(lm_recdev_detrend)

# Cook's distance
plot(lm_age0, which = 4)

# Trimmed analysis: remove three most anomalous years
combined_lagged_trimmed <- combined_lagged %>%
  filter(!year %in% c(1983, 1986, 1998))

combined_lagged_trimmed %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

lm_age0_trimmed         <- lm(age0_lead1 ~ larvae_index, data = combined_lagged_trimmed)
lm_age0_trimmed_detrend <- lm(age0_lead1 ~ larvae_index + year, data = combined_lagged_trimmed)
summary(lm_age0_trimmed)
summary(lm_age0_trimmed_detrend)

# 1. Is the underprediction concentrated in one delta component?
pred_obs <- pred_obs %>%
  mutate(
    rq_resid1 = residuals(cab_fit1, model = 1, type = "mle-mvn"),
    rq_resid2 = residuals(cab_fit1, model = 2, type = "mle-mvn")
  )

pred_obs %>%
  group_by(year) %>%
  summarise(
    med_resid1 = median(rq_resid1, na.rm = TRUE),
    med_resid2 = median(rq_resid2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    cusum1 = cumsum(replace_na(med_resid1, 0)),
    cusum2 = cumsum(replace_na(med_resid2, 0))
  ) %>%
  pivot_longer(cols = c(cusum1, cusum2),
               names_to = "component",
               values_to = "cusum") %>%
  ggplot(aes(x = year, y = cusum, color = component)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("cusum1" = "steelblue", "cusum2" = "firebrick"),
                     labels = c("cusum1" = "Delta (binomial)", "cusum2" = "Gamma")) +
  labs(x = "Year", y = "CUSUM",
       title = "CUSUM by Delta Model Component — cab_fit1") +
  theme_minimal()

# 2. Is the underprediction spatially structured?
pred_obs %>%
  group_by(station) %>%
  summarise(mean_resid = mean(rq_resid), .groups = "drop") %>%
  ggplot(aes(x = station, y = mean_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = TRUE) +
  labs(x = "Station", y = "Mean RQ Residual",
       title = "Mean Residual by Station — cab_fit1") +
  theme_minimal()
```


## 7.4. cab_fit01 — Correlation Analysis

```{r 7.4. cab_fit01 — Correlation Analysis, cache = T}
# RQ residuals for spatial-only model
pred_obs_sp %>%
  group_by(year) %>%
  mutate(med_resid = median(rq_resid)) %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method  = "loess", span = 0.4,
    se      = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Year",
    y     = "RQ Residual",
    title = "Annual RQ Residual Distributions — cab_fit01"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

pred_obs_sp %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x     = "Year",
    y     = "Cumulative Sum of Median RQ Residual",
    title = "CUSUM — cab_fit01"
  ) +
  theme_minimal()

#---- Simple lagged regression at lag -1

combined_sp_lagged <- combined_sp_obs %>%
  arrange(year) %>%
  mutate(
    age0_lead1    = lead(age0, 1),
    rec_dev_lead1 = lead(rec_dev, 1)
  )

lm_age0_sp   <- lm(age0_lead1    ~ larvae_index, data = combined_sp_lagged)
lm_recdev_sp <- lm(rec_dev_lead1 ~ larvae_index, data = combined_sp_lagged)

summary(lm_age0_sp)
summary(lm_recdev_sp)

combined_sp_lagged %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

# Sensitivity: remove 1983
combined_sp_lagged_no83 <- combined_sp_lagged %>% filter(year != 1983)
lm_age0_sp_no83 <- lm(age0_lead1 ~ larvae_index, data = combined_sp_lagged_no83)
summary(lm_age0_sp_no83)

# Detrended regressions
lm_age0_sp_detrend   <- lm(age0_lead1    ~ larvae_index + year, data = combined_sp_lagged)
lm_recdev_sp_detrend <- lm(rec_dev_lead1 ~ larvae_index + year, data = combined_sp_lagged)
summary(lm_age0_sp_detrend)
summary(lm_recdev_sp_detrend)

# Cook's distance
plot(lm_age0_sp, which = 4)

# Trimmed analysis: remove three most anomalous years
combined_sp_lagged_trimmed <- combined_sp_lagged %>%
  filter(!year %in% c(1983, 1986, 1998))

combined_sp_lagged_trimmed %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

lm_age0_sp_trimmed         <- lm(age0_lead1 ~ larvae_index, data = combined_sp_lagged_trimmed)
lm_age0_sp_trimmed_detrend <- lm(age0_lead1 ~ larvae_index + year, data = combined_sp_lagged_trimmed)
summary(lm_age0_sp_trimmed)
summary(lm_age0_sp_trimmed_detrend)

# 1. Is the underprediction concentrated in one delta component?
pred_obs_sp <- pred_obs_sp %>%
  mutate(
    rq_resid1 = residuals(cab_fit01, model = 1, type = "mle-mvn"),
    rq_resid2 = residuals(cab_fit01, model = 2, type = "mle-mvn")
  )

pred_obs_sp %>%
  group_by(year) %>%
  summarise(
    med_resid1 = median(rq_resid1, na.rm = TRUE),
    med_resid2 = median(rq_resid2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    cusum1 = cumsum(replace_na(med_resid1, 0)),
    cusum2 = cumsum(replace_na(med_resid2, 0))
  ) %>%
  pivot_longer(cols = c(cusum1, cusum2),
               names_to = "component", values_to = "cusum") %>%
  ggplot(aes(x = year, y = cusum, color = component)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("cusum1" = "steelblue", "cusum2" = "firebrick"),
                     labels = c("cusum1" = "Delta (binomial)", "cusum2" = "Gamma")) +
  labs(x = "Year", y = "CUSUM",
       title = "CUSUM by Delta Model Component — cab_fit01") +
  theme_minimal()

# 2. Is the underprediction spatially structured?
pred_obs_sp %>%
  group_by(station) %>%
  summarise(mean_resid = mean(rq_resid), .groups = "drop") %>%
  ggplot(aes(x = station, y = mean_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = TRUE) +
  labs(x = "Station", y = "Mean RQ Residual",
       title = "Mean Residual by Station — cab_fit01") +
  theme_minimal()
```



---

# PART II — CONSTRAINED ANALYSIS (SPAWNING SEASON + SHELF REGION)

---

# RE-RUNNING WITH CONSTRAINED TEMPORAL AND SPATIAL WINDOWS

To refresh our memory, all analysis to this point has been done using CalCOFI data in the core CalCOFI grid, extended to the maximum distance from shore that these transect lines go (in excess of some 250 mi). Additionally, our data compasses all months and seasons over the observed time-series from 1981-2015. 

This section will attempt to apply a finer grade to our analysis by constraining these parameters in the following ways:

1. We will focus on spawning season, defined in the 2019 stock assessment as **beginning in November and ending in March, with a peak in January and February.**

2. We will focus on the shelf rather than the full distance offshore, assuming that cabezon larvae occur most frequently and most regularly inshore.

It's worth keeping in mind, of course, that this constrains our spatial data further, *but,* our natural science might suggest that a narrower examination of spawning season larval abundance *in areas where we'd more likely expect larval occurrence* could be a better match to downstream adult distribution and abundance.

# 8. DATA PARING & EXPLORATION (CONSTRAINED)

```{r 8. DATA PARING AND EXPLORATION (CONSTRAINED), cache = T}

cabezon_shelf_spawn <- cabezon %>%
  filter(
    month %in% c(1, 2, 3),
    station <= 60.0
  )

cat("Rows retained:", nrow(cabezon_shelf_spawn), "of", nrow(cabezon), "\n")
cat("Years covered:", paste(range(cabezon_shelf_spawn$year), collapse = "–"), "\n")
cat("Months present:", paste(sort(unique(cabezon_shelf_spawn$month)), collapse = ", "), "\n")
cat("Station range:", paste(range(cabezon_shelf_spawn$station), collapse = "–"), "\n")
cat("Lines present:", paste(sort(unique(cabezon_shelf_spawn$line)), collapse = ", "), "\n")

cabezon_shelf_spawn %>%
  group_by(year) %>%
  summarize(
    n_tows = n(),
    n_pos = sum(larvae_100m3 > 0),
    pct_pos = round(100 * n_pos / n_tows, 1)
  ) %>%
  print(n = Inf)

# Composite station column (combines line + station)
cabezon_shelf_spawn <- cabezon_shelf_spawn %>%
  mutate(station_id = paste(line, station, sep = "_"))

# Calculating positive tow values for visualization
ss_pos_tows_year <- cabezon_shelf_spawn %>%
  group_by(year) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )

ss_pos_tows_month <- cabezon_shelf_spawn %>%
  group_by(month) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )

ss_pos_tows_season <- cabezon_shelf_spawn %>%
  group_by(season) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )

ss_pos_tows_station <- cabezon_shelf_spawn %>%
  group_by(station_id) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )

# Annual mean larval observations with SE
annual_cab_ss_summary <- cabezon_shelf_spawn %>%
  group_by(year) %>%
  summarize(
    mean_larvae = mean(larvae_100m3, na.rm = T),
    sd_larvae = sd(larvae_100m3, na.rm = T),
    n = n(),
    se = sd_larvae / sqrt(n)
  )

# ---- 8.b. Annual average larval observations

ggplot(annual_cab_ss_summary, aes(x = year, y = mean_larvae)) +
  geom_ribbon(aes(ymin = mean_larvae - se, ymax = mean_larvae + se),
              alpha = 0.5, fill = "skyblue") +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue") +
  labs(
    x = "Year",
    y = expression("Mean larvae per 100 m"^3),
    title = "Annual Mean Larval Observations, Constrained to Spawning Season and Coastal Shelf",
    subtitle = expression("Ribbon shows " %+-% " 1 SE")
  ) +
  theme_bw()

# ---- 8.c. Stacked bar chart: positive vs. total tows by year, month, season

ss_plot_data <- bind_rows(
  ss_pos_tows_year   %>% mutate(group_var = "Year",   level = as.character(year)),
  ss_pos_tows_month  %>% mutate(group_var = "Month",  level = as.character(month)),
  ss_pos_tows_season %>% mutate(group_var = "Season", level = as.character(season))
) %>%
  mutate(negative_tows = total_tows - positive_tows) %>%
  pivot_longer(
    cols      = c(positive_tows, negative_tows),
    names_to  = "tow_type",
    values_to = "count"
  )

ss_plot_data <- ss_plot_data %>%
  mutate(level = case_when(
    group_var == "Month" ~ factor(level, levels = as.character(1:12)),
    group_var == "Season" ~ factor(level, levels = c("winter", "spring", "summer", "fall")),
    T ~ factor(level)
  ))

# Shared elements
shared_fill <- scale_fill_manual(
  values = c("negative_tows" = "grey80", "positive_tows" = "steelblue"),
  labels = c("negative_tows" = "Larvae Absent", "positive_tows" = "Larvae Present")
)

shared_bar_theme <- theme_minimal(base_size = 12) +
  theme(
    strip.text         = element_text(face = "bold", size = 13),
    axis.text.x        = element_text(angle = 45, hjust = 1),
    legend.position    = "bottom",
    panel.grid.major.x = element_blank(),
    plot.title         = element_text(face = "bold", size = 15),
    plot.subtitle      = element_text(color = "grey40", size = 10)
  )

# One plot for years; one plot for months + seasons
ss_plot_data_year <- ss_plot_data %>% filter(group_var == "Year")
ss_plot_data_ms   <- ss_plot_data %>% filter(group_var %in% c("Month", "Season"))

# Years plot
ggplot(ss_plot_data_year, aes(x = level, y = count, fill = tow_type)) +
  geom_bar(stat = "identity", width = 0.7) +
  shared_fill +
  shared_bar_theme +
  labs(
    title    = "Cabezon Larvae: Positive vs. Total Tows",
    subtitle = "Stacked bars show larvae-present (blue) against total sampling effort (gray)",
    x        = NULL,
    y        = "Number of Tows",
    fill     = NULL
  )

# Months + Seasons plot
ggplot(ss_plot_data_ms, aes(x = level, y = count, fill = tow_type)) +
  geom_bar(stat = "identity", width = 0.7) +
  facet_wrap(~ group_var, scales = "free_x", nrow = 1) +
  shared_fill + shared_bar_theme +
  labs(
    title = "Cabezon Larvae: Positive vs. Total Tows",
    subtitle = "Stacked bar charts show larve-present (blue) against total sampling effort (gray)",
    x        = NULL,
    y        = "Number of Tows",
    fill     = NULL
  )

#---- 8.d. Boxplots: log-larvae by year, month, season (positive tows only)

# Data binning
ss_box_data <- bind_rows(
  cabezon_shelf_spawn %>% mutate(group_var = "Year",   level = as.character(year)),
  cabezon_shelf_spawn %>% mutate(group_var = "Month",  level = as.character(month))
) %>%
  filter(larvae_100m3 > 0) %>%
  mutate(
    log_larvae = log(larvae_100m3),
    group_var  = factor(group_var, levels = c("Year", "Month", "Season")),
    level = case_when(
      group_var == "Month"  ~ factor(level, levels = as.character(1:12)),
      group_var == "Season" ~ factor(level, levels = c("winter", "spring", "summer", "fall")),
      TRUE                  ~ factor(level)
    )
  )

# Shared elements
shared_box_theme <- theme_minimal(base_size = 12) +
  theme(
    strip.text         = element_text(face = "bold", size = 13),
    axis.text.x        = element_text(angle = 45, hjust = 1),
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    plot.title         = element_text(face = "bold", size = 15),
    plot.subtitle      = element_text(color = "grey40", size = 10)
  )

# Years plot
ggplot(ss_box_data %>% filter(group_var == "Year"), 
       aes(x = level, y = log_larvae)) +
  geom_boxplot(fill = "steelblue", color = "grey30", alpha = 0.7, outlier.size = 1) +
  shared_box_theme +
  labs(
    title    = "Cabezon Larvae: Average Count by Year (Positive Tows Only)",
    subtitle = "Log-transformed larvae per 100m³; zero-catch tows excluded",
    x        = NULL,
    y        = "log(Larvae per 100m³)"
  )

# Months + Seasons plot
ggplot(ss_box_data %>% filter(group_var %in% c("Month", "Season")),
       aes(x = level, y = log_larvae)) +
  geom_boxplot(fill = "steelblue", color = "grey30", alpha = 0.7, outlier.size = 1) +
  facet_wrap(~ group_var, scales = "free_x", nrow = 1) +
  shared_box_theme +
  labs(
    title    = "Cabezon Larvae: Average Count by Month and Season (Positive Tows Only)",
    subtitle = "Log-transformed larvae per 100m³; zero-catch tows excluded",
    x = NULL, y = "log(Larvae per 100m³)"
  )


#---- 8.e. Heatmaps: mean log-larvae by station x year, month, season

heatmap_theme <- theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 7),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey40", size = 10)
  )

heatmap_fill <- scale_fill_viridis_c(option = "mako", name = "log(Larvae\nper 100m³)", na.value = "white")

# Year plot
cabezon_shelf_spawn %>%
  filter(larvae_100m3 > 0) %>%
  group_by(line.station, year) %>%
  summarise(mean_log_larvae = mean(log(larvae_100m3)), .groups = "drop") %>%
  ggplot(aes(x = year, y = line.station, fill = mean_log_larvae)) +
  geom_tile() + heatmap_fill + scale_y_discrete(limits = rev) + heatmap_theme +
  labs(title    = "Cabezon Larval Observations by Station and Year",
       subtitle = "Positive tows only; white = no positive tows recorded",
       x = NULL, y = "South \u2190 Station \u2192 North")

# Month plot
cabezon_shelf_spawn %>%
  filter(larvae_100m3 > 0) %>%
  group_by(line.station, month) %>%
  summarise(mean_log_larvae = mean(log(larvae_100m3)), .groups = "drop") %>%
  ggplot(aes(x = month, y = line.station, fill = mean_log_larvae)) +
  geom_tile() + heatmap_fill +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_discrete(limits = rev) + heatmap_theme +
  labs(title    = "Cabezon Larval Observations by Station and Month",
       subtitle = "Positive tows only; white = no positive tows recorded",
       x = NULL, y = "South \u2190 Station \u2192 North")

# Season plot
cabezon_shelf_spawn %>%
  filter(larvae_100m3 > 0) %>%
  group_by(line.station, season) %>%
  summarise(mean_log_larvae = mean(log(larvae_100m3)), .groups = "drop") %>%
  ggplot(aes(x = season, y = line.station, fill = mean_log_larvae)) +
  geom_tile() + heatmap_fill + scale_y_discrete(limits = rev) + heatmap_theme +
  labs(title    = "Cabezon Larval Observations by Station and Season",
       subtitle = "Positive tows only; white = no positive tows recorded",
       x = NULL, y = "South \u2190 Station \u2192 North")

```

# 9. sdmTMB MODEL FITTING (CONSTRAINED)

```{r 9. sdmTMB MODEL FITTING (CONSTRAINED), cache = T}

set.seed(my.seed)

#---- 9.a. Build mesh
if (!all(c("X", "Y") %in% names(cabezon_shelf_spawn))) {
  cabezon_shelf_spawn <- sdmTMB::add_utm_columns(cabezon_shelf_spawn, ll_names = c("longitude", "latitude"))
}

ss_mesh <- fm_mesh_2d(
  loc = cabezon_shelf_spawn[,c("X","Y")],
  cutoff = 20,
  max.edge = c(75, 150),
  offset = c(45, 120)
)
ss_cab_mesh <- make_mesh(data = cabezon_shelf_spawn, c("X", "Y"), mesh = ss_mesh)
plot(ss_cab_mesh); title("SPDE Mesh for Cabezon CalCOFI analysis (Constrained)")

#---- 9.b. Fit candidate models

# Add year factor variable
cabezon_shelf_spawn$fyear <- as.factor(cabezon_shelf_spawn$year)

# NULL Delta-Gamma model

ss_cab_null <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = "off",
  spatiotemporal = "off",
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_null)

# Delta-Gamma: fixed year effect + iid: 
ss_cab_fit01 <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("iid", "iid"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit01)

# Delta-Gamma: fixed year effect - spatiotemporal: # Is this the best fit for a fixed year effect? Our sanity checks tell us that in models with both a fixed year effect and spatiotemporal fields cause the fields to collapse, suggesting maybe that any temporal variation is adequately captured by the fixed year effect already.
ss_cab_fit02 <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("off", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit02)

# Delta-Gamma: fixed year effect + iid delta:
ss_cab_fit03 <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("iid", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit03)

# Delta-Gamma: full AR1 spatial + spatiotemporal 
ss_cab_fit1 <- sdmTMB(
  list(larvae_100m3 ~ 1, larvae_100m3 ~ 1),
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("ar1", "ar1"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit1)

# Delta-Gamma: spatial only, spatiotemporal off
ss_cab_fit1a <- sdmTMB(
  list(larvae_100m3 ~ 1, larvae_100m3 ~ 1),
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("off", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit1a)

# Delta-Gamma; Delta component AR1, Gamma spatiotemporal off
ss_cab_fit1b <- sdmTMB(
  list(larvae_100m3 ~ 1, larvae_100m3 ~ 1),
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("ar1", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit1b)

# Delta-Gamma: Gamma component AR1, Delta spatiotemporal off
ss_cab_fit1c <- sdmTMB(
  list(larvae_100m3 ~ 1, larvae_100m3 ~ 1),
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("off", "ar1"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit1c)

# Tweedie: full AR1 spatial + spatiotemporal
ss_cab_fit2 <- sdmTMB(
  formula = larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = tweedie(),
  spatial = "on",
  spatiotemporal = "ar1",
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit2)

# Tweedie: spatial only, spatiotemporal off
ss_cab_fit2a <- sdmTMB(
  formula = larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = tweedie(),
  spatial = "on",
  spatiotemporal = "off",
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit2a)

#---- 9.c. Construct model comparison table

AIC_table <- data.frame(
  fit_name = c(
    "ss_cab_null",
    "ss_cab_fit01",
    "ss_cab_fit02",
    "ss_cab_fit03",
    "ss_cab_fit1",
    "ss_cab_fit1a",
    "ss_cab_fit1b",
    "ss_cab_fit1c",
    "ss_cab_fit2",
    "ss_cab_fit2a"
  ),
  model = c(
    "Delta-Gamma NULL (Constrained)",
    "Delta-Gamma spatial + spatiotemporal IID + fixed year effect (Constrained)",
    "Delta-Gamma spatial + fixed year effect - spatiotemporal (Constrained)",
    "Delta-Gamma: fixed year effect + iid delta (Constrained)",
    "Delta-Gamma spatial + spatiotemporal AR1 (Constrained)",
    "Delta-Gamma spatial (Constrained)",
    "Delta-Gamma spatiotemporal Delta AR1 (Constrained)",
    "Delta-Gamma spatiotemporal Gamma AR1 (Constrained)",
    "Tweedie spatial + spatiotemporal AR1 (Constrained)",
    "Tweedie spatial (Constrained)"
  ),
  AIC = c(
    AIC(ss_cab_null),
    AIC(ss_cab_fit01),
    AIC(ss_cab_fit02),
    AIC(ss_cab_fit03),
    AIC(ss_cab_fit1),
    AIC(ss_cab_fit1a),
    AIC(ss_cab_fit1b),
    AIC(ss_cab_fit1c),
    AIC(ss_cab_fit2),
    AIC(ss_cab_fit2a)
  ),
    delta_AIC = NA
  )

AIC_table$delta_AIC <- AIC_table$AIC - min(AIC_table$AIC)

AIC_table <- AIC_table[order(AIC_table$AIC, decreasing = F), ]

AIC_table

#---- 9.d. Sanity checks and random effect parameter estimates for best models

sanity(ss_cab_fit1)
sanity(ss_cab_fit01)
sanity(ss_cab_fit2)

tidy(ss_cab_fit1, effects = "ran_pars")
tidy(ss_cab_fit01, effects = "ran_pars")
tidy(ss_cab_fit2, effects = "ran_pars")


```


# 10. MODEL DIAGNOSTICS & ABUNDANCE INDICES (CONSTRAINED)

## 10.1. ss_cab_fit1 — Full Delta-Gamma Model

```{r 10.1. ss_cab_fit1 diagnostics, cache = T}

#---- 10.1.a. Calibration summaries

ss_full_pred_obs <- predict(ss_cab_fit1, type = "response")

# Overall mean calibration
ss_full_pred_obs %>%
  summarise(
    obs_mean = mean(larvae_100m3, na.rm = TRUE),
    pred_mean = mean(est, na.rm = TRUE),
    obs_occ = mean(larvae_100m3 > 0, na.rm = TRUE),
    pred_occ = mean(est1, na.rm = TRUE),
    obs_pos_mean = mean(larvae_100m3[larvae_100m3 > 0], na.rm = TRUE),
    pred_pos_mean = mean(est2, na.rm = TRUE)
  )

# Binomial calibration; i.e. predicted presence probability vs. observed presence rate
ss_full_pred_obs$presence <- ss_full_pred_obs$larvae_100m3 > 0

ss_full_calibration_summary <- ss_full_pred_obs %>%
  mutate(prob_bin = cut(est1, breaks = seq(0, 1, by = 0.1))) %>%
  group_by(prob_bin) %>%
  summarise(
    mean_pred = mean(est1),
    observed = mean(presence),
    n = n()
  )

plot(ss_full_calibration_summary$mean_pred, ss_full_calibration_summary$observed,
     xlab = "Predicted presence probability",
     ylab = "Observed presence frequency")
abline(0,1,lty=2)

#---- 10.1.b. Predicted vs. observed plots

# Raw scale comparison
plot(
  x = ss_full_pred_obs$est,
  y = cabezon_shelf_spawn$larvae_100m3,
  xlab = "Predicted",
  ylab = "Observed"
)
abline(0, 1, lty = 2)

# Log scale comparison
plot(log1p(ss_full_pred_obs$est), log1p(cabezon_shelf_spawn$larvae_100m3),
     xlab = "log(Predicted + 1)", ylab = "log(Observed + 1)")
abline(0, 1, lty = 2)

# Gamma (positive density) component only
pos_idx <- cabezon_shelf_spawn$larvae_100m3 > 0
plot(log(ss_full_pred_obs$est2[pos_idx]), log(cabezon_shelf_spawn$larvae_100m3[pos_idx]),
     xlab = "log(Predicted positive abundance)",
     ylab = "log(Observed positive abundance)")
abline(0, 1, lty = 2)

#---- 10.1.c. Randomized quantile residuals

ss_full_rq_res <- residuals(ss_cab_fit1, type = "mle-mvn")

# QQ plot
qqnorm(ss_full_rq_res)
qqline(ss_full_rq_res)

# Spatial residual map by year (kind of hard to see)
ss_full_pred_obs$rq_resid <- ss_full_rq_res

ggplot(ss_full_pred_obs, aes(x = longitude, y = latitude, color = rq_resid)) +
  geom_point() +
  scale_color_gradient2() +
  facet_wrap(~ year)

#---- 10.1.d. Temporal calibration

# Are predictions tracking temporal trends?
ss_full_pred_obs %>%
  group_by(year) %>%
  summarize(
    obs = mean(larvae_100m3),
    pred = mean(est)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = obs), color = "black") +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed")

#---- 10.1.e. Abundance index construction
# Note! The index uncertainty here was far too wide using a regular spatial grid, probably because interpolation over years with no CalCOFI coverage was causing inflation. I kept that plot in for reference, but 10.1.f. (observed station grid) is probably preferred for diagnosis.

ss_full_pred_grid <- expand.grid(
  X = seq(min(cabezon_shelf_spawn$X), max(cabezon_shelf_spawn$X), by = 5),
  Y = seq(min(cabezon_shelf_spawn$Y), max(cabezon_shelf_spawn$Y), by = 5)
) %>%
  tidyr::crossing(year = as.integer(sort(unique(cabezon_shelf_spawn$year)))) %>%
  mutate(fyear = as.factor(year))

ss_full_cab_map <- predict(ss_cab_fit1, newdata = ss_full_pred_grid, return_tmb_object = T)
ss_full_cab_index <- get_index(ss_full_cab_map, area = 5^2, bias_correct = T)

ggplot(ss_full_cab_index, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year", title = "Abundance Index: Constrained to Spawning Season & Coastal Samples")

#---- 10.1.f. Abundance index: constrained to observed station grid (probably preferred)

ss_full_pred_grid_obs <- cabezon_shelf_spawn %>%
  dplyr::select(X, Y, year) %>%
  distinct()

ss_full_cab_map_obs <- predict(ss_cab_fit1,
                       newdata = ss_full_pred_grid_obs,
                       return_tmb_object = T)

ss_full_cab_index_obs <- get_index(ss_full_cab_map_obs,
                           area = 5^2,
                           bias_correct = T)

ggplot(ss_full_cab_index_obs, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year", title = "Abundance Index: Observed Station Grid")
```

## 10.2. ss_cab_fit1 — STAR Comparisons & CCFs

```{r 10.2. ss_cab_fit1 STAR comparisons full extrapolation, cache = T}
#---- 10.2. ss_cab_fit1 STAR comparisons and CCF

# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 10.1.f. abundance index, assuming it's more stable.
ss_full_combined <- ss_full_cab_index %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
ss_full_combined_obs <- ss_full_combined %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf(ss_full_combined_obs$larvae_index, ss_full_combined_obs$age0,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Age-0 Recruits")

# CCF: Restricted larval index vs. recruitment deviations
ccf(ss_full_combined_obs$larvae_index, ss_full_combined_obs$rec_dev,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Recruitment Deviations")

# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf(ss_full_combined_obs$larvae_index, ss_full_combined_obs$ssb,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Spawning Stock Biomass (raw)")

larvae_detrended <- residuals(loess(larvae_index ~ year, data = ss_full_combined_obs, span = 0.75))
ssb_detrended    <- residuals(loess(ssb ~ year,          data = ss_full_combined_obs, span = 0.75))

ccf(larvae_detrended, ssb_detrended, 
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. SSB (detrended)")
```

```{r 10.2.2. ss_cab_fit1 STAR comparisons station extrapolation, cache = T}
#---- 10.2. ss_cab_fit1 STAR comparisons and CCF

# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 10.1.f. abundance index, assuming it's more stable.
ss_full_combined <- ss_full_cab_index_obs %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
ss_full_combined_obs <- ss_full_combined %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf(ss_full_combined_obs$larvae_index, ss_full_combined_obs$age0,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Age-0 Recruits")

# CCF: Restricted larval index vs. recruitment deviations
ccf(ss_full_combined_obs$larvae_index, ss_full_combined_obs$rec_dev,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Recruitment Deviations")

# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf(ss_full_combined_obs$larvae_index, ss_full_combined_obs$ssb,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Spawning Stock Biomass (raw)")

larvae_detrended <- residuals(loess(larvae_index ~ year, data = ss_full_combined_obs, span = 0.75))
ssb_detrended    <- residuals(loess(ssb ~ year,          data = ss_full_combined_obs, span = 0.75))

ccf(larvae_detrended, ssb_detrended, 
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. SSB (detrended)")
```

## 10.3. ss_cab_fit1 — Correlation Testing

```{r 10.3. ss_cab_fit1 correlation testing, cache = T}
# RQ residuals for full AR1 spatiotemporal model
ss_full_pred_obs %>%
  group_by(year) %>%
  mutate(med_resid = median(rq_resid)) %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method  = "loess", span = 0.4,
    se      = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Year",
    y     = "RQ Residual",
    title = "Annual RQ Residual Distributions — cab_fit1"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

ss_full_pred_obs %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x     = "Year",
    y     = "Cumulative Sum of Median RQ Residual",
    title = "CUSUM — cab_fit1"
  ) +
  theme_minimal()


#---- Simple lagged regression at -1 to check

ss_full_combined_lagged <- ss_full_combined_obs %>%
  arrange(year) %>%
  mutate(
    age0_lead1 = lead(age0, 1),
    rec_dev_lead1 = lead(rec_dev, 1)
  )

ss_full_lm_age0 <- lm(age0_lead1 ~ larvae_index, data = ss_full_combined_lagged)
ss_full_lm_recdev <- lm(rec_dev_lead1 ~ larvae_index, data = ss_full_combined_lagged)

summary(ss_full_lm_age0)
summary(ss_full_lm_recdev)

ss_full_combined_lagged %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = T) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

ss_full_combined_lagged_no83 <- ss_full_combined_lagged %>% filter(year != 1983)

ss_full_lm_age0_no83 <- lm(age0_lead1 ~ larvae_index, data = ss_full_combined_lagged_no83)
summary(ss_full_lm_age0_no83)

ss_full_lm_age0_detrend   <- lm(age0_lead1    ~ larvae_index + year, data = ss_full_combined_lagged)
ss_full_lm_recdev_detrend <- lm(rec_dev_lead1 ~ larvae_index + year, data = ss_full_combined_lagged)

summary(ss_full_lm_age0_detrend)
summary(ss_full_lm_recdev_detrend)

plot(ss_full_lm_age0, which = 4)  # Cook's distance plot

# Remove the three most anomalous years
ss_full_combined_lagged_trimmed <- ss_full_combined_lagged %>% 
  filter(!year %in% c(1983, 1986, 1998))

ss_full_combined_lagged_trimmed %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = T) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

ss_full_lm_age0_trimmed <- lm(age0_lead1 ~ larvae_index, data = ss_full_combined_lagged_trimmed)
summary(ss_full_lm_age0_trimmed)

ss_full_lm_age0_trimmed_detrend <- lm(age0_lead1 ~ larvae_index + year, data = ss_full_combined_lagged_trimmed)
summary(ss_full_lm_age0_trimmed_detrend)


# 1. Is the underprediction concentrated in one delta component?
# Separate RQ residuals by model component
ss_full_pred_obs <- ss_full_pred_obs %>%
  mutate(
    rq_resid1 = residuals(ss_cab_fit1, model = 1, type = "mle-mvn"),
    rq_resid2 = residuals(ss_cab_fit1, model = 2, type = "mle-mvn")
  )

# CUSUM for each component separately
ss_full_pred_obs %>%
  group_by(year) %>%
  summarise(
    med_resid1 = median(rq_resid1, na.rm = TRUE),
    med_resid2 = median(rq_resid2, na.rm = TRUE),  # positive tows only
    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    cusum1 = cumsum(replace_na(med_resid1, 0)),
    cusum2 = cumsum(replace_na(med_resid2, 0))
  ) %>%
  pivot_longer(cols = c(cusum1, cusum2),
               names_to = "component",
               values_to = "cusum") %>%
  ggplot(aes(x = year, y = cusum, color = component)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("cusum1" = "steelblue", "cusum2" = "firebrick"),
                     labels = c("Delta (binomial)", "Gamma")) +
  labs(x = "Year", y = "CUSUM", title = "CUSUM by Delta Model Component — ss_cab_fit1") +
  theme_minimal()

# 2. Is the underprediction spatially structured?
ss_full_pred_obs %>%
  group_by(station) %>%
  summarise(mean_resid = mean(rq_resid), .groups = "drop") %>%
  ggplot(aes(x = station, y = mean_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = TRUE) +
  labs(x = "Station", y = "Mean RQ Residual",
       title = "Mean Residual by Station — ss_cab_fit1") +
  theme_minimal()

# 3. AIC comparison: ss_cab_fit1 vs. ss_cab_fit1c (Gamma AR1 only)
AIC(ss_cab_fit1, ss_cab_fit1c)

#---- Breakpoint analysis: ss_cab_fit1 (constrained, AR1 spatiotemporal) ----

# ── Build annual median residual series ──

# Combined (overall)
ss_fit1_cusum <- ss_full_pred_obs %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year)

# Component-level
ss_fit1_cusum_components <- ss_full_pred_obs %>%
  group_by(year) %>%
  summarise(
    med_resid1 = median(rq_resid1, na.rm = TRUE),
    med_resid2 = median(rq_resid2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year)

# ── Segmented regression: combined CUSUM — one breakpoint ──
lm_ss_fit1       <- lm(med_resid ~ year, data = ss_fit1_cusum)
seg_ss_fit1      <- segmented(lm_ss_fit1, seg.Z = ~year, npsi = 1)
summary(seg_ss_fit1)
confint(seg_ss_fit1)

# ── Segmented regression: Gamma component — one breakpoint ──
# Gamma is the dominant signal; focus breakpoint search here
lm_ss_fit1_gamma  <- lm(med_resid2 ~ year, data = ss_fit1_cusum_components)
seg_ss_fit1_gamma <- segmented(lm_ss_fit1_gamma, seg.Z = ~year, npsi = 1)
summary(seg_ss_fit1_gamma)
confint(seg_ss_fit1_gamma)

# ── Segmented regression: Gamma component — two breakpoints ──
# Try two in case the early 1987-88 peak and mid-1990s descent are distinct
seg_ss_fit1_gamma2 <- segmented(lm_ss_fit1_gamma, seg.Z = ~year, npsi = 2)
summary(seg_ss_fit1_gamma2)
confint(seg_ss_fit1_gamma2)

run_era_ccf <- function(df, ss_var, start_yr, end_yr, lag.max = 10) {
  sub <- df %>% filter(year >= start_yr, year <= end_yr) %>% arrange(year)
  cat("Era:", start_yr, "–", end_yr, "| n rows:", nrow(sub), "\n")
  print(sub %>% dplyr::select(year, larvae_index, all_of(ss_var)))
  ccf(sub$larvae_index, sub[[ss_var]], lag.max = lag.max,
      main = paste0("CCF: ", start_yr, "–", end_yr, " | ", ss_var))
}

# ── Extract breakpoints and run era-stratified CCFs ──
bp_ss_fit1 <- round(seg_ss_fit1$psi[, "Est."])
bp_ss_fit1_gamma <- round(seg_ss_fit1_gamma$psi[, "Est."])

cat("Combined CUSUM breakpoint:", bp_ss_fit1, "\n")
cat("Gamma CUSUM breakpoint:", bp_ss_fit1_gamma, "\n")

# Era-stratified CCFs using combined breakpoint
# Two eras defined by single breakpoint
run_era_ccf(ss_full_combined_obs, "age0",    1981, bp_ss_fit1 - 1)
run_era_ccf(ss_full_combined_obs, "age0",    bp_ss_fit1, 2015)

run_era_ccf(ss_full_combined_obs, "rec_dev", 1981, bp_ss_fit1 - 1)
run_era_ccf(ss_full_combined_obs, "rec_dev", bp_ss_fit1, 2015)

# Era-stratified CCFs using Gamma breakpoint
run_era_ccf(ss_full_combined_obs, "age0",    1981, bp_ss_fit1_gamma - 1)
run_era_ccf(ss_full_combined_obs, "age0",    bp_ss_fit1_gamma, 2015)

run_era_ccf(ss_full_combined_obs, "rec_dev", 1981, bp_ss_fit1_gamma - 1)
run_era_ccf(ss_full_combined_obs, "rec_dev", bp_ss_fit1_gamma, 2015)
```

## 10.4. ss_cab_fit01 — Fixed Year Effects Delta-Gamma Model

```{r 10.4. ss_cab_fit01 diagnostics, cache = T}

#---- 10.4.a. Calibration summaries

ss_fy_pred_obs <- predict(ss_cab_fit01, type = "response")

# Overall mean calibration
ss_fy_pred_obs %>%
  summarise(
    obs_mean     = mean(larvae_100m3, na.rm = TRUE),
    pred_mean    = mean(est, na.rm = TRUE),
    obs_occ      = mean(larvae_100m3 > 0, na.rm = TRUE),
    pred_occ     = mean(est1, na.rm = TRUE),
    obs_pos_mean = mean(larvae_100m3[larvae_100m3 > 0], na.rm = TRUE),
    pred_pos_mean = mean(est2, na.rm = TRUE)
  )

# Binomial calibration
ss_fy_pred_obs$presence <- ss_fy_pred_obs$larvae_100m3 > 0

ss_fy_calibration_summary <- ss_fy_pred_obs %>%
  mutate(prob_bin = cut(est1, breaks = seq(0, 1, by = 0.1))) %>%
  group_by(prob_bin) %>%
  summarise(
    mean_pred = mean(est1),
    observed  = mean(presence),
    n         = n()
  )

plot(ss_fy_calibration_summary$mean_pred, ss_fy_calibration_summary$observed,
     xlab = "Predicted presence probability",
     ylab = "Observed presence frequency")
abline(0, 1, lty = 2)

#---- 10.4.b. Predicted vs. observed plots

# Raw scale
plot(
  x = ss_fy_pred_obs$est,
  y = cabezon_shelf_spawn$larvae_100m3,
  xlab = "Predicted",
  ylab = "Observed"
)
abline(0, 1, lty = 2)

# Log scale
plot(log1p(ss_fy_pred_obs$est), log1p(cabezon_shelf_spawn$larvae_100m3),
     xlab = "log(Predicted + 1)", ylab = "log(Observed + 1)")
abline(0, 1, lty = 2)

# Gamma (positive density) component only
pos_idx <- cabezon_shelf_spawn$larvae_100m3 > 0
plot(log(ss_fy_pred_obs$est2[pos_idx]), log(cabezon_shelf_spawn$larvae_100m3[pos_idx]),
     xlab = "log(Predicted positive abundance)",
     ylab = "log(Observed positive abundance)")
abline(0, 1, lty = 2)

#---- 10.4.c. Randomized quantile residuals

ss_fy_rq_res <- residuals(ss_cab_fit01, type = "mle-mvn")

# QQ plot
qqnorm(ss_fy_rq_res)
qqline(ss_fy_rq_res)

# Spatial residual map by year
ss_fy_pred_obs$rq_resid <- ss_fy_rq_res

ggplot(ss_fy_pred_obs, aes(x = longitude, y = latitude, color = rq_resid)) +
  geom_point() +
  scale_color_gradient2() +
  facet_wrap(~ year)

#---- 10.4.d. Temporal calibration

ss_fy_pred_obs %>%
  group_by(year) %>%
  summarize(
    obs  = mean(larvae_100m3),
    pred = mean(est)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = obs),  color = "black") +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed")

#---- 10.4.e. Abundance index: regular spatial grid

ss_fy_pred_grid <- expand.grid(
  X = seq(min(cabezon_shelf_spawn$X), max(cabezon_shelf_spawn$X), by = 5),
  Y = seq(min(cabezon_shelf_spawn$Y), max(cabezon_shelf_spawn$Y), by = 5)
) %>%
  tidyr::crossing(year = as.integer(sort(unique(cabezon_shelf_spawn$year)))) %>%
  mutate(fyear = as.factor(year))

ss_fy_cab_map   <- predict(ss_cab_fit01, newdata = ss_fy_pred_grid, return_tmb_object = TRUE)
ss_fy_cab_index <- get_index(ss_fy_cab_map, area = 5^2, bias_correct = TRUE)

ggplot(ss_fy_cab_index, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Constrained to Spawning Season & Coastal Samples")

#---- 10.4.f. Abundance index: observed station grid (preferred)

ss_fy_pred_grid_obs <- cabezon_shelf_spawn %>%
  dplyr::select(X, Y, year, fyear) %>%
  distinct()

ss_fy_cab_map_obs   <- predict(ss_cab_fit01,
                               newdata = ss_fy_pred_grid_obs,
                               return_tmb_object = TRUE)

ss_fy_cab_index_obs <- get_index(ss_fy_cab_map_obs,
                                 area = 5^2,
                                 bias_correct = TRUE)

ggplot(ss_fy_cab_index_obs, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Observed Station Grid")

```

## 10.5. ss_cab_fit01 — STAR Comparisons & CCFs

```{r 10.5. ss_cab_fit01 STAR comparisons full extrapolation, cache = T}

#---- 10.5. ss_cab_fit01 STAR comparisons and CCF

# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 10.4.f. abundance index, assuming it's more stable.
ss_fy_combined <- ss_fy_cab_index %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
ss_fy_combined_obs <- ss_fy_combined %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf(ss_fy_combined_obs$larvae_index, ss_fy_combined_obs$age0,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Age-0 Recruits")

# CCF: Restricted larval index vs. recruitment deviations
ccf(ss_fy_combined_obs$larvae_index, ss_fy_combined_obs$rec_dev,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Recruitment Deviations")

# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf(ss_fy_combined_obs$larvae_index, ss_fy_combined_obs$ssb,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Spawning Stock Biomass (raw)")

ss_fy_larvae_detrended <- residuals(loess(larvae_index ~ year, data = ss_fy_combined_obs, span = 0.75))
ss_fy_ssb_detrended    <- residuals(loess(ssb ~ year,          data = ss_fy_combined_obs, span = 0.75))

ccf(ss_fy_larvae_detrended, ss_fy_ssb_detrended, 
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. SSB (detrended)")

```

```{r 10.5.2. ss_cab_fit01 STAR comparisons station extrapolation, cache = T}

#---- 10.5. ss_cab_fit01 STAR comparisons and CCF

# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 10.4.f. abundance index, assuming it's more stable.
ss_fy_combined <- ss_fy_cab_index_obs %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
ss_fy_combined_obs <- ss_fy_combined %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf(ss_fy_combined_obs$larvae_index, ss_fy_combined_obs$age0,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Age-0 Recruits")

# CCF: Restricted larval index vs. recruitment deviations
ccf(ss_fy_combined_obs$larvae_index, ss_fy_combined_obs$rec_dev,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Recruitment Deviations")

# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf(ss_fy_combined_obs$larvae_index, ss_fy_combined_obs$ssb,
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. Spawning Stock Biomass (raw)")

ss_fy_larvae_detrended <- residuals(loess(larvae_index ~ year, data = ss_fy_combined_obs, span = 0.75))
ss_fy_ssb_detrended    <- residuals(loess(ssb ~ year,          data = ss_fy_combined_obs, span = 0.75))

ccf(ss_fy_larvae_detrended, ss_fy_ssb_detrended, 
    lag.max = 10,
    main = "CCF: Restricted Larval Index vs. SSB (detrended)")

```

## 10.6. ss_cab_fit01 — Correlation Testing

```{r 10.6. ss_cab_fit01 correlation testing, cache = T}
# RQ residuals for fixed year effect + IID model
ss_fy_pred_obs %>%
  group_by(year) %>%
  mutate(med_resid = median(rq_resid)) %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method  = "loess", span = 0.4,
    se      = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Year",
    y     = "RQ Residual",
    title = "Annual RQ Residual Distributions — ss_cab_fit01"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

ss_fy_pred_obs %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x     = "Year",
    y     = "Cumulative Sum of Median RQ Residual",
    title = "CUSUM — ss_cab_fit01"
  ) +
  theme_minimal()

#---- Simple lagged regression at lag -1

ss_fy_combined_lagged <- ss_fy_combined_obs %>%
  arrange(year) %>%
  mutate(
    age0_lead1    = lead(age0, 1),
    rec_dev_lead1 = lead(rec_dev, 1)
  )

ss_fy_lm_age0   <- lm(age0_lead1    ~ larvae_index, data = ss_fy_combined_lagged)
ss_fy_lm_recdev <- lm(rec_dev_lead1 ~ larvae_index, data = ss_fy_combined_lagged)

summary(ss_fy_lm_age0)
summary(ss_fy_lm_recdev)

ss_fy_combined_lagged %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

# Sensitivity: remove 1983
ss_fy_combined_lagged_no83 <- ss_fy_combined_lagged %>% filter(year != 1983)
ss_fy_lm_age0_no83 <- lm(age0_lead1 ~ larvae_index, data = ss_fy_combined_lagged_no83)
summary(ss_fy_lm_age0_no83)

# Detrended regressions
ss_fy_lm_age0_detrend   <- lm(age0_lead1    ~ larvae_index + year, data = ss_fy_combined_lagged)
ss_fy_lm_recdev_detrend <- lm(rec_dev_lead1 ~ larvae_index + year, data = ss_fy_combined_lagged)
summary(ss_fy_lm_age0_detrend)
summary(ss_fy_lm_recdev_detrend)

# Cook's distance
plot(ss_fy_lm_age0, which = 4)

# Trimmed analysis: remove three most anomalous years
ss_fy_combined_lagged_trimmed <- ss_fy_combined_lagged %>%
  filter(!year %in% c(1983, 1986, 1998))

ss_fy_combined_lagged_trimmed %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

ss_fy_lm_age0_trimmed <- lm(age0_lead1 ~ larvae_index, data = ss_fy_combined_lagged_trimmed)
summary(ss_fy_lm_age0_trimmed)

ss_fy_lm_age0_trimmed_detrend <- lm(age0_lead1 ~ larvae_index + year, data = ss_fy_combined_lagged_trimmed)
summary(ss_fy_lm_age0_trimmed_detrend)

# 1. Is the underprediction concentrated in one delta component?
ss_fy_pred_obs <- ss_fy_pred_obs %>%
  mutate(
    rq_resid1 = residuals(ss_cab_fit01, model = 1, type = "mle-mvn"),
    rq_resid2 = residuals(ss_cab_fit01, model = 2, type = "mle-mvn")
  )

ss_fy_pred_obs %>%
  group_by(year) %>%
  summarise(
    med_resid1 = median(rq_resid1, na.rm = TRUE),
    med_resid2 = median(rq_resid2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    cusum1 = cumsum(replace_na(med_resid1, 0)),
    cusum2 = cumsum(replace_na(med_resid2, 0))
  ) %>%
  pivot_longer(cols = c(cusum1, cusum2),
               names_to  = "component",
               values_to = "cusum") %>%
  ggplot(aes(x = year, y = cusum, color = component)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("cusum1" = "steelblue", "cusum2" = "firebrick"),
                     labels = c("cusum1" = "Delta (binomial)", "cusum2" = "Gamma")) +
  labs(x = "Year", y = "CUSUM",
       title = "CUSUM by Delta Model Component — ss_cab_fit01") +
  theme_minimal()

# 2. Is the underprediction spatially structured?
ss_fy_pred_obs %>%
  group_by(station) %>%
  summarise(mean_resid = mean(rq_resid), .groups = "drop") %>%
  ggplot(aes(x = station, y = mean_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = TRUE) +
  labs(x = "Station", y = "Mean RQ Residual",
       title = "Mean Residual by Station — ss_cab_fit01") +
  theme_minimal()

# 3. AIC comparison: AR1 vs. fixed year effect (constrained dataset)
AIC(ss_cab_fit01, ss_cab_fit1, ss_cab_fit1c)



```



---

# PART III — APPENDIX / ADDITIONAL MODELS

---

# A1. TWEEDIE MODEL — cab_fit2

```{r A1. cab_fit2 Tweedie, eval = F}
#---- A1. cab_fit2 Tweedie — Diagnostics & Abundance Index
#---- A1.a. Calibration summaries
pred_obs_tw <- predict(cab_fit2, type = "response")

pred_obs_tw %>%
  summarise(
    obs_mean  = mean(cabezon$larvae_100m3, na.rm = TRUE),
    pred_mean = mean(est, na.rm = TRUE)
  )

# Note: Tweedie is single-component — binomial calibration and gamma plots are skipped

#---- A1.b. Predicted vs. observed
plot(pred_obs_tw$est, cabezon$larvae_100m3,
     xlab = "Predicted", ylab = "Observed",
     main = "Raw scale — cab_fit2 Tweedie")
abline(0, 1, lty = 2)

plot(log1p(pred_obs_tw$est), log1p(cabezon$larvae_100m3),
     xlab = "log(Predicted + 1)", ylab = "log(Observed + 1)",
     main = "Log scale — cab_fit2 Tweedie")
abline(0, 1, lty = 2)

#---- A1.c. Randomized quantile residuals
rq_res_tw <- residuals(cab_fit2, type = "mle-mvn")

qqnorm(rq_res_tw, main = "QQ — cab_fit2 Tweedie")
qqline(rq_res_tw)

pred_obs_tw$rq_resid <- rq_res_tw

ggplot(pred_obs_tw, aes(x = X, y = Y, color = rq_resid)) +
  geom_point() +
  scale_color_gradient2() +
  facet_wrap(~ year) +
  labs(title = "Spatial residuals — cab_fit2 Tweedie")

#---- A1.d. Temporal calibration
pred_obs_tw %>%
  group_by(year) %>%
  summarize(
    obs  = mean(larvae_100m3),
    pred = mean(est)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = obs), color = "black") +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed") +
  labs(title = "Temporal calibration — cab_fit2 Tweedie")

#---- A1.e. Annual RQ distributions
pred_obs_tw %>%
  group_by(year) %>%
  mutate(med_resid = median(rq_resid)) %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method = "loess", span = 0.4,
    se = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Year", y = "RQ Residual",
       title = "Annual RQ Residual Distributions — cab_fit2 Tweedie") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

#---- A1.f. CUSUM
pred_obs_tw %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(x = "Year", y = "Cumulative Sum of Median RQ Residual",
       title = "CUSUM — cab_fit2 Tweedie") +
  theme_minimal()

#---- A1.g. Abundance index: observed station grid
pred_grid_obs_tw <- cabezon %>%
  dplyr::select(X, Y, year) %>%
  distinct()

cab_map_obs_tw  <- predict(cab_fit2,
                           newdata = pred_grid_obs_tw,
                           return_tmb_object = TRUE)
cab_index_obs_tw <- get_index(cab_map_obs_tw, area = 5^2, bias_correct = TRUE)

ggplot(cab_index_obs_tw, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Observed Station Grid — cab_fit2 Tweedie")



#---- 7.7. cab_fit2 TWEEDIE — STAR COMPARISONS & CCFs
combined_tw <- cab_index_obs_tw %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

combined_tw_obs <- combined_tw %>%
  filter(!is.na(larvae_index))

#---- A1.a. Pooled CCFs
ccf(combined_tw_obs$larvae_index, combined_tw_obs$age0,
    lag.max = 10,
    main = "CCF: Tweedie Larval Index vs. Age-0 Recruits")

ccf(combined_tw_obs$larvae_index, combined_tw_obs$rec_dev,
    lag.max = 10,
    main = "CCF: Tweedie Larval Index vs. Recruitment Deviations")

ccf(combined_tw_obs$larvae_index, combined_tw_obs$ssb,
    lag.max = 10,
    main = "CCF: Tweedie Larval Index vs. SSB (raw)")

larvae_detrended_tw <- residuals(lm(larvae_index ~ year, data = combined_tw_obs))
ssb_detrended_tw    <- residuals(lm(ssb ~ year,          data = combined_tw_obs))

ccf(larvae_detrended_tw, ssb_detrended_tw,
    lag.max = 10,
    main = "CCF: Tweedie Larval Index vs. SSB (detrended)")

#---- A1.STAR.b. Breakpoint analysis
cusum_tw <- pred_obs_tw %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year)

lm_tw  <- lm(med_resid ~ year, data = cusum_tw)
seg_tw <- segmented(lm_tw, seg.Z = ~year, npsi = 1)
summary(seg_tw)
confint(seg_tw)
```

# A2. ALTERNATIVE MODEL — cab_fit03

```{r A2. cab_fit03, eval = F}
#---- A2. cab_fit03 DG Fixed Year / IID — Diagnostics & Abundance Index
#---- A2.a. Calibration summaries
pred_obs_iid <- predict(cab_fit03, type = "response")

pred_obs_iid %>%
  summarise(
    obs_mean      = mean(cabezon$larvae_100m3, na.rm = TRUE),
    pred_mean     = mean(est, na.rm = TRUE),
    obs_occ       = mean(cabezon$larvae_100m3 > 0, na.rm = TRUE),
    pred_occ      = mean(est1, na.rm = TRUE),
    obs_pos_mean  = mean(cabezon$larvae_100m3[cabezon$larvae_100m3 > 0], na.rm = TRUE),
    pred_pos_mean = mean(est2, na.rm = TRUE)
  )

pred_obs_iid$presence <- cabezon$larvae_100m3 > 0
calibration_summary_iid <- pred_obs_iid %>%
  mutate(prob_bin = cut(est1, breaks = seq(0, 1, by = 0.1))) %>%
  group_by(prob_bin) %>%
  summarise(mean_pred = mean(est1), observed = mean(presence), n = n())

plot(calibration_summary_iid$mean_pred, calibration_summary_iid$observed,
     xlab = "Predicted presence probability",
     ylab = "Observed presence frequency",
     main = "Binomial calibration — cab_fit03 DG IID")
abline(0, 1, lty = 2)

#---- A2.b. Predicted vs. observed
plot(pred_obs_iid$est, cabezon$larvae_100m3,
     xlab = "Predicted", ylab = "Observed",
     main = "Raw scale — cab_fit03 DG IID")
abline(0, 1, lty = 2)

plot(log1p(pred_obs_iid$est), log1p(cabezon$larvae_100m3),
     xlab = "log(Predicted + 1)", ylab = "log(Observed + 1)",
     main = "Log scale — cab_fit03 DG IID")
abline(0, 1, lty = 2)

pos_idx_iid <- cabezon$larvae_100m3 > 0
plot(log(pred_obs_iid$est2[pos_idx_iid]), log(cabezon$larvae_100m3[pos_idx_iid]),
     xlab = "log(Predicted positive abundance)",
     ylab = "log(Observed positive abundance)",
     main = "Gamma component — cab_fit03 DG IID")
abline(0, 1, lty = 2)

#---- A2.c. Randomized quantile residuals
rq_res_iid <- residuals(cab_fit03, type = "mle-mvn")

qqnorm(rq_res_iid, main = "QQ — cab_fit03 DG IID")
qqline(rq_res_iid)

pred_obs_iid$rq_resid <- rq_res_iid

ggplot(pred_obs_iid, aes(x = X, y = Y, color = rq_resid)) +
  geom_point() +
  scale_color_gradient2() +
  facet_wrap(~ year) +
  labs(title = "Spatial residuals — cab_fit03 DG IID")

#---- A2.d. Temporal calibration
pred_obs_iid %>%
  group_by(year) %>%
  summarize(
    obs  = mean(larvae_100m3),
    pred = mean(est)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = obs), color = "black") +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed") +
  labs(title = "Temporal calibration — cab_fit03 DG IID")

#---- A2.e. Annual RQ distributions
pred_obs_iid %>%
  group_by(year) %>%
  mutate(med_resid = median(rq_resid)) %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method = "loess", span = 0.4,
    se = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Year", y = "RQ Residual",
       title = "Annual RQ Residual Distributions — cab_fit03 DG IID") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

#---- A2.f. CUSUM
pred_obs_iid %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(x = "Year", y = "Cumulative Sum of Median RQ Residual",
       title = "CUSUM — cab_fit03 DG IID") +
  theme_minimal()

#---- A2.g. Abundance index: observed station grid
pred_grid_obs_iid <- cabezon %>%
  dplyr::select(X, Y, year) %>%
  distinct() %>%
  mutate(fyear = as.factor(year))

cab_map_obs_iid  <- predict(cab_fit03,
                            newdata = pred_grid_obs_iid,
                            return_tmb_object = TRUE)
cab_index_obs_iid <- get_index(cab_map_obs_iid, area = 5^2, bias_correct = TRUE)

ggplot(cab_index_obs_iid, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Observed Station Grid — cab_fit03 DG IID")


#---- A2.STAR. cab_fit03 DG IID — STAR COMPARISONS & CCFs
combined_iid <- cab_index_obs_iid %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

combined_iid_obs <- combined_iid %>%
  filter(!is.na(larvae_index))

#---- A2.STAR.a. Pooled CCFs
ccf(combined_iid_obs$larvae_index, combined_iid_obs$age0,
    lag.max = 10,
    main = "CCF: DG IID Larval Index vs. Age-0 Recruits")

ccf(combined_iid_obs$larvae_index, combined_iid_obs$rec_dev,
    lag.max = 10,
    main = "CCF: DG IID Larval Index vs. Recruitment Deviations")

ccf(combined_iid_obs$larvae_index, combined_iid_obs$ssb,
    lag.max = 10,
    main = "CCF: DG IID Larval Index vs. SSB (raw)")

larvae_detrended_iid <- residuals(lm(larvae_index ~ year, data = combined_iid_obs))
ssb_detrended_iid    <- residuals(lm(ssb ~ year,          data = combined_iid_obs))

ccf(larvae_detrended_iid, ssb_detrended_iid,
    lag.max = 10,
    main = "CCF: DG IID Larval Index vs. SSB (detrended)")

#---- A2.STAR.b. Breakpoint analysis
cusum_iid <- pred_obs_iid %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year)

lm_iid  <- lm(med_resid ~ year, data = cusum_iid)
seg_iid <- segmented(lm_iid, seg.Z = ~year, npsi = 1)
summary(seg_iid)
confint(seg_iid)
```


# A3. EXPERIMENTS

```{r A3. Experiments, eval = F}

# RQ residuals for full AR1 spatiotemporal model
pred_obs %>%
  group_by(year) %>%
  mutate(med_resid = median(rq_resid)) %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method  = "loess", span = 0.4,
    se      = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Year",
    y     = "RQ Residual",
    title = "Annual RQ Residual Distributions — cab_fit1"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

pred_obs %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x     = "Year",
    y     = "Cumulative Sum of Median RQ Residual",
    title = "CUSUM — cab_fit1"
  ) +
  theme_minimal()

# RQ residuals for spatial only, fixed effects model
pred_obs_sp %>%
  group_by(year) %>%
  mutate(med_resid = median(rq_resid)) %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method  = "loess", span = 0.4,
    se      = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Year",
    y     = "RQ Residual",
    title = "Annual RQ Residual Distributions — cab_fit01"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))


pred_obs_sp %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x     = "Year",
    y     = "Cumulative Sum of Median RQ Residual",
    title = "CUSUM — cab_fit01"
  ) +
  theme_minimal()

#---- Are there any meaningful breakpoints in our time-series to inform segmented regression analysis?

# Build the annual median series for each model
cusum_sp <- pred_obs_sp %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year)

cusum_ar1 <- pred_obs %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year)

# Fit a linear base model, then let segmented find the breakpoint
# Spatial model — one breakpoint
lm_sp  <- lm(med_resid ~ year, data = cusum_sp)
seg_sp <- segmented(lm_sp, seg.Z = ~year, npsi = 1)
summary(seg_sp)
confint(seg_sp)  # confidence interval on the breakpoint location

# AR1 model — two breakpoints
lm_ar1  <- lm(med_resid ~ year, data = cusum_ar1)
seg_ar1 <- segmented(lm_ar1, seg.Z = ~year, npsi = 2)
summary(seg_ar1)
confint(seg_ar1)


bp_sp <- round(seg_sp$psi[, "Est."])  # extract estimated breakpoint

run_era_ccf <- function(df, ss_var, start_yr, end_yr, lag.max = 10) {
  sub <- df %>% filter(year >= start_yr, year <= end_yr) %>% arrange(year)
  cat("Era:", start_yr, "–", end_yr, "| n rows:", nrow(sub), "\n")
  print(sub %>% dplyr::select(year, larvae_index, all_of(ss_var)))
  ccf(sub$larvae_index, sub[[ss_var]], lag.max = lag.max,
      main = paste0("CCF: ", start_yr, "–", end_yr, " | ", ss_var))
}

# AR1 model: three eras
bp1 <- round(seg_ar1$psi[1, "Est."])
bp2 <- round(seg_ar1$psi[2, "Est."])

str(combined_obs)

run_era_ccf(combined_obs, "rec_dev", 1981, bp1 - 1)
run_era_ccf(combined_obs, "rec_dev", bp1,  bp2 - 1)
run_era_ccf(combined_obs, "rec_dev", bp2,  2015)

#---- Simple lagged regression at -1 to check

combined_lagged <- combined_obs %>%
  arrange(year) %>%
  mutate(
    age0_lead1 = lead(age0, 1),
    rec_dev_lead1 = lead(rec_dev, 1)
  )

lm_age0 <- lm(age0_lead1 ~ larvae_index, data = combined_lagged)
lm_recdev <- lm(rec_dev_lead1 ~ larvae_index, data = combined_lagged)

summary(lm_age0)
summary(lm_recdev)

combined_lagged %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = T) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

combined_lagged_no83 <- combined_lagged %>% filter(year != 1983)

lm_age0_no83 <- lm(age0_lead1 ~ larvae_index, data = combined_lagged_no83)
summary(lm_age0_no83)

lm_age0_detrend   <- lm(age0_lead1    ~ larvae_index + year, data = combined_lagged)
lm_recdev_detrend <- lm(rec_dev_lead1 ~ larvae_index + year, data = combined_lagged)

summary(lm_age0_detrend)
summary(lm_recdev_detrend)

plot(lm_age0, which = 4)  # Cook's distance plot

# Remove the three most visually anomalous years
combined_lagged_trimmed <- combined_lagged %>% 
  filter(!year %in% c(1983, 1986, 2001))

lm_age0_trimmed <- lm(age0_lead1 ~ larvae_index, data = combined_lagged_trimmed)
summary(lm_age0_trimmed)
```

# A4. EXTRAS

```{r A4. Extras, error = TRUE, eval = F}

world <- ne_coastline(scale = "medium", returnclass = "sf")

# ── In-sample predictions ─────────────────────────────────────────────────────

# Delta-Gamma: est1 = occurrence prob, est2 = positive catch rate, combined = est1 * est2
pred_dg <- predict(cab_fit1, type = "response")
pred_dg$est_combined <- pred_dg$est1 * pred_dg$est2

# Tweedie: est is the combined expected value in response scale
pred_tw <- predict(cab_fit2, type = "response")

# Attach predicted values to observed data
obs_dg <- cabezon %>% mutate(predicted = pred_dg$est_combined, observed = larvae_100m3)
obs_tw <- cabezon %>% mutate(predicted = pred_tw$est,           observed = larvae_100m3)

# ── Grid predictions for spatial maps ─────────────────────────────────────────

# Delta-Gamma (cab_map already exists; add combined column)
cab_map_df <- cab_map$data %>%
  mutate(est_combined = est1 * est2)

plot_spatial_fit(cab_map_df, "est_combined", cabezon, "Delta-Gamma")

# Tweedie grid predictions
cab_map_tw <- predict(cab_fit2, newdata = pred_grid, type = "response")

# ── 1. Predicted vs. Observed scatter plots ───────────────────────────────────

scatter_dg <- ggplot(obs_dg, aes(x = predicted, y = observed)) +
  geom_point(alpha = 0.35, size = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_x_continuous(trans = "log1p", labels = scales::comma) +
  scale_y_continuous(trans = "log1p", labels = scales::comma) +
  labs(
    title = "Delta-Gamma: Predicted vs. Observed",
    x = expression("Predicted larvae per 100 m"^3),
    y = expression("Observed larvae per 100 m"^3)
  ) +
  theme_bw()

scatter_tw <- ggplot(obs_tw, aes(x = predicted, y = observed)) +
  geom_point(alpha = 0.35, size = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_x_continuous(trans = "log1p", labels = scales::comma) +
  scale_y_continuous(trans = "log1p", labels = scales::comma) +
  labs(
    title = "Tweedie: Predicted vs. Observed",
    x = expression("Predicted larvae per 100 m"^3),
    y = expression("Observed larvae per 100 m"^3)
  ) +
  theme_bw()

scatter_dg
scatter_tw

# ── 2. Faceted spatial maps: predicted surface + observed points ───────────────

plot_spatial_fit <- function(grid_preds, est_col, obs_data, model_name) {
  ggplot() +
    geom_raster(
      data = grid_preds,
      aes(x = longitude, y = latitude, fill = log1p(.data[[est_col]])),
      interpolate = TRUE
    ) +
    geom_sf(data = world, color = "gray30", fill = "gray85", linewidth = 0.3) +
    geom_point(
      data = obs_data,
      aes(x = longitude, y = latitude, size = log1p(larvae_100m3)),
      shape = 21, fill = "white", color = "black", stroke = 0.3, alpha = 0.75
    ) +
    scale_fill_viridis_c(
      name = expression(log(1 + hat(y))),
      option = "plasma",
      na.value = NA
    ) +
    scale_size_continuous(
      name = expression(log(1 + y)),
      range = c(0.3, 3.5)
    ) +
    coord_sf(xlim = x_limits, ylim = y_limits, expand = FALSE) +
    facet_wrap(~ year) +
    labs(title = paste(model_name, "— Predicted Surface & Observed Points")) +
    theme_bw(base_size = 9) +
    theme(
      strip.text     = element_text(size = 7),
      axis.text      = element_text(size = 6),
      legend.title   = element_text(size = 8),
      legend.key.height = unit(0.6, "cm")
    )
}

plot_spatial_fit(cab_map_df, "est_combined", cabezon, "Delta-Gamma")
plot_spatial_fit(cab_map_tw, "est",          cabezon, "Tweedie")

```


