---
title: "Cabezon sdmTMB — Constrained Analysis"
author: "Dan Portnof"
date: "2026-02-25"
output: word_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


---

# PART II — CONSTRAINED ANALYSIS (SPAWNING SEASON + SHELF REGION)

---

To refresh our memory, all analysis to this point has been done using CalCOFI data in the core CalCOFI grid, extended to the maximum distance from shore that these transect lines go (in excess of some 250 mi). Additionally, our data compasses all months and seasons over the observed time-series from 1981-2015.

This section will attempt to apply a finer grade to our analysis by constraining these parameters in the following ways:

1. We will focus on spawning season, defined in the 2019 stock assessment as **beginning in November and ending in March, with a peak in January and February.**

2. We will focus on the shelf rather than the full distance offshore, assuming that cabezon larvae occur most frequently and most regularly inshore.

It's worth keeping in mind, of course, that this constrains our spatial data further, *but,* our natural science might suggest that a narrower examination of spawning season larval abundance *in areas where we'd more likely expect larval occurrence* could be a better match to downstream adult distribution and abundance.

# 0. SETUP & STAR REPORT (CONSTRAINED)

```{r 0. SETUP & STAR REPORT (Part II)}
# ---- 0. SETUP & STAR REPORT (Part II)

# Load packages
library(tidyverse)
library(r4ss)
library(sdmTMB)
library(fmesher)
library(funtimes)

my.seed <- 666
set.seed(my.seed)

wd <- "Cab_SCS_BC_STAR" # Mac

# This reads the output files
pp <- SS_output(wd)

# Pull out derived quantities.  These are things like SSB and age-0 recruits.
derived_quants <- pp[["derived_quants"]]

# Pull out parameter estimates.  Recruitment deviations are estimated parameters.
params <- pp[["parameters"]]

# Plot recruitment deviations for the main estimation period.
devyrs <- 1970:2018
devdf <- pp$parameters[28:76, ]
devvalue <- devdf[,3]
devsd <- devdf[,11]
devlower <- (devvalue-1.96*devsd)
devmed   <- devvalue
devupper <- (devvalue+1.96*devsd)
STAR_recdevs <- data.frame(year=devyrs, value=devmed,lo=devlower,hi=devupper)

STAR_recdevs_plot <- ggplot(data=STAR_recdevs, aes(x=year, ymin=devlower, ymax=devupper)) +
  geom_ribbon(fill="skyblue", alpha=0.5) +
  geom_line(aes(y=devmed)) +
  labs(title="Cabezon SoCal Recruitment Deviations") +
  theme_bw()

STAR_recdevs_plot

# Plot age-0 recruits
age0yrs <- 1970:2018
startrowindex <- which(rownames(derived_quants)=="Recr_1970")
endrowindex <- which(rownames(derived_quants)=="Recr_2018")
age0 <- derived_quants[174:222, "Value"]
age0sd <- derived_quants[174:222, "StdDev"]
age0lower <- (age0-1.96*age0sd)
age0upper <- (age0+1.96*age0sd)

STAR_age0 <- data.frame(year=age0yrs, value=age0,lo=age0lower,hi=age0upper)

STAR_age0_plot <- ggplot(data=STAR_age0, aes(x=year, ymin=age0lower, ymax=age0upper)) +
  geom_ribbon(fill="skyblue", alpha=0.5) +
  geom_line(aes(y=age0)) +
  labs(title="Cabezon SoCal Age 0 Recruits") +
  theme_bw()

STAR_age0_plot

# Plot SSB
ssbyrs <- 1916:2018
startrowindex <- which(rownames(derived_quants)=="SSB_1916")
endrowindex <- which(rownames(derived_quants)=="SSB_2018")
ssb <- derived_quants[3:105, "Value"]
ssbsd <- derived_quants[3:105, "StdDev"]
ssblower <- (ssb-1.96*ssbsd)
ssbupper <- (ssb+1.96*ssbsd)

STAR_SSB <- data.frame(year=ssbyrs, value=ssb,lo=ssblower,hi=ssbupper)

STAR_SSB_plot <- ggplot(data=STAR_SSB, aes(x=year, ymin=ssblower, ymax=ssbupper)) +
  geom_ribbon(fill="skyblue", alpha=0.5) +
  geom_line(aes(y=ssb)) +
  labs(title="Cabezon SoCal SSB") +
  theme_bw()

STAR_SSB_plot

```

# 1. DATA QC + EXPLORATION (CONSTRAINED)

```{r 1. DATA QC (Part II), cache = T}
#---- DATA QC + EXPLORATION (Part II)

cabezon <- read.csv("cabezon_calcofi_data.csv")

str(cabezon)

# Add factor variable for season
cabezon <- cabezon %>%
  mutate(season = as.factor(season))

# ---- 1. DATA QC + EXPLORATION (Part II)

# Check for how many observations per ...

# year
cabezon %>%
  count(year, season, line.station) %>%
  filter(n > 1)
# month
cabezon %>%
  count(year, month, line.station) %>%
  filter(n > 1)
# cruise
cabezon %>%
  count(cruise, line.station) %>%
  filter(n > 1)

# Station coverage across years
cabezon %>%
  group_by(year) %>%
  summarize(n_stations = n_distinct(line.station))
table(cabezon$line.station, cabezon$year)
  # Pretty steady coverage, especially post-1989

# Check the spatial grid for a quick visual
ggplot(cabezon, aes(longitude, latitude)) +
  geom_point(alpha = 0.3)

# How much zero inflation do we have?
cabezon %>%
  mutate(zero = larvae_100m3 == 0) %>%
  count(zero)
  # Roughly ~85% zeros

# Raw distribution visual
plot(cabezon$volume_sampled, cabezon$larvae_100m3)

pos_vals <- cabezon$larvae_100m3[cabezon$larvae_100m3 > 0]
hist(log(pos_vals), breaks = 50, main = "log(positive larvae_100m3)")

cabezon %>%
  filter(larvae_100m3 > 0) %>%
  group_by(year) %>%
  summarize(m = mean(larvae_100m3), v = var(larvae_100m3), n = n()) %>%
  filter(n > 5) %>%
   ggplot(aes(log(m), log(v))) + geom_point() + geom_smooth(method = "lm") +
  geom_abline(slope = 2, intercept = 0, linetype = "dashed", color = "red") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Mean-variance relationship (dashed: red=gamma slope 2, blue=Poisson slope 1)")

# Also useful: what fraction are zeros?
mean(cabezon$larvae_100m3 == 0)

```

# 2. DATA PARING & EXPLORATION (CONSTRAINED)

```{r 2. DATA PARING AND EXPLORATION (CONSTRAINED), cache = T}

cabezon_shelf_spawn <- cabezon %>%
  filter(
    month %in% c(1, 2, 3),
    station <= 60.0
  )

cat("Rows retained:", nrow(cabezon_shelf_spawn), "of", nrow(cabezon), "\n")
cat("Years covered:", paste(range(cabezon_shelf_spawn$year), collapse = "–"), "\n")
cat("Months present:", paste(sort(unique(cabezon_shelf_spawn$month)), collapse = ", "), "\n")
cat("Station range:", paste(range(cabezon_shelf_spawn$station), collapse = "–"), "\n")
cat("Lines present:", paste(sort(unique(cabezon_shelf_spawn$line)), collapse = ", "), "\n")

cabezon_shelf_spawn %>%
  group_by(year) %>%
  summarize(
    n_tows = n(),
    n_pos = sum(larvae_100m3 > 0),
    pct_pos = round(100 * n_pos / n_tows, 1)
  ) %>%
  print(n = Inf)

# Composite station column (combines line + station)
cabezon_shelf_spawn <- cabezon_shelf_spawn %>%
  mutate(station_id = paste(line, station, sep = "_"))

# Do non-standard hauls have systematically different standard_haul_factor values?
summary(cabezon_shelf_spawn$standard_haul_factor)

# Are any values notably far from 1.0?
hist(cabezon_shelf_spawn$standard_haul_factor)
hist(log(cabezon_shelf_spawn$larvae_100m3))

# Check the spike at larvae_100m3 == 1
cabezon_shelf_spawn %>% count(larvae_100m3 == 1)

# Check how many true zeros are present
cabezon_shelf_spawn %>% count(larvae_100m3 == 0)

# Confirm no -Inf values making it into model inputs
sum(is.infinite(log(cabezon_shelf_spawn$larvae_100m3)))

cabezon_shelf_spawn %>%
  filter(standard_haul_factor != 0.02) %>%
  count(year) 

# Calculating positive tow values for visualization
ss_pos_tows_year <- cabezon_shelf_spawn %>%
  group_by(year) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )

ss_pos_tows_month <- cabezon_shelf_spawn %>%
  group_by(month) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )

ss_pos_tows_season <- cabezon_shelf_spawn %>%
  group_by(season) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )

ss_pos_tows_station <- cabezon_shelf_spawn %>%
  group_by(station_id) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )

# Annual mean larval observations with SE
annual_cab_ss_summary <- cabezon_shelf_spawn %>%
  group_by(year) %>%
  summarize(
    mean_larvae = mean(larvae_100m3, na.rm = T),
    sd_larvae = sd(larvae_100m3, na.rm = T),
    n = n(),
    se = sd_larvae / sqrt(n)
  )

# ---- 8.b. Annual average larval observations

ggplot(annual_cab_ss_summary, aes(x = year, y = mean_larvae)) +
  geom_ribbon(aes(ymin = mean_larvae - se, ymax = mean_larvae + se),
              alpha = 0.5, fill = "skyblue") +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue") +
  labs(
    x = "Year",
    y = expression("Mean larvae per 100 m"^3),
    title = "Annual Mean Larval Observations, Constrained to Spawning Season and Coastal Shelf",
    subtitle = expression("Ribbon shows " %+-% " 1 SE")
  ) +
  theme_bw()

# ---- 8.c. Stacked bar chart: positive vs. total tows by year, month, season

ss_plot_data <- bind_rows(
  ss_pos_tows_year   %>% mutate(group_var = "Year",   level = as.character(year)),
  ss_pos_tows_month  %>% mutate(group_var = "Month",  level = as.character(month)),
  ss_pos_tows_season %>% mutate(group_var = "Season", level = as.character(season))
) %>%
  mutate(negative_tows = total_tows - positive_tows) %>%
  pivot_longer(
    cols      = c(positive_tows, negative_tows),
    names_to  = "tow_type",
    values_to = "count"
  )

ss_plot_data <- ss_plot_data %>%
  mutate(level = case_when(
    group_var == "Month" ~ factor(level, levels = as.character(1:12)),
    group_var == "Season" ~ factor(level, levels = c("winter", "spring", "summer", "fall")),
    T ~ factor(level)
  ))

# Shared elements
shared_fill <- scale_fill_manual(
  values = c("negative_tows" = "grey80", "positive_tows" = "steelblue"),
  labels = c("negative_tows" = "Larvae Absent", "positive_tows" = "Larvae Present")
)

shared_bar_theme <- theme_minimal(base_size = 12) +
  theme(
    strip.text         = element_text(face = "bold", size = 13),
    axis.text.x        = element_text(angle = 45, hjust = 1),
    legend.position    = "bottom",
    panel.grid.major.x = element_blank(),
    plot.title         = element_text(face = "bold", size = 15),
    plot.subtitle      = element_text(color = "grey40", size = 10)
  )

# One plot for years; one plot for months + seasons
ss_plot_data_year <- ss_plot_data %>% filter(group_var == "Year")
ss_plot_data_ms   <- ss_plot_data %>% filter(group_var %in% c("Month", "Season"))

# Years plot
ggplot(ss_plot_data_year, aes(x = level, y = count, fill = tow_type)) +
  geom_bar(stat = "identity", width = 0.7) +
  shared_fill +
  shared_bar_theme +
  labs(
    title    = "Cabezon Larvae: Positive vs. Total Tows",
    subtitle = "Stacked bars show larvae-present (blue) against total sampling effort (gray)",
    x        = NULL,
    y        = "Number of Tows",
    fill     = NULL
  )

# Months + Seasons plot
ggplot(ss_plot_data_ms, aes(x = level, y = count, fill = tow_type)) +
  geom_bar(stat = "identity", width = 0.7) +
  facet_wrap(~ group_var, scales = "free_x", nrow = 1) +
  shared_fill + shared_bar_theme +
  labs(
    title = "Cabezon Larvae: Positive vs. Total Tows",
    subtitle = "Stacked bar charts show larve-present (blue) against total sampling effort (gray)",
    x        = NULL,
    y        = "Number of Tows",
    fill     = NULL
  )

#---- 8.d. Boxplots: log-larvae by year, month, season (positive tows only)

# Data binning
ss_box_data <- bind_rows(
  cabezon_shelf_spawn %>% mutate(group_var = "Year",   level = as.character(year)),
  cabezon_shelf_spawn %>% mutate(group_var = "Month",  level = as.character(month))
) %>%
  filter(larvae_100m3 > 0) %>%
  mutate(
    log_larvae = log(larvae_100m3),
    group_var  = factor(group_var, levels = c("Year", "Month", "Season")),
    level = case_when(
      group_var == "Month"  ~ factor(level, levels = as.character(1:12)),
      group_var == "Season" ~ factor(level, levels = c("winter", "spring", "summer", "fall")),
      TRUE                  ~ factor(level)
    )
  )

# Shared elements
shared_box_theme <- theme_minimal(base_size = 12) +
  theme(
    strip.text         = element_text(face = "bold", size = 13),
    axis.text.x        = element_text(angle = 45, hjust = 1),
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    plot.title         = element_text(face = "bold", size = 15),
    plot.subtitle      = element_text(color = "grey40", size = 10)
  )

# Years plot
ggplot(ss_box_data %>% filter(group_var == "Year"), 
       aes(x = level, y = log_larvae)) +
  geom_boxplot(fill = "steelblue", color = "grey30", alpha = 0.7, outlier.size = 1) +
  shared_box_theme +
  labs(
    title    = "Cabezon Larvae: Average Count by Year (Positive Tows Only)",
    subtitle = "Log-transformed larvae per 100m³; zero-catch tows excluded",
    x        = NULL,
    y        = "log(Larvae per 100m³)"
  )

# Months + Seasons plot
ggplot(ss_box_data %>% filter(group_var %in% c("Month", "Season")),
       aes(x = level, y = log_larvae)) +
  geom_boxplot(fill = "steelblue", color = "grey30", alpha = 0.7, outlier.size = 1) +
  facet_wrap(~ group_var, scales = "free_x", nrow = 1) +
  shared_box_theme +
  labs(
    title    = "Cabezon Larvae: Average Count by Month and Season (Positive Tows Only)",
    subtitle = "Log-transformed larvae per 100m³; zero-catch tows excluded",
    x = NULL, y = "log(Larvae per 100m³)"
  )


#---- 8.e. Heatmaps: mean log-larvae by station x year, month, season

heatmap_theme <- theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 7),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey40", size = 10)
  )

heatmap_fill <- scale_fill_viridis_c(option = "mako", name = "log(Larvae\nper 100m³)", na.value = "white")

# Year plot
cabezon_shelf_spawn %>%
  filter(larvae_100m3 > 0) %>%
  group_by(line.station, year) %>%
  summarise(mean_log_larvae = mean(log(larvae_100m3)), .groups = "drop") %>%
  ggplot(aes(x = year, y = line.station, fill = mean_log_larvae)) +
  geom_tile() + heatmap_fill + scale_y_discrete(limits = rev) + heatmap_theme +
  labs(title    = "Cabezon Larval Observations by Station and Year",
       subtitle = "Positive tows only; white = no positive tows recorded",
       x = NULL, y = "South \u2190 Station \u2192 North")

# Month plot
cabezon_shelf_spawn %>%
  filter(larvae_100m3 > 0) %>%
  group_by(line.station, month) %>%
  summarise(mean_log_larvae = mean(log(larvae_100m3)), .groups = "drop") %>%
  ggplot(aes(x = month, y = line.station, fill = mean_log_larvae)) +
  geom_tile() + heatmap_fill +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_discrete(limits = rev) + heatmap_theme +
  labs(title    = "Cabezon Larval Observations by Station and Month",
       subtitle = "Positive tows only; white = no positive tows recorded",
       x = NULL, y = "South \u2190 Station \u2192 North")

# Season plot
cabezon_shelf_spawn %>%
  filter(larvae_100m3 > 0) %>%
  group_by(line.station, season) %>%
  summarise(mean_log_larvae = mean(log(larvae_100m3)), .groups = "drop") %>%
  ggplot(aes(x = season, y = line.station, fill = mean_log_larvae)) +
  geom_tile() + heatmap_fill + scale_y_discrete(limits = rev) + heatmap_theme +
  labs(title    = "Cabezon Larval Observations by Station and Season",
       subtitle = "Positive tows only; white = no positive tows recorded",
       x = NULL, y = "South \u2190 Station \u2192 North")

```

# 3. sdmTMB MODEL FITTING (CONSTRAINED)

```{r 3. sdmTMB MODEL FITTING (CONSTRAINED), cache = T}

set.seed(my.seed)

#---- 3.a. Build mesh
if (!all(c("X", "Y") %in% names(cabezon_shelf_spawn))) {
  cabezon_shelf_spawn <- sdmTMB::add_utm_columns(cabezon_shelf_spawn, ll_names = c("longitude", "latitude"))
}

ss_mesh <- fm_mesh_2d(
  loc = cabezon_shelf_spawn[,c("X","Y")],
  cutoff = 20,
  max.edge = c(75, 150),
  offset = c(45, 120)
)
ss_cab_mesh <- make_mesh(data = cabezon_shelf_spawn, c("X", "Y"), mesh = ss_mesh)
plot(ss_cab_mesh); title("SPDE Mesh for Cabezon CalCOFI analysis (Constrained)")

#---- 3.b. Fit candidate models

# Add year factor variable
cabezon_shelf_spawn$fyear <- as.factor(cabezon_shelf_spawn$year)

# NULL Delta-Gamma model

ss_cab_null <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = "off",
  spatiotemporal = "off",
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_null)

# NULL Delta-Lognormal model

ss_cab_null_dln <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = "off",
  spatiotemporal = "off",
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_null_dln)

# NULL Tweedie model

ss_cab_null_tw <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = tweedie(),
  spatial = "off",
  spatiotemporal = "off",
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_null_tw)

null_AIC_table <- data.frame(
  fit_name = c(
    "ss_cab_null",
    "ss_cab_null_dln",
    "ss_cab_null_tw"
  ),
  model = c(
    "Delta-Gamma NULL (Constrained)",
    "Delta-Lognormal NULL (Constrained)",
    "Tweedie NULL (Constrained)"
  ),
  AIC = c(
    AIC(ss_cab_null),
    AIC(ss_cab_null_dln),
    AIC(ss_cab_null_tw)
  ),
    delta_AIC = NA
  )

null_AIC_table$delta_AIC <- null_AIC_table$AIC - min(null_AIC_table$AIC)

null_AIC_table <- null_AIC_table[order(null_AIC_table$AIC, decreasing = F), ]

null_AIC_table

# Delta-Lognormal: Global Intercept, spatial on, spatiotemporal delta AR1, spatiotemporal lognormal AR1
ss_cab_fit_dln1 <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("ar1", "ar1"),
  offset = NULL,
  extra_time = 1982,
  priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 50,
                                             sigma_lt = 5),
                        matern_st = pc_matern(range_gt = 50,
                                              sigma_lt = 5))
)

sanity(ss_cab_fit_dln1)

AIC(ss_cab_fit_dln1)

# Delta-Lognormal: Global Intercept, spatial on, delta AR1, spatiotemporal delta AR1, spatiotemporal lognormal iid
ss_cab_fit_dln2 <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("ar1", "iid"),
  offset = NULL,
  extra_time = 1982,
  priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 50,
                                             sigma_lt = 5),
                        matern_st = pc_matern(range_gt = 50,
                                              sigma_lt = 5))
)

sanity(ss_cab_fit_dln2)

AIC(ss_cab_fit_dln2)

# Delta-Lognormal: Global Intercept, spatial on, delta AR1, spatiotemporal delta AR1, spatiotemporal lognormal off
ss_cab_fit_dln3 <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("ar1", "off"),
  offset = NULL,
  extra_time = 1982,
  priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 50,
                                             sigma_lt = 5))
)

sanity(ss_cab_fit_dln3)

AIC(ss_cab_fit_dln3)

# Delta-Lognormal: fixed year effect, spatial on, spatiotemopral off
ss_cab_fit_dln4 <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("off", "off"),
  offset = NULL,
  extra_time = 1982,
  priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 50,
                                             sigma_lt = 5))
)

sanity(ss_cab_fit_dln4)

AIC(ss_cab_fit_dln4)

# Delta-Lognormal: fixed year effect, spatial on, spatiotemporal iid
ss_cab_fit_dln5 <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("iid", "iid"),
  offset = NULL,
  extra_time = 1982,
  priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 50,
                                             sigma_lt = 5),
                        matern_st = pc_matern(range_gt = 50,
                                              sigma_lt = 5))
)

sanity(ss_cab_fit_dln5)

AIC(ss_cab_fit_dln5)

# Delta-Lognormal: fixed year effect, spatial on, iid delta spatiotemporal, gamma spatiotemporal off
ss_cab_fit_dln6 <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("iid", "off"),
  offset = NULL,
  extra_time = 1982,
  priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 50,
                                             sigma_lt = 5))
)

sanity(ss_cab_fit_dln6)

AIC(ss_cab_fit_dln6)



#---- 3.c. Random effect parameter estimates for best models

tidy(ss_cab_fit_dln1, effects = "ran_pars")
tidy(ss_cab_fit_dln2, effects = "ran_pars")
tidy(ss_cab_fit_dln3, effects = "ran_pars")
tidy(ss_cab_fit_dln4, effects = "ran_pars")
tidy(ss_cab_fit_dln5, effects = "ran_pars")

#---- 3.d. Construct model comparison table

AIC_table <- data.frame(
  fit_name = c(
    "ss_cab_fit_dln1",
    "ss_cab_fit_dln2",
    "ss_cab_fit_dln3",
    "ss_cab_fit_dln4",
    "ss_cab_fit_dln5"
  ),
  model = c(
    "Delta-Lognormal Global Intercept + spatial + spatiotemporal delta AR1 + spatiotemporal lognormal iid (Spawning Season)",
    "Delta-Lognormal Global Intercept + spatial + spatiotemporal delta AR1 + spatiotemporal gamma off (Spawning Season)",
    "Delta-Lognormal fixed year effect + spatial + spatiotemporal off (Spawning Season)",
    "Delta-Lognormal fixed year effect + spatial + spatiotemporal iid (Spawning Season)",
    "Delta-Lognormal fixed year effect + spatial + spatiotemporal delta iid + spatiotemporal gamma off (Spawning Season)"
  ),
  AIC = c(
    AIC(ss_cab_fit_dln1),
    AIC(ss_cab_fit_dln2),
    AIC(ss_cab_fit_dln3),
    AIC(ss_cab_fit_dln4),
    AIC(ss_cab_fit_dln5)
  ),
    delta_AIC = NA
  )

AIC_table$delta_AIC <- AIC_table$AIC - min(AIC_table$AIC)

AIC_table <- AIC_table[order(AIC_table$AIC, decreasing = F), ]

AIC_table

ss_cab_fit_dg1 <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("ar1", "off"),
  offset = NULL,
  extra_time = 1982,
  priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 50,
                                             sigma_lt = 5))
)

sanity(ss_cab_fit_dg1)
AIC(ss_cab_fit_dg1)


```


# 4. MODEL DIAGNOSTICS & ABUNDANCE INDICES (CONSTRAINED)

## 4.1. ss_cab_fit1 — Full Delta-Gamma Model

```{r 4.1. ss_cab_fit1 diagnostics, cache = T}

#---- 4.1.a. Calibration summaries

ss_full_pred_obs <- predict(ss_cab_fit1, type = "response")

# Overall mean calibration
ss_full_pred_obs %>%
  summarise(
    obs_mean = mean(larvae_100m3, na.rm = TRUE),
    pred_mean = mean(est, na.rm = TRUE),
    obs_occ = mean(larvae_100m3 > 0, na.rm = TRUE),
    pred_occ = mean(est1, na.rm = TRUE),
    obs_pos_mean = mean(larvae_100m3[larvae_100m3 > 0], na.rm = TRUE),
    pred_pos_mean = mean(est2, na.rm = TRUE)
  )

# Binomial calibration; i.e. predicted presence probability vs. observed presence rate
ss_full_pred_obs$presence <- ss_full_pred_obs$larvae_100m3 > 0

ss_full_calibration_summary <- ss_full_pred_obs %>%
  mutate(prob_bin = cut(est1, breaks = seq(0, 1, by = 0.1))) %>%
  group_by(prob_bin) %>%
  summarise(
    mean_pred = mean(est1),
    observed = mean(presence),
    n = n()
  )

plot(ss_full_calibration_summary$mean_pred, ss_full_calibration_summary$observed,
     xlab = "Predicted presence probability",
     ylab = "Observed presence frequency")
abline(0,1,lty=2)

#---- 4.1.b. Predicted vs. observed plots

# Raw scale comparison
plot(
  x = ss_full_pred_obs$est,
  y = cabezon_shelf_spawn$larvae_100m3,
  xlab = "Predicted",
  ylab = "Observed"
)
abline(0, 1, lty = 2)

# Log scale comparison
plot(log1p(ss_full_pred_obs$est), log1p(cabezon_shelf_spawn$larvae_100m3),
     xlab = "log(Predicted + 1)", ylab = "log(Observed + 1)")
abline(0, 1, lty = 2)

# Gamma (positive density) component only
pos_idx <- cabezon_shelf_spawn$larvae_100m3 > 0
plot(log(ss_full_pred_obs$est2[pos_idx]), log(cabezon_shelf_spawn$larvae_100m3[pos_idx]),
     xlab = "log(Predicted positive abundance)",
     ylab = "log(Observed positive abundance)")
abline(0, 1, lty = 2)

#---- 4.1.c. Randomized quantile residuals

ss_full_rq_res <- residuals(ss_cab_fit1, type = "mle-mvn")

# QQ plot
qqnorm(ss_full_rq_res)
qqline(ss_full_rq_res)

# Spatial residual map by year (kind of hard to see)
ss_full_pred_obs$rq_resid <- ss_full_rq_res

ggplot(ss_full_pred_obs, aes(x = longitude, y = latitude, color = rq_resid)) +
  geom_point() +
  scale_color_gradient2() +
  facet_wrap(~ year)

#---- 4.1.d. Temporal calibration

# Are predictions tracking temporal trends?
ss_full_pred_obs %>%
  group_by(year) %>%
  summarize(
    obs = mean(larvae_100m3),
    pred = mean(est)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = obs), color = "black") +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed")

#---- 4.1.e. Abundance index construction
# Note! The index uncertainty here was far too wide using a regular spatial grid, probably because interpolation over years with no CalCOFI coverage was causing inflation. I kept that plot in for reference, but 10.1.f. (observed station grid) is probably preferred for diagnosis.

ss_full_pred_grid <- expand.grid(
  X = seq(min(cabezon_shelf_spawn$X), max(cabezon_shelf_spawn$X), by = 5),
  Y = seq(min(cabezon_shelf_spawn$Y), max(cabezon_shelf_spawn$Y), by = 5)
) %>%
  tidyr::crossing(year = as.integer(sort(unique(cabezon_shelf_spawn$year)))) %>%
  mutate(fyear = as.factor(year))

ss_full_cab_map <- predict(ss_cab_fit1, newdata = ss_full_pred_grid, return_tmb_object = T)
ss_full_cab_index <- get_index(ss_full_cab_map, area = 1, bias_correct = T)

ggplot(ss_full_cab_index, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year", title = "Abundance Index: Delta-Gamma on Global Intercept with AR1 Spatiotemporal Fields")

#---- 4.1.f. Abundance index: constrained to observed station grid (probably preferred)

ss_full_pred_grid_obs <- cabezon_shelf_spawn %>%
  dplyr::select(X, Y, year) %>%
  distinct()

ss_full_cab_map_obs <- predict(ss_cab_fit1,
                       newdata = ss_full_pred_grid_obs,
                       return_tmb_object = T)

ss_full_cab_index_obs <- get_index(ss_full_cab_map_obs,
                           area = 1,
                           bias_correct = T)

ggplot(ss_full_cab_index_obs, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year", title = "Abundance Index: Observed Station Grid")
```

## 4.2. ss_cab_fit_dln1 — AR1 Delta-Lognormal Model

```{r 4.2. ss_cab_fit_dln1 diagnostics, cache = T}

#---- 4.2.a. Calibration summaries

ss_dln_fy_pred_obs <- predict(ss_cab_fit_dln1, type = "response")

# Overall mean calibration
ss_dln_fy_pred_obs %>%
  summarise(
    obs_mean     = mean(larvae_100m3, na.rm = TRUE),
    pred_mean    = mean(est, na.rm = TRUE),
    obs_occ      = mean(larvae_100m3 > 0, na.rm = TRUE),
    pred_occ     = mean(est1, na.rm = TRUE),
    obs_pos_mean = mean(larvae_100m3[larvae_100m3 > 0], na.rm = TRUE),
    pred_pos_mean = mean(est2, na.rm = TRUE)
  )

# Binomial calibration
ss_dln_fy_pred_obs$presence <- ss_dln_fy_pred_obs$larvae_100m3 > 0

ss_dln_fy_calibration_summary <- ss_dln_fy_pred_obs %>%
  mutate(prob_bin = cut(est1, breaks = seq(0, 1, by = 0.1))) %>%
  group_by(prob_bin) %>%
  summarise(
    mean_pred = mean(est1),
    observed  = mean(presence),
    n         = n()
  )

plot(ss_dln_fy_calibration_summary$mean_pred, ss_dln_fy_calibration_summary$observed,
     xlab = "Predicted presence probability",
     ylab = "Observed presence frequency")
abline(0, 1, lty = 2)

#---- 4.2.b. Predicted vs. observed plots

# Raw scale
plot(
  x = ss_dln_fy_pred_obs$est,
  y = cabezon_shelf_spawn$larvae_100m3,
  xlab = "Predicted",
  ylab = "Observed"
)
abline(0, 1, lty = 2)

# Log scale
plot(log1p(ss_dln_fy_pred_obs$est), log1p(cabezon_shelf_spawn$larvae_100m3),
     xlab = "log(Predicted + 1)", ylab = "log(Observed + 1)")
abline(0, 1, lty = 2)

# Gamma (positive density) component only
pos_idx <- cabezon_shelf_spawn$larvae_100m3 > 0
plot(log(ss_dln_fy_pred_obs$est2[pos_idx]), log(cabezon_shelf_spawn$larvae_100m3[pos_idx]),
     xlab = "log(Predicted positive abundance)",
     ylab = "log(Observed positive abundance)")
abline(0, 1, lty = 2)

#---- 4.2.c. Randomized quantile residuals

ss_dln_fy_rq_res <- residuals(ss_cab_fit_dln1, type = "mle-mvn")

# QQ plot
qqnorm(ss_dln_fy_rq_res)
qqline(ss_dln_fy_rq_res)

# Spatial residual map by year
ss_dln_fy_pred_obs$rq_resid <- ss_dln_fy_rq_res

ggplot(ss_dln_fy_pred_obs, aes(x = longitude, y = latitude, color = rq_resid)) +
  geom_point() +
  scale_color_gradient2() +
  facet_wrap(~ year)

#---- 4.2.d. Temporal calibration

ss_dln_fy_pred_obs %>%
  group_by(year) %>%
  summarize(
    obs  = mean(larvae_100m3),
    pred = mean(est)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = obs),  color = "black") +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed")

#---- 4.2.e. Abundance index: regular spatial grid

ss_dln_fy_pred_grid <- expand.grid(
  X = seq(min(cabezon_shelf_spawn$X), max(cabezon_shelf_spawn$X), by = 5),
  Y = seq(min(cabezon_shelf_spawn$Y), max(cabezon_shelf_spawn$Y), by = 5)
) %>%
  tidyr::crossing(year = as.integer(sort(unique(cabezon_shelf_spawn$year)))) %>%
  mutate(fyear = as.factor(year))

ss_dln_fy_cab_map   <- predict(ss_cab_fit_dln1, newdata = ss_dln_fy_pred_grid, return_tmb_object = TRUE)
ss_dln_fy_cab_index <- get_index(ss_dln_fy_cab_map, area = 1, bias_correct = TRUE)

ggplot(ss_dln_fy_cab_index, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Delta-Lognormal on Global Intercept with AR1 Spatiotemporal Fields")

#---- 4.2.f. Abundance index: observed station grid (preferred)

ss_dln_fy_pred_grid_obs <- cabezon_shelf_spawn %>%
  dplyr::select(X, Y, year, fyear) %>%
  distinct()

ss_dln_fy_cab_map_obs   <- predict(ss_cab_fit01,
                               newdata = ss_dln_fy_pred_grid_obs,
                               return_tmb_object = TRUE)

ss_dln_fy_cab_index_obs <- get_index(ss_dln_fy_cab_map_obs,
                                 area = 1,
                                 bias_correct = TRUE)

ggplot(ss_dln_fy_cab_index_obs, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Observed Station Grid")

```

# 5. STAR COMPARISONS & CCFs (CONSTRAINED)

## 5.1. ss_cab_fit1 — STAR Comparisons & CCFs

```{r 5.1. ss_cab_fit1 STAR comparisons full extrapolation, cache = T}
#---- 5.1. ss_cab_fit1 STAR comparisons and CCF

# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 10.1.f. abundance index, assuming it's more stable.
ss_full_combined <- ss_full_cab_index %>%
  dplyr::select(year, est, lwr, upr) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
ss_full_combined_obs <- ss_full_combined %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf_boot(
  x = ss_full_combined_obs$larvae_index,
  y = ss_full_combined_obs$age0,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)

# CCF: Restricted larval index vs. recruitment deviations
ccf_boot(
  x = ss_full_combined_obs$larvae_index,
  y = ss_full_combined_obs$rec_dev,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Recruitment Deviations")


# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf_boot(
  x = ss_full_combined_obs$larvae_index,
  y = ss_full_combined_obs$ssb,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)

# First-differenced
ccf_boot(
  x = diff(ss_full_combined_obs$larvae_index),
  y = diff(ss_full_combined_obs$ssb),
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)

# And only differenced SSB on raw larvae
ccf_boot(
  x = ss_full_combined_obs$larvae_index,
  y = diff(ss_full_combined_obs$ssb),
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)

# Join differenced SSB back to index
ssb_diff_df <- ss_full_combined_obs %>%
  arrange(year) %>%
  mutate(ssb_diff = c(NA, diff(ssb))) %>%
  filter(!is.na(ssb_diff))

lm_lag0 <- lm(ssb_diff ~ larvae_index, data = ssb_diff_df)
summary(lm_lag0)

# Visual
ssb_diff_df %>%
  ggplot(aes(x = larvae_index, y = ssb_diff, label = year)) +
  geom_point() +
  geom_text(nudge_y = 50, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index", y = "ΔSSB (mt)",
       title = "Contemporaneous larval index vs. change in SSB") +
  theme_minimal()

# Assuming ecologically that SSB would have the strongest relationship to larval abundance, we'll plot those two time series visually for a qualitative look
scale_factor_ssb <- max(ss_full_combined_obs$larvae_index, na.rm = TRUE) /
  max(ss_full_combined_obs$ssb, na.rm = TRUE)

ggplot(ss_full_combined_obs %>% filter(year >= 1984, year <= 2015),
       aes(x = year)) +
  # SSB scaled UP to primary axis units
  geom_ribbon(aes(ymin = ssb_lo * scale_factor_ssb,
                  ymax = ssb_hi * scale_factor_ssb),
              fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = ssb * scale_factor_ssb), color = "steelblue", linewidth = 0.8) +
  # Larvae index on primary axis
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "coral", alpha = 0.3) +
  geom_line(aes(y = larvae_index), color = "coral", linewidth = 0.8) +
  scale_y_continuous(
    name = "Estimated Larval Abundance Index; Delta-Gamma full AR1",
    sec.axis = sec_axis(~ . / scale_factor_ssb,
                        name = "SSB (mt)")
  ) +
  scale_x_continuous(breaks = seq(1984, 2015, by = 4)) +
  labs(title = "SoCal Cabezon: Larval Index vs. SSB (1984–2015)", x = "Year") +
  theme_bw()

```

```{r 5.1.2. ss_cab_fit1 STAR comparisons station extrapolation, cache = T}
#---- 10.2. ss_cab_fit1 STAR comparisons and CCF

# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 10.1.f. abundance index, assuming it's more stable.
ss_full_combined_sta <- ss_full_cab_index_obs %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
ss_full_combined_sta <- ss_full_combined_sta %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf_boot(
  x = ss_full_combined_sta$larvae_index,
  y = ss_full_combined_sta$age0,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)

# CCF: Restricted larval index vs. recruitment deviations
ccf_boot(
  x = ss_full_combined_sta$larvae_index,
  y = ss_full_combined_sta$rec_dev,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Recruitment Deviations")


# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf_boot(
  x = ss_full_combined_sta$larvae_index,
  y = ss_full_combined_sta$ssb,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)

# First-differenced
ccf_boot(
  x = diff(ss_full_combined_sta$larvae_index),
  y = diff(ss_full_combined_sta$ssb),
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)
```

## 5.2. ss_cab_fit01 — STAR Comparisons & CCFs

```{r 5.2. ss_cab_fit01 STAR comparisons full extrapolation, cache = T}

#---- 5.2. ss_cab_fit01 STAR comparisons and CCF

# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 10.4.f. abundance index, assuming it's more stable.
ss_dln_fy_combined <- ss_dln_fy_cab_index %>%
  dplyr::select(year, est, lwr, upr) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
ss_dln_fy_combined_obs <- ss_dln_fy_combined %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf_boot(
  x = ss_dln_fy_combined_obs$larvae_index,
  y = ss_dln_fy_combined_obs$age0,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)

# CCF: Restricted larval index vs. recruitment deviations
ccf_boot(
  x = ss_dln_fy_combined_obs$larvae_index,
  y = ss_dln_fy_combined_obs$rec_dev,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Recruitment Deviations")


# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf_boot(
  x = ss_dln_fy_combined_obs$larvae_index,
  y = ss_dln_fy_combined_obs$ssb,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)

# First-differenced
ccf_boot(
  x = diff(ss_dln_fy_combined_obs$larvae_index),
  y = diff(ss_dln_fy_combined_obs$ssb),
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)

# Assuming ecologically that SSB would have the strongest relationship to larval abundance, we'll plot those two time series visually for a qualitative look
dln_scale_factor_ssb <- max(ss_dln_fy_combined$larvae_index, na.rm = TRUE) /
  max(ss_dln_fy_combined$ssb, na.rm = TRUE)

ggplot(ss_dln_fy_combined %>% filter(year >= 1984, year <= 2015),
       aes(x = year)) +
  # SSB scaled UP to primary axis units
  geom_ribbon(aes(ymin = ssb_lo * dln_scale_factor_ssb,
                  ymax = ssb_hi * dln_scale_factor_ssb),
              fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = ssb * dln_scale_factor_ssb), color = "steelblue", linewidth = 0.8) +
  # Larvae index on primary axis
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "coral", alpha = 0.3) +
  geom_line(aes(y = larvae_index), color = "coral", linewidth = 0.8) +
  scale_y_continuous(
    name = "Estimated Larval Abundance Index; Delta-Lognormal Fixed Year",
    sec.axis = sec_axis(~ . / scale_factor_ssb,
                        name = "SSB (mt)")
  ) +
  scale_x_continuous(breaks = seq(1984, 2015, by = 4)) +
  labs(title = "SoCal Cabezon: Larval Index vs. SSB (1984–2015)", x = "Year") +
  theme_bw()
```

```{r 5.2.2. ss_cab_fit01 STAR comparisons station extrapolation, cache = T}

#---- 10.5. ss_cab_fit01 STAR comparisons and CCF

# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 10.4.f. abundance index, assuming it's more stable.
ss_fy_combined_sta <- ss_dln_fy_cab_index_obs %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
ss_fy_combined_sta <- ss_fy_combined_sta %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf_boot(
  x = ss_fy_combined_sta$larvae_index,
  y = ss_fy_combined_sta$age0,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)

# CCF: Restricted larval index vs. recruitment deviations
ccf_boot(
  x = ss_fy_combined_sta$larvae_index,
  y = ss_fy_combined_sta$rec_dev,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Recruitment Deviations")


# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf_boot(
  x = ss_fy_combined_sta$larvae_index,
  y = ss_fy_combined_sta$ssb,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)

# First-differenced
ccf_boot(
  x = diff(ss_fy_combined_sta$larvae_index),
  y = diff(ss_fy_combined_sta$ssb),
  lag.max = 10,
  plot = "Spearman",
  B = 1000
)

```

# 6. REGRESSION ANALYSES (CONSTRAINED)

## 6.1. ss_cab_fit1 — Regression Analysis

```{r 6.1. ss_cab_fit1 correlation testing, cache = T}
# RQ residuals for full AR1 spatiotemporal model
ss_full_pred_obs %>%
  group_by(year) %>%
  mutate(med_resid = median(rq_resid)) %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method  = "loess", span = 0.4,
    se      = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Year",
    y     = "RQ Residual",
    title = "Annual RQ Residual Distributions — cab_fit1"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

ss_full_pred_obs %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x     = "Year",
    y     = "Cumulative Sum of Median RQ Residual",
    title = "CUSUM — cab_fit1"
  ) +
  theme_minimal()

# Is the underprediction concentrated in one delta component?
# Separate RQ residuals by model component
ss_full_pred_obs <- ss_full_pred_obs %>%
  mutate(
    rq_resid1 = residuals(ss_cab_fit1, model = 1, type = "mle-mvn"),
    rq_resid2 = residuals(ss_cab_fit1, model = 2, type = "mle-mvn")
  )

# CUSUM for each component separately
ss_full_pred_obs %>%
  group_by(year) %>%
  summarise(
    med_resid1 = median(rq_resid1, na.rm = TRUE),
    med_resid2 = median(rq_resid2, na.rm = TRUE),  # positive tows only
    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    cusum1 = cumsum(replace_na(med_resid1, 0)),
    cusum2 = cumsum(replace_na(med_resid2, 0))
  ) %>%
  pivot_longer(cols = c(cusum1, cusum2),
               names_to = "component",
               values_to = "cusum") %>%
  ggplot(aes(x = year, y = cusum, color = component)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("cusum1" = "steelblue", "cusum2" = "firebrick"),
                     labels = c("Delta (binomial)", "Gamma")) +
  labs(x = "Year", y = "CUSUM", title = "CUSUM by Delta Model Component — ss_cab_fit1") +
  theme_minimal()

# Is the underprediction spatially structured?
ss_full_pred_obs %>%
  group_by(station) %>%
  summarise(mean_resid = mean(rq_resid), .groups = "drop") %>%
  ggplot(aes(x = station, y = mean_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = TRUE) +
  labs(x = "Station", y = "Mean RQ Residual",
       title = "Mean Residual by Station — ss_cab_fit1") +
  theme_minimal()


#---- Simple lagged regression at -1 to check

ss_full_combined_lagged <- ss_full_combined_obs %>%
  arrange(year) %>%
  mutate(
    age0_lead1 = lead(age0, 1),
    rec_dev_lead1 = lead(rec_dev, 1)
  )

ss_full_lm_age0 <- lm(age0_lead1 ~ larvae_index, data = ss_full_combined_lagged)
ss_full_lm_recdev <- lm(rec_dev_lead1 ~ larvae_index, data = ss_full_combined_lagged)

summary(ss_full_lm_age0)
summary(ss_full_lm_recdev)

ss_full_combined_lagged %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = T) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

ss_full_combined_lagged_no83 <- ss_full_combined_lagged %>% filter(year != 1983)

ss_full_lm_age0_no83 <- lm(age0_lead1 ~ larvae_index, data = ss_full_combined_lagged_no83)
summary(ss_full_lm_age0_no83)

ss_full_lm_age0_detrend   <- lm(age0_lead1    ~ larvae_index + year, data = ss_full_combined_lagged)
ss_full_lm_recdev_detrend <- lm(rec_dev_lead1 ~ larvae_index + year, data = ss_full_combined_lagged)

summary(ss_full_lm_age0_detrend)
summary(ss_full_lm_recdev_detrend)

plot(ss_full_lm_age0, which = 4)  # Cook's distance plot

# Remove the three most anomalous years
ss_full_combined_lagged_trimmed <- ss_full_combined_lagged %>% 
  filter(!year %in% c(1983, 1986, 1998))

ss_full_combined_lagged_trimmed %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = T) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

ss_full_lm_age0_trimmed <- lm(age0_lead1 ~ larvae_index, data = ss_full_combined_lagged_trimmed)
summary(ss_full_lm_age0_trimmed)

ss_full_lm_age0_trimmed_detrend <- lm(age0_lead1 ~ larvae_index + year, data = ss_full_combined_lagged_trimmed)
summary(ss_full_lm_age0_trimmed_detrend)

```

## 6.2. ss_cab_fit01 — Regression Analysis

```{r 6.2. ss_cab_fit01 correlation testing, cache = T}
# RQ residuals for fixed year effect + IID model
ss_dln_fy_pred_obs %>%
  group_by(year) %>%
  mutate(med_resid = median(rq_resid)) %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method  = "loess", span = 0.4,
    se      = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Year",
    y     = "RQ Residual",
    title = "Annual RQ Residual Distributions — ss_cab_fit_dln1"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

ss_dln_fy_pred_obs %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x     = "Year",
    y     = "Cumulative Sum of Median RQ Residual",
    title = "CUSUM — ss_cab_fit_dln1"
  ) +
  theme_minimal()

# 1. Is the underprediction concentrated in one delta component?
ss_dln_fy_pred_obs <- ss_dln_fy_pred_obs %>%
  mutate(
    rq_resid1 = residuals(ss_cab_fit_dln1, model = 1, type = "mle-mvn"),
    rq_resid2 = residuals(ss_cab_fit_dln1, model = 2, type = "mle-mvn")
  )

ss_dln_fy_pred_obs %>%
  group_by(year) %>%
  summarise(
    med_resid1 = median(rq_resid1, na.rm = TRUE),
    med_resid2 = median(rq_resid2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    cusum1 = cumsum(replace_na(med_resid1, 0)),
    cusum2 = cumsum(replace_na(med_resid2, 0))
  ) %>%
  pivot_longer(cols = c(cusum1, cusum2),
               names_to  = "component",
               values_to = "cusum") %>%
  ggplot(aes(x = year, y = cusum, color = component)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("cusum1" = "steelblue", "cusum2" = "firebrick"),
                     labels = c("cusum1" = "Delta (binomial)", "cusum2" = "Gamma")) +
  labs(x = "Year", y = "CUSUM",
       title = "CUSUM by Delta Model Component — ss_cab_fit_dln1") +
  theme_minimal()

# 2. Is the underprediction spatially structured?
ss_dln_fy_pred_obs %>%
  group_by(station) %>%
  summarise(mean_resid = mean(rq_resid), .groups = "drop") %>%
  ggplot(aes(x = station, y = mean_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = TRUE) +
  labs(x = "Station", y = "Mean RQ Residual",
       title = "Mean Residual by Station — ss_cab_fit_dln1") +
  theme_minimal()

#---- Simple lagged regression at lag -1

ss_dln_fy_combined_lagged <- ss_dln_fy_combined_obs %>%
  arrange(year) %>%
  mutate(
    age0_lead1    = lead(age0, 1),
    rec_dev_lead1 = lead(rec_dev, 1)
  )

ss_dln_fy_lm_age0   <- lm(age0_lead1    ~ larvae_index, data = ss_dln_fy_combined_lagged)
ss_dln_fy_lm_recdev <- lm(rec_dev_lead1 ~ larvae_index, data = ss_dln_fy_combined_lagged)

summary(ss_dln_fy_lm_age0)
summary(ss_dln_fy_lm_recdev)

ss_dln_fy_combined_lagged %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

# Sensitivity: remove 1983
ss_dln_fy_combined_lagged_no83 <- ss_dln_fy_combined_lagged %>% filter(year != 1983)
ss_dln_fy_lm_age0_no83 <- lm(age0_lead1 ~ larvae_index, data = ss_dln_fy_combined_lagged_no83)
summary(ss_dln_fy_lm_age0_no83)

# Detrended regressions
ss_dln_fy_lm_age0_detrend   <- lm(age0_lead1    ~ larvae_index + year, data = ss_dln_fy_combined_lagged)
ss_dln_fy_lm_recdev_detrend <- lm(rec_dev_lead1 ~ larvae_index + year, data = ss_dln_fy_combined_lagged)
summary(ss_dln_fy_lm_age0_detrend)
summary(ss_dln_fy_lm_recdev_detrend)

# Cook's distance
plot(ss_dln_fy_lm_age0, which = 4)

# Trimmed analysis: remove three most anomalous years
ss_dln_fy_combined_lagged_trimmed <- ss_dln_fy_combined_lagged %>%
  filter(!year %in% c(1983, 1986, 1998))

ss_dln_fy_combined_lagged_trimmed %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

ss_dln_fy_lm_age0_trimmed <- lm(age0_lead1 ~ larvae_index, data = ss_dln_fy_combined_lagged_trimmed)
summary(ss_dln_fy_lm_age0_trimmed)

ss_dln_fy_lm_age0_trimmed_detrend <- lm(age0_lead1 ~ larvae_index + year, data = ss_dln_fy_combined_lagged_trimmed)
summary(ss_dln_fy_lm_age0_trimmed_detrend)



```


# Notes

**Check** if CDFW or another source has data that could proxy better than STAR for SSB. For example, taking Erica's approach, is there useful landings data for cabezon?

**Take** annual values for our significant covariates (temp and O2) and run them against your indices. Is there any visual relationship? Qualitative similarity between the time-series?

# Heatmaps for winning models

```{r}
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

utm_crs <- 32611

ss_map_pred <- predict(
  ss_cab_fit1,
  newdata = ss_full_pred_grid,
  type = "response"
)

pred_plot <- ss_map_pred |>
  mutate(
    expected = est1 * est2,
    X_m = X * 1000,
    Y_m = Y * 1000
  )

ca_coast <- ne_states(
  country = "United States of America",
  returnclass = "sf"
) |>
  dplyr::filter(name == "California") |>
  st_transform(utm_crs)

calcofi_stations <- cabezon_shelf_spawn |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000
  ) |>
  distinct(line, station, X_m, Y_m) |>
  st_as_sf(coords = c("X_m", "Y_m"), crs = utm_crs)

calcofi_lines <- calcofi_stations |>
  arrange(line, station) |>
  group_by(line) |>
  summarize(do_union = FALSE, .groups = "drop") |>
  st_cast("LINESTRING")

pred_plot <- ss_map_pred |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000
  )


# Plot showing larval observations by station over time
obs_plot <- cabezon_shelf_spawn |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000,
    larvae_positive = larvae_100m3 > 0
  )

ggplot() +
  geom_sf(
    data = ca_coast,
    inherit.aes = FALSE,
    fill = "grey80",
    color = "grey40"
  ) +
  geom_sf(
    data = calcofi_lines,
    inherit.aes = FALSE,
    color = "grey95",
    linewidth = 0.25
  ) +
  geom_point(
    data = obs_plot |> filter(!larvae_positive),
    aes(x = X_m, y = Y_m),
    size = 0.4,
    alpha = 0.25
  ) +
  geom_point(
    data = obs_plot |> filter(larvae_positive),
    aes(x = X_m, y = Y_m, size = larvae_100m3),
    alpha = 0.7
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(obs_plot$X_m, na.rm = TRUE),
    ylim = range(obs_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  facet_wrap(~ year) +
  scale_size_continuous(
    name = "Observed larvae / 100 m³",
    range = c(1, 7)
  ) +
  labs(
    title = "Observed cabezon larval abundance by CalCOFI station",
    x = NULL,
    y = NULL
  ) +
  theme_minimal()

# Plot showing only delta (distribution)
ggplot() +
  geom_raster(
    data = pred_plot,
    aes(x = X_m, y = Y_m, fill = est1)
  ) +
  geom_sf(data = ca_coast, inherit.aes = FALSE, fill = "grey80") +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(pred_plot$X_m, na.rm = TRUE),
    ylim = range(pred_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  facet_wrap(~ year) +
  scale_fill_viridis_c(name = "Predicted larvae") +
  theme_minimal()

ggplot() +
  geom_raster(
    data = pred_plot,
    aes(x = X_m, y = Y_m, fill = est2)
  ) +
  geom_sf(data = ca_coast, inherit.aes = FALSE, fill = "grey80") +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(pred_plot$X_m, na.rm = TRUE),
    ylim = range(pred_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  facet_wrap(~ year) +
  scale_fill_viridis_c(name = "Predicted larvae") +
  theme_minimal()

ggplot() +
  geom_raster(
    data = pred_plot,
    aes(x = X_m, y = Y_m, fill = log(est))
  ) +
  geom_sf(data = ca_coast, inherit.aes = FALSE, fill = "grey80") +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(pred_plot$X_m, na.rm = TRUE),
    ylim = range(pred_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  facet_wrap(~ year) +
  scale_fill_viridis_c(name = "Predicted larvae ln(abun)") +
  theme_minimal()

```


# Additional Models


##10.7. ss_cab_fit2a — Tweedie with iid spatiotemporal

```{r}
#---- 10.2.a. Calibration summaries (Tweedie; single component)

ss_fit2a_pred_obs <- predict(ss_cab_fit2a, type = "response")

# Overall mean calibration (no est1/est2 for Tweedie)
ss_fit2a_pred_obs %>%
  summarise(
    obs_mean  = mean(larvae_100m3, na.rm = TRUE),
    pred_mean = mean(est, na.rm = TRUE),
    obs_occ   = mean(larvae_100m3 > 0, na.rm = TRUE),
    # Tweedie has no explicit p(encounter); use threshold on est as a rough proxy
    pred_occ_proxy = mean(est > 0, na.rm = TRUE)
  )

#---- 10.2.b. Predicted vs. observed plots

# Raw scale
plot(
  x = ss_fit2a_pred_obs$est,
  y = cabezon_shelf_spawn$larvae_100m3,
  xlab = "Predicted",
  ylab = "Observed"
)
abline(0, 1, lty = 2)

# Log scale
plot(
  log1p(ss_fit2a_pred_obs$est),
  log1p(cabezon_shelf_spawn$larvae_100m3),
  xlab = "log(Predicted + 1)",
  ylab = "log(Observed + 1)"
)
abline(0, 1, lty = 2)

#---- 10.2.c. Randomized quantile residuals

ss_fit2a_rq_res <- residuals(ss_cab_fit2a, type = "mle-mvn")

# QQ plot
qqnorm(ss_fit2a_rq_res)
qqline(ss_fit2a_rq_res)

# Spatial residual map by year
ss_fit2a_pred_obs$rq_resid <- ss_fit2a_rq_res

ggplot(ss_fit2a_pred_obs, aes(x = longitude, y = latitude, color = rq_resid)) +
  geom_point() +
  scale_color_gradient2() +
  facet_wrap(~ year)

#---- 10.2.d. Temporal calibration

ss_fit2a_pred_obs %>%
  group_by(year) %>%
  summarize(
    obs  = mean(larvae_100m3),
    pred = mean(est)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = obs), color = "black") +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed")

#---- 10.2.e. Abundance index: regular prediction grid
# fyear must be present and match levels in training data
# Note: 1982 is extra_time — fixed effect for that level is interpolated by sdmTMB

ss_fit2a_pred_grid <- expand.grid(
  X = seq(min(cabezon_shelf_spawn$X), max(cabezon_shelf_spawn$X), by = 5),
  Y = seq(min(cabezon_shelf_spawn$Y), max(cabezon_shelf_spawn$Y), by = 5)
) %>%
  tidyr::crossing(year = as.integer(sort(unique(cabezon_shelf_spawn$year)))) %>%
  mutate(fyear = as.factor(year))

ss_fit2a_cab_map <- predict(ss_cab_fit2a,
                            newdata   = ss_fit2a_pred_grid,
                            return_tmb_object = TRUE)

ss_fit2a_cab_index <- get_index(ss_fit2a_cab_map, area = 1, bias_correct = TRUE)

ggplot(ss_fit2a_cab_index, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Tweedie IID (fixed year effects, regular grid)")

#---- 10.2.f. Abundance index: observed station grid (preferred)

ss_fit2a_pred_grid_obs <- cabezon_shelf_spawn %>%
  dplyr::select(X, Y, year) %>%
  distinct() %>%
  mutate(fyear = as.factor(year))

ss_fit2a_cab_map_obs <- predict(ss_cab_fit2a,
                                newdata = ss_fit2a_pred_grid_obs,
                                return_tmb_object = TRUE)

ss_fit2a_cab_index_obs <- get_index(ss_fit2a_cab_map_obs,
                                    area = 1,
                                    bias_correct = TRUE)

ggplot(ss_fit2a_cab_index_obs, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Tweedie IID (fixed year effects, observed grid)")


# CUSUM
ss_full_pred_obs %>%
  group_by(year) %>%
  summarise(
    med_resid1 = median(rq_resid1, na.rm = TRUE),
    med_resid2 = median(rq_resid2, na.rm = TRUE),  # positive tows only
    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    cusum1 = cumsum(replace_na(med_resid1, 0)),
    cusum2 = cumsum(replace_na(med_resid2, 0))
  ) %>%
  pivot_longer(cols = c(cusum1, cusum2),
               names_to = "component",
               values_to = "cusum") %>%
  ggplot(aes(x = year, y = cusum, color = component)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("cusum1" = "steelblue", "cusum2" = "firebrick"),
                     labels = c("Delta (binomial)", "Gamma")) +
  labs(x = "Year", y = "CUSUM", title = "CUSUM by Delta Model Component — ss_cab_fit1") +
  theme_minimal()


library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

utm_crs <- 32611

ca_coast <- ne_states(
  country = "United States of America",
  returnclass = "sf"
) |>
  dplyr::filter(name == "California") |>
  st_transform(utm_crs)

calcofi_stations <- cabezon_shelf_spawn |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000
  ) |>
  distinct(line, station, X_m, Y_m) |>
  st_as_sf(coords = c("X_m", "Y_m"), crs = utm_crs)

calcofi_lines <- calcofi_stations |>
  arrange(line, station) |>
  group_by(line) |>
  summarize(do_union = FALSE, .groups = "drop") |>
  st_cast("LINESTRING")

pred_plot <- ss_map_pred |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000
  )

# Presence + Density
ggplot() +
  geom_raster(
    data = pred_plot_tw,
    aes(x = X_m, y = Y_m, fill = est)
  ) +
  geom_sf(
    data = ca_coast,
    inherit.aes = FALSE,
    fill = "grey80",
    color = "grey40"
  ) +
  geom_sf(
    data = calcofi_lines,
    inherit.aes = FALSE,
    color = "white",
    linewidth = 0.3
  ) +
  geom_sf(
    data = calcofi_stations,
    inherit.aes = FALSE,
    size = 0.5
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(pred_plot$X_m, na.rm = TRUE),
    ylim = range(pred_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "Predicted larvae") +
  theme_minimal()

# Density component
ggplot() +
  geom_raster(
    data = pred_plot_tw,
    aes(x = X_m, y = Y_m, fill = est2)
  ) +
  geom_sf(
    data = ca_coast,
    inherit.aes = FALSE,
    fill = "grey80",
    color = "grey40"
  ) +
  geom_sf(
    data = calcofi_lines,
    inherit.aes = FALSE,
    color = "white",
    linewidth = 0.3
  ) +
  geom_sf(
    data = calcofi_stations,
    inherit.aes = FALSE,
    size = 0.5
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(pred_plot$X_m, na.rm = TRUE),
    ylim = range(pred_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "Predicted larvae") +
  theme_minimal()

# Tweedie maps
ss_map_pred_tw <- predict(
  ss_cab_fit2,
  newdata = ss_fit2a_pred_grid,
  type = "response"
)

pred_plot_tw <- ss_map_pred_tw |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000
  )

# Plot showing larval observations by station over time
obs_plot_tw <- cabezon_shelf_spawn |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000,
    larvae_positive = larvae_100m3 > 0
  )

ggplot() +
  geom_sf(
    data = ca_coast,
    inherit.aes = FALSE,
    fill = "grey80",
    color = "grey40"
  ) +
  geom_sf(
    data = calcofi_lines,
    inherit.aes = FALSE,
    color = "grey95",
    linewidth = 0.25
  ) +
  geom_point(
    data = obs_plot_tw |> filter(!larvae_positive),
    aes(x = X_m, y = Y_m),
    size = 0.4,
    alpha = 0.25
  ) +
  geom_point(
    data = obs_plot_tw |> filter(larvae_positive),
    aes(x = X_m, y = Y_m, size = larvae_100m3),
    alpha = 0.7
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(obs_plot$X_m, na.rm = TRUE),
    ylim = range(obs_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  facet_wrap(~ year) +
  scale_size_continuous(
    name = "Observed larvae / 100 m³",
    range = c(1, 7)
  ) +
  labs(
    title = "Observed cabezon larval abundance by CalCOFI station",
    x = NULL,
    y = NULL
  ) +
  theme_minimal()

# Plot showing delta + hurdle components (distribution and abundance)
ggplot() +
  geom_raster(
    data = pred_plot_tw,
    aes(x = X_m, y = Y_m, fill = est)
  ) +
  geom_sf(data = ca_coast, inherit.aes = FALSE, fill = "grey80") +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(pred_plot_tw$X_m, na.rm = TRUE),
    ylim = range(pred_plot_tw$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  facet_wrap(~ year) +
  scale_fill_viridis_c(name = "Predicted larvae") +
  theme_minimal()
```

## 10.8. ss_cab_fit2a — STAR correlation analysis

```{r}
#---- 10.8. ss_cab_fit2a STAR comparisons and CCF

# Create one data frame to house all STAR + larval index data
# Note: using observed station grid index (10.2.f) as preferred basis
tw_fy_combined <- ss_fit2a_cab_index_obs %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
tw_fy_combined_obs <- tw_fy_combined %>%
  filter(!is.na(larvae_index))

# CCF: Tweedie IID larval index vs. age-0 recruits
ccf(tw_fy_combined_obs$larvae_index, tw_fy_combined_obs$age0,
    lag.max = 10,
    main = "CCF: Tweedie IID Larval Index vs. Age-0 Recruits")

# CCF: Tweedie IID larval index vs. recruitment deviations
ccf(tw_fy_combined_obs$larvae_index, tw_fy_combined_obs$rec_dev,
    lag.max = 10,
    main = "CCF: Tweedie IID Larval Index vs. Recruitment Deviations")

# CCF: Tweedie IID larval index vs. SSB (raw)
# Note: shared declining trend may inflate signal — detrended version below is preferred
ccf(tw_fy_combined_obs$larvae_index, tw_fy_combined_obs$ssb,
    lag.max = 10,
    main = "CCF: Tweedie IID Larval Index vs. Spawning Stock Biomass (raw)")

# Detrend both series via loess before comparing against SSB
tw_fy_larvae_detrended <- residuals(loess(larvae_index ~ year, data = tw_fy_combined_obs, span = 0.75))
tw_fy_ssb_detrended    <- residuals(loess(ssb ~ year,          data = tw_fy_combined_obs, span = 0.75))

ccf(tw_fy_larvae_detrended, tw_fy_ssb_detrended,
    lag.max = 10,
    main = "CCF: Tweedie IID Larval Index vs. SSB (detrended)")
```

## 10.9. ss_cab_fit2a — Correlation analysis

```{r}
#---- 10.9. ss_cab_fit2a diagnostics and lagged regression

# RQ residuals: annual distributions
ss_fit2a_pred_obs %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method    = "loess", span = 0.4,
    se        = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Year",
    y     = "RQ Residual",
    title = "Annual RQ Residual Distributions — ss_cab_fit2a"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

# CUSUM of median RQ residuals (single component; Tweedie)
ss_fit2a_pred_obs %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x     = "Year",
    y     = "Cumulative Sum of Median RQ Residual",
    title = "CUSUM — ss_cab_fit2a"
  ) +
  theme_minimal()

#---- Simple lagged regression at lag -1

tw_fy_combined_lagged <- tw_fy_combined_obs %>%
  arrange(year) %>%
  mutate(
    age0_lead1    = lead(age0, 1),
    rec_dev_lead1 = lead(rec_dev, 1)
  )

tw_fy_lm_age0   <- lm(age0_lead1    ~ larvae_index, data = tw_fy_combined_lagged)
tw_fy_lm_recdev <- lm(rec_dev_lead1 ~ larvae_index, data = tw_fy_combined_lagged)

summary(tw_fy_lm_age0)
summary(tw_fy_lm_recdev)

tw_fy_combined_lagged %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

# Sensitivity: remove 1983
tw_fy_combined_lagged_no83 <- tw_fy_combined_lagged %>% filter(year != 1983)
tw_fy_lm_age0_no83 <- lm(age0_lead1 ~ larvae_index, data = tw_fy_combined_lagged_no83)
summary(tw_fy_lm_age0_no83)

# Detrended regressions
tw_fy_lm_age0_detrend   <- lm(age0_lead1    ~ larvae_index + year, data = tw_fy_combined_lagged)
tw_fy_lm_recdev_detrend <- lm(rec_dev_lead1 ~ larvae_index + year, data = tw_fy_combined_lagged)
summary(tw_fy_lm_age0_detrend)
summary(tw_fy_lm_recdev_detrend)

# Cook's distance
plot(tw_fy_lm_age0, which = 4)

# Trimmed analysis: remove three most anomalous years
tw_fy_combined_lagged_trimmed <- tw_fy_combined_lagged %>%
  filter(!year %in% c(1983, 1986, 1998))

tw_fy_combined_lagged_trimmed %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

tw_fy_lm_age0_trimmed <- lm(age0_lead1 ~ larvae_index, data = tw_fy_combined_lagged_trimmed)
summary(tw_fy_lm_age0_trimmed)

tw_fy_lm_age0_trimmed_detrend <- lm(age0_lead1 ~ larvae_index + year, data = tw_fy_combined_lagged_trimmed)
summary(tw_fy_lm_age0_trimmed_detrend)

# Is underprediction spatially structured?
ss_fit2a_pred_obs %>%
  group_by(station) %>%
  summarise(mean_resid = mean(rq_resid), .groups = "drop") %>%
  ggplot(aes(x = station, y = mean_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = TRUE) +
  labs(x = "Station", y = "Mean RQ Residual",
       title = "Mean Residual by Station — ss_cab_fit2a") +
  theme_minimal()

# AIC comparison: Tweedie variants
AIC(ss_cab_fit2, ss_cab_fit2a)
```


