---
title: "Cabezon sdmTMB — Constrained Analysis"
author: "Dan Portnof"
date: "2026-02-25"
output: word_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


---

# PART II — CONSTRAINED ANALYSIS (SPAWNING SEASON + SHELF REGION)

---

To refresh our memory, all analysis to this point has been done using CalCOFI data in the core CalCOFI grid, extended to the maximum distance from shore that these transect lines go (in excess of some 250 mi). Additionally, our data compasses all months and seasons over the observed time-series from 1981-2015.

This section will attempt to apply a finer grade to our analysis by constraining these parameters in the following ways:

1. We will focus on spawning season, defined in the 2019 stock assessment as **beginning in November and ending in March, with a peak in January and February.**

2. We will focus on the shelf rather than the full distance offshore, assuming that cabezon larvae occur most frequently and most regularly inshore.

It's worth keeping in mind, of course, that this constrains our spatial data further, *but,* our natural science might suggest that a narrower examination of spawning season larval abundance *in areas where we'd more likely expect larval occurrence* could be a better match to downstream adult distribution and abundance.

# 0. SETUP & STAR REPORT (CONSTRAINED)

```{r 0. SETUP & STAR REPORT (Part II)}
# ---- 0. SETUP & STAR REPORT (Part II)

# Load packages
library(tidyverse)
library(r4ss)
library(sdmTMB)
library(fmesher)
library(funtimes)

my.seed <- 666
set.seed(my.seed)

wd <- "Cab_SCS_BC_STAR" # Mac

# This reads the output files
pp <- SS_output(wd)

# Pull out derived quantities.  These are things like SSB and age-0 recruits.
derived_quants <- pp[["derived_quants"]]

# Pull out parameter estimates.  Recruitment deviations are estimated parameters.
params <- pp[["parameters"]]

# Plot recruitment deviations for the main estimation period.
devyrs <- 1970:2018
devdf <- pp$parameters[28:76, ]
devvalue <- devdf[,3]
devsd <- devdf[,11]
devlower <- (devvalue-1.96*devsd)
devmed   <- devvalue
devupper <- (devvalue+1.96*devsd)
STAR_recdevs <- data.frame(year=devyrs, value=devmed,lo=devlower,hi=devupper)

STAR_recdevs_plot <- ggplot(data=STAR_recdevs, aes(x=year, ymin=devlower, ymax=devupper)) +
  geom_ribbon(fill="skyblue", alpha=0.5) +
  geom_line(aes(y=devmed)) +
  labs(title="Cabezon SoCal Recruitment Deviations") +
  theme_bw()

STAR_recdevs_plot

# Plot age-0 recruits
age0yrs <- 1970:2018
startrowindex <- which(rownames(derived_quants)=="Recr_1970")
endrowindex <- which(rownames(derived_quants)=="Recr_2018")
age0 <- derived_quants[174:222, "Value"]
age0sd <- derived_quants[174:222, "StdDev"]
age0lower <- (age0-1.96*age0sd)
age0upper <- (age0+1.96*age0sd)

STAR_age0 <- data.frame(year=age0yrs, value=age0,lo=age0lower,hi=age0upper)

STAR_age0_plot <- ggplot(data=STAR_age0, aes(x=year, ymin=age0lower, ymax=age0upper)) +
  geom_ribbon(fill="skyblue", alpha=0.5) +
  geom_line(aes(y=age0)) +
  labs(title="Cabezon SoCal Age 0 Recruits") +
  theme_bw()

STAR_age0_plot

# Plot SSB
ssbyrs <- 1916:2018
startrowindex <- which(rownames(derived_quants)=="SSB_1916")
endrowindex <- which(rownames(derived_quants)=="SSB_2018")
ssb <- derived_quants[3:105, "Value"]
ssbsd <- derived_quants[3:105, "StdDev"]
ssblower <- (ssb-1.96*ssbsd)
ssbupper <- (ssb+1.96*ssbsd)

STAR_SSB <- data.frame(year=ssbyrs, value=ssb,lo=ssblower,hi=ssbupper)

STAR_SSB_plot <- ggplot(data=STAR_SSB, aes(x=year, ymin=ssblower, ymax=ssbupper)) +
  geom_ribbon(fill="skyblue", alpha=0.5) +
  geom_line(aes(y=ssb)) +
  labs(title="Cabezon SoCal SSB") +
  theme_bw()

STAR_SSB_plot

```

# 1. DATA QC + EXPLORATION (CONSTRAINED)

```{r 1. DATA QC (Part II), cache = T}
#---- DATA QC + EXPLORATION (Part II)

cabezon <- read.csv("cabezon_calcofi_data.csv")

str(cabezon)

# Add factor variable for season
cabezon <- cabezon %>%
  mutate(season = as.factor(season))

# ---- 1. DATA QC + EXPLORATION (Part II)

# Check for how many observations per ...

# year
cabezon %>%
  count(year, season, line.station) %>%
  filter(n > 1)
# month
cabezon %>%
  count(year, month, line.station) %>%
  filter(n > 1)
# cruise
cabezon %>%
  count(cruise, line.station) %>%
  filter(n > 1)

# Station coverage across years
cabezon %>%
  group_by(year) %>%
  summarize(n_stations = n_distinct(line.station))
table(cabezon$line.station, cabezon$year)
  # Pretty steady coverage, especially post-1989

# Check the spatial grid for a quick visual
ggplot(cabezon, aes(longitude, latitude)) +
  geom_point(alpha = 0.3)

# How much zero inflation do we have?
cabezon %>%
  mutate(zero = larvae_100m3 == 0) %>%
  count(zero)
  # Roughly ~85% zeros

# Raw distribution visual
plot(cabezon$volume_sampled, cabezon$larvae_100m3)

pos_vals <- cabezon$larvae_100m3[cabezon$larvae_100m3 > 0]
hist(log(pos_vals), breaks = 50, main = "log(positive larvae_100m3)")
hist(log(cabezon$larvae_100m3[cabezon$larvae_100m3 > 0]))

cabezon %>%
  filter(larvae_100m3 > 0) %>%
  group_by(year) %>%
  summarize(m = mean(larvae_100m3), v = var(larvae_100m3), n = n()) %>%
  filter(n > 5) %>%
   ggplot(aes(log(m), log(v))) + geom_point() + geom_smooth(method = "lm") +
  geom_abline(slope = 2, intercept = 0, linetype = "dashed", color = "red") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Mean-variance relationship (dashed: red=gamma slope 2, blue=Poisson slope 1)")

# Also useful: what fraction are zeros?
mean(cabezon$larvae_100m3 == 0)

```

# 2. DATA PARING & EXPLORATION (CONSTRAINED)

```{r 2. DATA PARING AND EXPLORATION (CONSTRAINED), cache = T}

cabezon_shelf_spawn <- cabezon %>%
  filter(
    month %in% c(1, 2, 3),
    station <= 60.0
  )

cat("Rows retained:", nrow(cabezon_shelf_spawn), "of", nrow(cabezon), "\n")
cat("Years covered:", paste(range(cabezon_shelf_spawn$year), collapse = "–"), "\n")
cat("Months present:", paste(sort(unique(cabezon_shelf_spawn$month)), collapse = ", "), "\n")
cat("Station range:", paste(range(cabezon_shelf_spawn$station), collapse = "–"), "\n")
cat("Lines present:", paste(sort(unique(cabezon_shelf_spawn$line)), collapse = ", "), "\n")

cabezon_shelf_spawn %>%
  group_by(year) %>%
  summarize(
    n_tows = n(),
    n_pos = sum(larvae_100m3 > 0),
    pct_pos = round(100 * n_pos / n_tows, 1)
  ) %>%
  print(n = Inf)

# Composite station column (combines line + station)
cabezon_shelf_spawn <- cabezon_shelf_spawn %>%
  mutate(station_id = paste(line, station, sep = "_"))

# Do non-standard hauls have systematically different standard_haul_factor values?
summary(cabezon_shelf_spawn$standard_haul_factor)

# Are any values notably far from 1.0?
hist(cabezon_shelf_spawn$standard_haul_factor)
hist(log(cabezon_shelf_spawn$larvae_100m3))

# Check the spike at larvae_100m3 == 1
cabezon_shelf_spawn %>% count(larvae_100m3 == 1)

# Check how many true zeros are present
cabezon_shelf_spawn %>% count(larvae_100m3 == 0)

# Confirm no -Inf values making it into model inputs
sum(is.infinite(log(cabezon_shelf_spawn$larvae_100m3)))

cabezon_shelf_spawn %>%
  filter(standard_haul_factor != 0.02) %>%
  count(year) 

# Calculating positive tow values for visualization
ss_pos_tows_year <- cabezon_shelf_spawn %>%
  group_by(year) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )

ss_pos_tows_month <- cabezon_shelf_spawn %>%
  group_by(month) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )

ss_pos_tows_season <- cabezon_shelf_spawn %>%
  group_by(season) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )

ss_pos_tows_station <- cabezon_shelf_spawn %>%
  group_by(station_id) %>%
  summarise(
    total_tows        = n(),
    positive_tows     = sum(larvae_100m3 > 0),
    pct_positive_tows = positive_tows / total_tows * 100
  )

# Annual mean larval observations with SE
annual_cab_ss_summary <- cabezon_shelf_spawn %>%
  group_by(year) %>%
  summarize(
    mean_larvae = mean(larvae_100m3, na.rm = T),
    sd_larvae = sd(larvae_100m3, na.rm = T),
    n = n(),
    se = sd_larvae / sqrt(n)
  )

# ---- 8.b. Annual average larval observations

ggplot(annual_cab_ss_summary, aes(x = year, y = mean_larvae)) +
  geom_ribbon(aes(ymin = mean_larvae - se, ymax = mean_larvae + se),
              alpha = 0.5, fill = "skyblue") +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue") +
  labs(
    x = "Year",
    y = expression("Mean larvae per 100 m"^3),
    title = "Annual Mean Larval Observations, Constrained to Spawning Season and Coastal Shelf",
    subtitle = expression("Ribbon shows " %+-% " 1 SE")
  ) +
  theme_bw()

# ---- 8.c. Stacked bar chart: positive vs. total tows by year, month, season

ss_plot_data <- bind_rows(
  ss_pos_tows_year   %>% mutate(group_var = "Year",   level = as.character(year)),
  ss_pos_tows_month  %>% mutate(group_var = "Month",  level = as.character(month)),
  ss_pos_tows_season %>% mutate(group_var = "Season", level = as.character(season))
) %>%
  mutate(negative_tows = total_tows - positive_tows) %>%
  pivot_longer(
    cols      = c(positive_tows, negative_tows),
    names_to  = "tow_type",
    values_to = "count"
  )

ss_plot_data <- ss_plot_data %>%
  mutate(level = case_when(
    group_var == "Month" ~ factor(level, levels = as.character(1:12)),
    group_var == "Season" ~ factor(level, levels = c("winter", "spring", "summer", "fall")),
    T ~ factor(level)
  ))

# Shared elements
shared_fill <- scale_fill_manual(
  values = c("negative_tows" = "grey80", "positive_tows" = "steelblue"),
  labels = c("negative_tows" = "Larvae Absent", "positive_tows" = "Larvae Present")
)

shared_bar_theme <- theme_minimal(base_size = 12) +
  theme(
    strip.text         = element_text(face = "bold", size = 13),
    axis.text.x        = element_text(angle = 45, hjust = 1),
    legend.position    = "bottom",
    panel.grid.major.x = element_blank(),
    plot.title         = element_text(face = "bold", size = 15),
    plot.subtitle      = element_text(color = "grey40", size = 10)
  )

# One plot for years; one plot for months + seasons
ss_plot_data_year <- ss_plot_data %>% filter(group_var == "Year")
ss_plot_data_ms   <- ss_plot_data %>% filter(group_var %in% c("Month", "Season"))

# Years plot
ggplot(ss_plot_data_year, aes(x = level, y = count, fill = tow_type)) +
  geom_bar(stat = "identity", width = 0.7) +
  shared_fill +
  shared_bar_theme +
  labs(
    title    = "Cabezon Larvae: Positive vs. Total Tows",
    subtitle = "Stacked bars show larvae-present (blue) against total sampling effort (gray)",
    x        = NULL,
    y        = "Number of Tows",
    fill     = NULL
  )

# Months + Seasons plot
ggplot(ss_plot_data_ms, aes(x = level, y = count, fill = tow_type)) +
  geom_bar(stat = "identity", width = 0.7) +
  facet_wrap(~ group_var, scales = "free_x", nrow = 1) +
  shared_fill + shared_bar_theme +
  labs(
    title = "Cabezon Larvae: Positive vs. Total Tows",
    subtitle = "Stacked bar charts show larve-present (blue) against total sampling effort (gray)",
    x        = NULL,
    y        = "Number of Tows",
    fill     = NULL
  )

#---- 8.d. Boxplots: log-larvae by year, month, season (positive tows only)

# Data binning
ss_box_data <- bind_rows(
  cabezon_shelf_spawn %>% mutate(group_var = "Year",   level = as.character(year)),
  cabezon_shelf_spawn %>% mutate(group_var = "Month",  level = as.character(month))
) %>%
  filter(larvae_100m3 > 0) %>%
  mutate(
    log_larvae = log(larvae_100m3),
    group_var  = factor(group_var, levels = c("Year", "Month", "Season")),
    level = case_when(
      group_var == "Month"  ~ factor(level, levels = as.character(1:12)),
      group_var == "Season" ~ factor(level, levels = c("winter", "spring", "summer", "fall")),
      TRUE                  ~ factor(level)
    )
  )

# Shared elements
shared_box_theme <- theme_minimal(base_size = 12) +
  theme(
    strip.text         = element_text(face = "bold", size = 13),
    axis.text.x        = element_text(angle = 45, hjust = 1),
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    plot.title         = element_text(face = "bold", size = 15),
    plot.subtitle      = element_text(color = "grey40", size = 10)
  )

# Years plot
ggplot(ss_box_data %>% filter(group_var == "Year"), 
       aes(x = level, y = log_larvae)) +
  geom_boxplot(fill = "steelblue", color = "grey30", alpha = 0.7, outlier.size = 1) +
  shared_box_theme +
  labs(
    title    = "Cabezon Larvae: Average Count by Year (Positive Tows Only)",
    subtitle = "Log-transformed larvae per 100m³; zero-catch tows excluded",
    x        = NULL,
    y        = "log(Larvae per 100m³)"
  )

# Months + Seasons plot
ggplot(ss_box_data %>% filter(group_var %in% c("Month", "Season")),
       aes(x = level, y = log_larvae)) +
  geom_boxplot(fill = "steelblue", color = "grey30", alpha = 0.7, outlier.size = 1) +
  facet_wrap(~ group_var, scales = "free_x", nrow = 1) +
  shared_box_theme +
  labs(
    title    = "Cabezon Larvae: Average Count by Month and Season (Positive Tows Only)",
    subtitle = "Log-transformed larvae per 100m³; zero-catch tows excluded",
    x = NULL, y = "log(Larvae per 100m³)"
  )


#---- 8.e. Heatmaps: mean log-larvae by station x year, month, season

heatmap_theme <- theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 7),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey40", size = 10)
  )

heatmap_fill <- scale_fill_viridis_c(option = "mako", name = "log(Larvae\nper 100m³)", na.value = "white")

# Year plot
cabezon_shelf_spawn %>%
  filter(larvae_100m3 > 0) %>%
  group_by(line.station, year) %>%
  summarise(mean_log_larvae = mean(log(larvae_100m3)), .groups = "drop") %>%
  ggplot(aes(x = year, y = line.station, fill = mean_log_larvae)) +
  geom_tile() + heatmap_fill + scale_y_discrete(limits = rev) + heatmap_theme +
  labs(title    = "Cabezon Larval Observations by Station and Year",
       subtitle = "Positive tows only; white = no positive tows recorded",
       x = NULL, y = "South \u2190 Station \u2192 North")

# Month plot
cabezon_shelf_spawn %>%
  filter(larvae_100m3 > 0) %>%
  group_by(line.station, month) %>%
  summarise(mean_log_larvae = mean(log(larvae_100m3)), .groups = "drop") %>%
  ggplot(aes(x = month, y = line.station, fill = mean_log_larvae)) +
  geom_tile() + heatmap_fill +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_discrete(limits = rev) + heatmap_theme +
  labs(title    = "Cabezon Larval Observations by Station and Month",
       subtitle = "Positive tows only; white = no positive tows recorded",
       x = NULL, y = "South \u2190 Station \u2192 North")

# Season plot
cabezon_shelf_spawn %>%
  filter(larvae_100m3 > 0) %>%
  group_by(line.station, season) %>%
  summarise(mean_log_larvae = mean(log(larvae_100m3)), .groups = "drop") %>%
  ggplot(aes(x = season, y = line.station, fill = mean_log_larvae)) +
  geom_tile() + heatmap_fill + scale_y_discrete(limits = rev) + heatmap_theme +
  labs(title    = "Cabezon Larval Observations by Station and Season",
       subtitle = "Positive tows only; white = no positive tows recorded",
       x = NULL, y = "South \u2190 Station \u2192 North")

ss_pos_vals <- cabezon_shelf_spawn$larvae_100m3[cabezon_shelf_spawn$larvae_100m3 > 0]
hist(log(ss_pos_vals), breaks = 50, main = "log(positive larvae_100m3)")
hist(log(cabezon_shelf_spawn$larvae_100m3[cabezon_shelf_spawn$larvae_100m3 > 0]))

cabezon_shelf_spawn %>%
  filter(larvae_100m3 > 0) %>%
  group_by(year) %>%
  summarize(m = mean(larvae_100m3), v = var(larvae_100m3), n = n()) %>%
  filter(n > 5) %>%
   ggplot(aes(log(m), log(v))) + geom_point() + geom_smooth(method = "lm") +
  geom_abline(slope = 2, intercept = 0, linetype = "dashed", color = "red") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Mean-variance relationship (dashed: red=gamma slope 2, blue=Poisson slope 1)")

# Back out the raw count from density and volume
cabezon_shelf_spawn %>%
  filter(larvae_100m3 > 0) %>%
  mutate(
    raw_count_est = larvae_100m3 * volume_sampled / 100,
    log_density = log(larvae_100m3)
  ) %>%
  ggplot(aes(log_density, fill = factor(round(raw_count_est)))) +
  geom_histogram(bins = 50, position = "stack") +
  scale_fill_viridis_d() +
  labs(fill = "Estimated raw count",
       title = "Does multi-modality decompose into integer count strata?")

```

# 3. sdmTMB MODEL FITTING (CONSTRAINED)

```{r 3. sdmTMB MODEL FITTING (CONSTRAINED), cache = T}

set.seed(my.seed)

#---- 3.a. Build mesh
if (!all(c("X", "Y") %in% names(cabezon_shelf_spawn))) {
  cabezon_shelf_spawn <- sdmTMB::add_utm_columns(cabezon_shelf_spawn, ll_names = c("longitude", "latitude"))
}

ss_mesh <- fm_mesh_2d(
  loc = cabezon_shelf_spawn[,c("X","Y")],
  cutoff = 20,
  max.edge = c(75, 150),
  offset = c(45, 120)
)
ss_cab_mesh <- make_mesh(data = cabezon_shelf_spawn, c("X", "Y"), mesh = ss_mesh)
plot(ss_cab_mesh); title("SPDE Mesh for Cabezon CalCOFI analysis (Constrained)")

#---- 3.b. Fit candidate models

# Add year factor variable
cabezon_shelf_spawn$fyear <- as.factor(cabezon_shelf_spawn$year)

# NULL Delta-Gamma model

ss_cab_null <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = "off",
  spatiotemporal = "off",
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_null)

# NULL Delta-Lognormal model

ss_cab_null_dln <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = "off",
  spatiotemporal = "off",
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_null_dln)
AIC(ss_cab_null_dln)

# NULL Tweedie model

ss_cab_null_tw <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = tweedie(),
  spatial = "off",
  spatiotemporal = "off",
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_null_tw)

null_AIC_table <- data.frame(
  fit_name = c(
    "ss_cab_null",
    "ss_cab_null_dln",
    "ss_cab_null_tw"
  ),
  model = c(
    "Delta-Gamma NULL (Constrained)",
    "Delta-Lognormal NULL (Constrained)",
    "Tweedie NULL (Constrained)"
  ),
  AIC = c(
    AIC(ss_cab_null),
    AIC(ss_cab_null_dln),
    AIC(ss_cab_null_tw)
  ),
    delta_AIC = NA
  )

null_AIC_table$delta_AIC <- null_AIC_table$AIC - min(null_AIC_table$AIC)

null_AIC_table <- null_AIC_table[order(null_AIC_table$AIC, decreasing = F), ]

null_AIC_table

# Delta-Lognormal: Global Intercept, spatial on, spatiotemporal off
ss_cab_fit_dln0 <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("off", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit_dln0)

AIC(ss_cab_fit_dln0)

# Delta-Lognormal: Global Intercept, spatial on, spatiotemporal delta AR1, spatiotemporal lognormal AR1
ss_cab_fit_dln1 <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("ar1", "ar1"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit_dln1)

AIC(ss_cab_fit_dln1)

# Delta-Lognormal: Global Intercept, spatial on, spatiotemporal delta AR1, spatiotemporal lognormal iid
ss_cab_fit_dln2 <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("ar1", "iid"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit_dln2)

AIC(ss_cab_fit_dln2)

# Delta-Lognormal: Global Intercept, spatial on, spatiotemporal delta iid, spatiotemporal lognormal off
ss_cab_fit_dln3 <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("iid", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit_dln3)

AIC(ss_cab_fit_dln3)

# Delta-Lognormal: fixed year effect, spatial on, spatiotemporal off
ss_cab_fit_dln4 <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("off", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit_dln4)

AIC(ss_cab_fit_dln4)

# Delta-Lognormal: fixed year effect, spatial on, spatiotemporal iid
ss_cab_fit_dln5 <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("iid", "iid"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit_dln5)

AIC(ss_cab_fit_dln5)

# Delta-Lognormal: fixed year effect, spatial on, iid delta spatiotemporal, lognormal spatiotemporal off
ss_cab_fit_dln6 <- sdmTMB(
  larvae_100m3 ~ 0 + fyear,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("iid", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit_dln6)

AIC(ss_cab_fit_dln6)

# Delta-Lognormal: Global Intercept, spatial on, spatiotemporal delta AR1, spatiotemporal lognormal off
ss_cab_fit_dln7 <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_lognormal(),
  spatial = list("on", "on"),
  spatiotemporal = list("ar1", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit_dln7)

AIC(ss_cab_fit_dln7)


#---- 3.c. Random effect parameter estimates for best models

tidy(ss_cab_fit_dln1, effects = "ran_pars", model = 1); tidy(ss_cab_fit_dln1, effects = "ran_pars", model = 2)
tidy(ss_cab_fit_dln2, effects = "ran_pars", model = 1); tidy(ss_cab_fit_dln2, effects = "ran_pars", model = 2)
tidy(ss_cab_fit_dln3, effects = "ran_pars", model = 1); tidy(ss_cab_fit_dln3, effects = "ran_pars", model = 2)
tidy(ss_cab_fit_dln4, effects = "ran_pars")
tidy(ss_cab_fit_dln5, effects = "ran_pars")

#---- 3.d. Construct model comparison table

AIC_table <- data.frame(
  fit_name = c(
    "ss_cab_fit_dln0",
    "ss_cab_fit_dln1",
    "ss_cab_fit_dln2",
    "ss_cab_fit_dln3",
    "ss_cab_fit_dln4",
    "ss_cab_fit_dln5",
    "ss_cab_fit_dln6",
    "ss_cab_fit_dln7"
  ),
  model = c(
    "Delta-Lognormal Global Intercept, spatial on, spatiotemporal off",
    "Delta-Lognormal Global Intercept + spatial + spatiotemporal delta AR1 + spatiotemporal lognormal AR1",
    "Delta-Lognormal Global Intercept + spatial + spatiotemporal delta AR1 + spatiotemporal lognormal iid",
    "Delta-Lognormal Global Intercept + spatial + spatiotemporal delta iid",
    "Delta-Lognormal fixed year effect + spatial + spatiotemporal off",
    "Delta-Lognormal fixed year effect + spatial + spatiotemporal delta iid + spatiotemporal lognormal iid",
    "Delta-Lognormal fixed year effect + spatial + spatiotemporal delta iid + spatiotemporal lognormal off",
    "Delta-Lognormal Global Intercept + spatial + spatiotemporal delta AR1 + spatiotemporal lognormal off"
  ),
  AIC = c(
    AIC(ss_cab_fit_dln0),
    AIC(ss_cab_fit_dln1),
    AIC(ss_cab_fit_dln2),
    AIC(ss_cab_fit_dln3),
    AIC(ss_cab_fit_dln4),
    AIC(ss_cab_fit_dln5),
    AIC(ss_cab_fit_dln6),
    AIC(ss_cab_fit_dln7)
  ),
    delta_AIC = NA
  )

AIC_table$delta_AIC <- AIC_table$AIC - min(AIC_table$AIC)

AIC_table <- AIC_table[order(AIC_table$AIC, decreasing = F), ]

AIC_table

ss_cab_fit_dg1 <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = delta_gamma(),
  spatial = list("on", "on"),
  spatiotemporal = list("ar1", "off"),
  offset = NULL,
  extra_time = 1982
)

sanity(ss_cab_fit_dg1)
AIC(ss_cab_fit_dg1)

tw_iid_fit <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = tweedie(),
  spatial = "on",
  spatiotemporal = "iid",
  offset = NULL,
  extra_time = 1982
)
sanity(tw_iid_fit)

tw_stoff_fit <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = tweedie(),
  spatial = "on",
  spatiotemporal = "off",
  offset = NULL,
  extra_time = 1982
)
sanity(tw_stoff_fit)

tw_ar1_fit <- sdmTMB(
  larvae_100m3 ~ 1,
  data = cabezon_shelf_spawn,
  mesh = ss_cab_mesh,
  time = "year",
  family = tweedie(),
  spatial = "on",
  spatiotemporal = "ar1",
  offset = NULL,
  extra_time = 1982
)
sanity(tw_ar1_fit)

AIC(tw_iid_fit, tw_stoff_fit, tw_ar1_fit)


```


# 4. MODEL DIAGNOSTICS & ABUNDANCE INDICES (CONSTRAINED)

## 4.1. ss_cab_fit1 — Full Delta-Gamma Model

```{r 4.1. ss_cab_fit1 diagnostics, cache = T}

summary(ss_cab_fit_dln3)

#---- 4.1.a. Calibration summaries

set.seed(my.seed)
ss_dln3_pred_obs <- predict(ss_cab_fit_dln3, type = "response")

# Overall mean calibration
ss_dln3_pred_obs %>%
  summarise(
    obs_mean = mean(larvae_100m3, na.rm = TRUE),
    pred_mean = mean(est, na.rm = TRUE),
    obs_occ = mean(larvae_100m3 > 0, na.rm = TRUE),
    pred_occ = mean(est1, na.rm = TRUE),
    obs_pos_mean = mean(larvae_100m3[larvae_100m3 > 0], na.rm = TRUE),
    pred_pos_mean = mean(est2, na.rm = TRUE)
  )

# Binomial calibration; i.e. predicted presence probability vs. observed presence rate
ss_dln3_pred_obs$presence <- ss_dln3_pred_obs$larvae_100m3 > 0

ss_dln3_calibration_summary <- ss_dln3_pred_obs %>%
  mutate(prob_bin = cut(est1, breaks = seq(0, 1, by = 0.1))) %>%
  group_by(prob_bin) %>%
  summarise(
    mean_pred = mean(est1),
    observed = mean(presence),
    n = n()
  )

plot(ss_dln3_calibration_summary$mean_pred, ss_dln3_calibration_summary$observed,
     xlab = "Predicted presence probability",
     ylab = "Observed presence frequency")
abline(0, 1, lty = 2)

#---- 4.1.b. Predicted vs. observed plots

# Raw scale comparison
plot(
  x = ss_dln3_pred_obs$est,
  y = cabezon_shelf_spawn$larvae_100m3,
  xlab = "Predicted",
  ylab = "Observed"
)
abline(0, 1, lty = 2)

# Log scale comparison
plot(log1p(ss_dln3_pred_obs$est), log1p(cabezon_shelf_spawn$larvae_100m3),
     xlab = "log(Predicted + 1)", ylab = "log(Observed + 1)")
abline(0, 1, lty = 2)

# Lognormal (positive density) component only
pos_idx <- cabezon_shelf_spawn$larvae_100m3 > 0
plot(log(ss_dln3_pred_obs$est2[pos_idx]), log(cabezon_shelf_spawn$larvae_100m3[pos_idx]),
     xlab = "log(Predicted positive abundance)",
     ylab = "log(Observed positive abundance)")
abline(0, 1, lty = 2)

#---- 4.1.c. Randomized quantile residuals

set.seed(my.seed)
ss_dln3_rq_res <- residuals(ss_cab_fit_dln3, type = "mle-mvn")

# QQ plot
qqnorm(ss_dln3_rq_res)
qqline(ss_dln3_rq_res)

# Spatial residual map by year
ss_dln3_pred_obs$rq_resid <- ss_dln3_rq_res

ggplot(ss_dln3_pred_obs, aes(x = longitude, y = latitude, color = rq_resid)) +
  geom_point() +
  scale_color_gradient2() +
  facet_wrap(~ year)

#---- 4.1.d. Temporal calibration

# Are predictions tracking temporal trends?
ss_dln3_pred_obs %>%
  group_by(year) %>%
  summarize(
    obs = mean(larvae_100m3),
    pred = mean(est)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = obs), color = "black") +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed")

#---- 4.1.e. Abundance index construction
# Note! The index uncertainty here was far too wide using a regular spatial grid, probably because interpolation over years with no CalCOFI coverage was causing inflation. I kept that plot in for reference, but 4.1.f. (observed station grid) is probably preferred for diagnosis.

ss_dln3_pred_grid <- expand.grid(
  X = seq(min(cabezon_shelf_spawn$X), max(cabezon_shelf_spawn$X), by = 5),
  Y = seq(min(cabezon_shelf_spawn$Y), max(cabezon_shelf_spawn$Y), by = 5)
) %>%
  tidyr::crossing(year = as.integer(sort(unique(cabezon_shelf_spawn$year)))) %>%
  mutate(fyear = as.factor(year))

ss_dln3_cab_map <- predict(ss_cab_fit_dln3, newdata = ss_dln3_pred_grid, return_tmb_object = TRUE)
ss_dln3_cab_index <- get_index(ss_dln3_cab_map, area = 1, bias_correct = TRUE)

ggplot(ss_dln3_cab_index, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Delta-Lognormal on Global Intercept with AR1 (encounter) Spatiotemporal Field")

#---- 4.1.f. Abundance index: constrained to observed station grid (probably preferred)

ss_dln3_pred_grid_obs <- cabezon_shelf_spawn %>%
  dplyr::select(X, Y, year) %>%
  distinct()

ss_dln3_cab_map_obs <- predict(ss_cab_fit_dln3,
                               newdata = ss_dln3_pred_grid_obs,
                               return_tmb_object = TRUE)

ss_dln3_cab_index_obs <- get_index(ss_dln3_cab_map_obs,
                                   area = 1,
                                   bias_correct = TRUE)

ggplot(ss_dln3_cab_index_obs, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Observed Station Grid (Delta-Lognormal, AR1 Encounter)")
```

## 4.2. ss_cab_fit_dln1 — AR1 Delta-Lognormal Model

```{r 4.2. ss_cab_fit_dln1 diagnostics, cache = T}
#---- 4.2.a. Calibration summaries

set.seed(my.seed)
ss_dln_fy_pred_obs <- predict(ss_cab_fit_dln4, type = "response")

# Overall mean calibration
ss_dln_fy_pred_obs %>%
  summarise(
    obs_mean      = mean(larvae_100m3, na.rm = TRUE),
    pred_mean     = mean(est, na.rm = TRUE),
    obs_occ       = mean(larvae_100m3 > 0, na.rm = TRUE),
    pred_occ      = mean(est1, na.rm = TRUE),
    obs_pos_mean  = mean(larvae_100m3[larvae_100m3 > 0], na.rm = TRUE),
    pred_pos_mean = mean(est2, na.rm = TRUE)
  )

# Binomial calibration
ss_dln_fy_pred_obs$presence <- ss_dln_fy_pred_obs$larvae_100m3 > 0

ss_dln_fy_calibration_summary <- ss_dln_fy_pred_obs %>%
  mutate(prob_bin = cut(est1, breaks = seq(0, 1, by = 0.1))) %>%
  group_by(prob_bin) %>%
  summarise(
    mean_pred = mean(est1),
    observed  = mean(presence),
    n         = n()
  )

plot(ss_dln_fy_calibration_summary$mean_pred, ss_dln_fy_calibration_summary$observed,
     xlab = "Predicted presence probability",
     ylab = "Observed presence frequency")
abline(0, 1, lty = 2)

#---- 4.2.b. Predicted vs. observed plots

# Raw scale
plot(
  x    = ss_dln_fy_pred_obs$est,
  y    = cabezon_shelf_spawn$larvae_100m3,
  xlab = "Predicted",
  ylab = "Observed"
)
abline(0, 1, lty = 2)

# Log scale
plot(log1p(ss_dln_fy_pred_obs$est), log1p(cabezon_shelf_spawn$larvae_100m3),
     xlab = "log(Predicted + 1)", ylab = "log(Observed + 1)")
abline(0, 1, lty = 2)

# Lognormal (positive density) component only
pos_idx <- cabezon_shelf_spawn$larvae_100m3 > 0
plot(log(ss_dln_fy_pred_obs$est2[pos_idx]), log(cabezon_shelf_spawn$larvae_100m3[pos_idx]),
     xlab = "log(Predicted positive abundance)",
     ylab = "log(Observed positive abundance)")
abline(0, 1, lty = 2)

#---- 4.2.c. Randomized quantile residuals

set.seed(my.seed)
ss_dln_fy_rq_res <- residuals(ss_cab_fit_dln4, type = "mle-mvn")

# QQ plot
qqnorm(ss_dln_fy_rq_res)
qqline(ss_dln_fy_rq_res)

# Spatial residual map by year
ss_dln_fy_pred_obs$rq_resid <- ss_dln_fy_rq_res

ggplot(ss_dln_fy_pred_obs, aes(x = longitude, y = latitude, color = rq_resid)) +
  geom_point() +
  scale_color_gradient2() +
  facet_wrap(~ year)

#---- 4.2.d. Temporal calibration

ss_dln_fy_pred_obs %>%
  group_by(year) %>%
  summarize(
    obs  = mean(larvae_100m3),
    pred = mean(est)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = obs),  color = "black") +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed")

#---- 4.2.e. Abundance index: regular spatial grid

ss_dln_fy_pred_grid <- expand.grid(
  X = seq(min(cabezon_shelf_spawn$X), max(cabezon_shelf_spawn$X), by = 5),
  Y = seq(min(cabezon_shelf_spawn$Y), max(cabezon_shelf_spawn$Y), by = 5)
) %>%
  tidyr::crossing(year = as.integer(sort(unique(cabezon_shelf_spawn$year)))) %>%
  mutate(fyear = as.factor(year))

ss_dln_fy_cab_map   <- predict(ss_cab_fit_dln4, newdata = ss_dln_fy_pred_grid, return_tmb_object = TRUE)
ss_dln_fy_cab_index <- get_index(ss_dln_fy_cab_map, area = 1, bias_correct = TRUE)

ggplot(ss_dln_fy_cab_index, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Delta-Lognormal Fixed Year Effect, Spatial + Spatiotemporal Off")

#---- 4.2.f. Abundance index: observed station grid (preferred)

ss_dln_fy_pred_grid_obs <- cabezon_shelf_spawn %>%
  dplyr::select(X, Y, year, fyear) %>%
  distinct()

ss_dln_fy_cab_map_obs   <- predict(ss_cab_fit_dln4,
                                   newdata = ss_dln_fy_pred_grid_obs,
                                   return_tmb_object = TRUE)

ss_dln_fy_cab_index_obs <- get_index(ss_dln_fy_cab_map_obs,
                                     area = 1,
                                     bias_correct = TRUE)

ggplot(ss_dln_fy_cab_index_obs, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Observed Station Grid (Delta-Lognormal Fixed Year)")

```

# 5. STAR COMPARISONS & CCFs (CONSTRAINED)

## 5.1. ss_cab_fit1 — STAR Comparisons & CCFs

```{r 5.1. ss_cab_fit1 STAR comparisons full extrapolation, cache = T}
#---- 5.1. ss_cab_fit_dln3 STAR comparisons and CCF
# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 4.1.f. abundance index, assuming it's more stable.
ss_dln3_combined <- ss_dln3_cab_index %>%
  dplyr::select(year, est, lwr, upr) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )
# Filter to observed years
ss_dln3_combined_obs <- ss_dln3_combined %>%
  filter(!is.na(larvae_index))
# CCF: Restricted larval index vs. age-0 recruits
ccf_boot(
  x = ss_dln3_combined_obs$larvae_index,
  y = ss_dln3_combined_obs$age0,
  lag.max = 5,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Age-0 Recruits (Delta-Lognormal)")
# CCF: Restricted larval index vs. recruitment deviations
ccf_boot(
  x = ss_dln3_combined_obs$larvae_index,
  y = ss_dln3_combined_obs$rec_dev,
  lag.max = 5,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Recruitment Deviations (Delta-Lognormal)")
# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf_boot(
  x = ss_dln3_combined_obs$larvae_index,
  y = ss_dln3_combined_obs$ssb,
  lag.max = 5,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. SSB (Delta-Lognormal)")
# First-differenced
ccf_boot(
  x = diff(ss_dln3_combined_obs$larvae_index),
  y = diff(ss_dln3_combined_obs$ssb),
  lag.max = 5,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Δ Larval Index vs. Δ SSB (Delta-Lognormal)")
# And only differenced SSB on raw larvae
ccf_boot(
  x = ss_dln3_combined_obs$larvae_index,
  y = diff(ss_dln3_combined_obs$ssb),
  lag.max = 5,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Larval Index vs. Δ SSB (Delta-Lognormal)")
# Join differenced SSB back to index
ssb_diff_df_dln3 <- ss_dln3_combined_obs %>%
  arrange(year) %>%
  mutate(ssb_diff = c(NA, diff(ssb))) %>%
  filter(!is.na(ssb_diff))
lm_lag0_dln3 <- lm(ssb_diff ~ larvae_index, data = ssb_diff_df_dln3)
summary(lm_lag0_dln3)
# Visual
ssb_diff_df_dln3 %>%
  ggplot(aes(x = larvae_index, y = ssb_diff, label = year)) +
  geom_point() +
  geom_text(nudge_y = 50, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index", y = "ΔSSB (mt)",
       title = "Contemporaneous larval index vs. change in SSB (Delta-Lognormal)") +
  theme_minimal()
# Assuming ecologically that SSB would have the strongest relationship to larval abundance, we'll plot those two time series visually for a qualitative look
scale_factor_ssb_dln3 <- max(ss_dln3_combined_obs$larvae_index, na.rm = TRUE) /
  max(ss_dln3_combined_obs$ssb, na.rm = TRUE)
ggplot(ss_dln3_combined_obs %>% filter(year >= 1984, year <= 2015),
       aes(x = year)) +
  # SSB scaled UP to primary axis units
  geom_ribbon(aes(ymin = ssb_lo * scale_factor_ssb_dln3,
                  ymax = ssb_hi * scale_factor_ssb_dln3),
              fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = ssb * scale_factor_ssb_dln3), color = "steelblue", linewidth = 0.8) +
  # Larvae index on primary axis
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "coral", alpha = 0.3) +
  geom_line(aes(y = larvae_index), color = "coral", linewidth = 0.8) +
  scale_y_continuous(
    name = "Estimated Larval Abundance Index; Delta-Lognormal AR1 (encounter)",
    sec.axis = sec_axis(~ . / scale_factor_ssb_dln3,
                        name = "SSB (mt)")
  ) +
  scale_x_continuous(breaks = seq(1984, 2015, by = 4)) +
  labs(title = "SoCal Cabezon: Larval Index vs. SSB (1984–2015), Delta-Lognormal",
       x = "Year") +
  theme_bw()

# ---- Larval Index vs. Recruitment Deviations ----
# Rec devs are log-scale and centered at 0, so a multiplicative scale_factor
# doesn't behave nicely. Instead, shift + scale rec devs into the index range.

idx_range_dln3   <- range(ss_dln3_combined_obs$larvae_index, na.rm = TRUE)
recdev_range_dln3 <- range(c(ss_dln3_combined_obs$rec_dev_lo,
                             ss_dln3_combined_obs$rec_dev_hi), na.rm = TRUE)

# Linear map from rec dev space -> index space
slope_rd_dln3 <- diff(idx_range_dln3) / diff(recdev_range_dln3)
intercept_rd_dln3 <- idx_range_dln3[1] - slope_rd_dln3 * recdev_range_dln3[1]

ggplot(ss_dln3_combined_obs %>% filter(year >= 1984, year <= 2015),
       aes(x = year)) +
  # Rec devs scaled to primary axis units
  geom_ribbon(aes(ymin = intercept_rd_dln3 + slope_rd_dln3 * rec_dev_lo,
                  ymax = intercept_rd_dln3 + slope_rd_dln3 * rec_dev_hi),
              fill = "darkgreen", alpha = 0.25) +
  geom_line(aes(y = intercept_rd_dln3 + slope_rd_dln3 * rec_dev),
            color = "darkgreen", linewidth = 0.8) +
  geom_hline(yintercept = intercept_rd_dln3 + slope_rd_dln3 * 0,
             linetype = "dashed", color = "darkgreen", alpha = 0.6) +
  # Larvae index on primary axis
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "coral", alpha = 0.3) +
  geom_line(aes(y = larvae_index), color = "coral", linewidth = 0.8) +
  scale_y_continuous(
    name = "Estimated Larval Abundance Index; Delta-Lognormal AR1 (encounter)",
    sec.axis = sec_axis(~ (. - intercept_rd_dln3) / slope_rd_dln3,
                        name = "Recruitment Deviation (log-scale)")
  ) +
  scale_x_continuous(breaks = seq(1984, 2015, by = 4)) +
  labs(title = "SoCal Cabezon: Larval Index vs. Recruitment Deviations (1984–2015), Delta-Lognormal",
       x = "Year") +
  theme_bw()


# ---- Larval Index vs. Age-0 Recruits ----
# Age-0 is strictly positive like SSB, so the multiplicative scale_factor approach works.

scale_factor_age0_dln3 <- max(ss_dln3_combined_obs$larvae_index, na.rm = TRUE) /
  max(ss_dln3_combined_obs$age0, na.rm = TRUE)

ggplot(ss_dln3_combined_obs %>% filter(year >= 1984, year <= 2015),
       aes(x = year)) +
  # Age-0 scaled UP to primary axis units
  geom_ribbon(aes(ymin = age0_lo * scale_factor_age0_dln3,
                  ymax = age0_hi * scale_factor_age0_dln3),
              fill = "purple", alpha = 0.25) +
  geom_line(aes(y = age0 * scale_factor_age0_dln3),
            color = "purple", linewidth = 0.8) +
  # Larvae index on primary axis
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "coral", alpha = 0.3) +
  geom_line(aes(y = larvae_index), color = "coral", linewidth = 0.8) +
  scale_y_continuous(
    name = "Estimated Larval Abundance Index; Delta-Lognormal AR1 (encounter)",
    sec.axis = sec_axis(~ . / scale_factor_age0_dln3,
                        name = "Age-0 Recruits")
  ) +
  scale_x_continuous(breaks = seq(1984, 2015, by = 4)) +
  labs(title = "SoCal Cabezon: Larval Index vs. Age-0 Recruits (1984–2015), Delta-Lognormal",
       x = "Year") +
  theme_bw()
```

```{r 5.1.2. ss_cab_fit1 STAR comparisons station extrapolation, cache = T}
#---- 5.1. ss_cab_fit_dln3 STAR comparisons and CCF
# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 4.1.f. abundance index, assuming it's more stable.
ss_dln3_combined <- ss_dln3_cab_index_obs %>%
  dplyr::select(year, est, lwr, upr) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )
# Filter to observed years
ss_dln3_combined_obs <- ss_dln3_combined %>%
  filter(!is.na(larvae_index))
# CCF: Restricted larval index vs. age-0 recruits
ccf_boot(
  x = ss_dln3_combined_obs$larvae_index,
  y = ss_dln3_combined_obs$age0,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Age-0 Recruits (Delta-Lognormal)")
# CCF: Restricted larval index vs. recruitment deviations
ccf_boot(
  x = ss_dln3_combined_obs$larvae_index,
  y = ss_dln3_combined_obs$rec_dev,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Recruitment Deviations (Delta-Lognormal)")
# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.
# CCF: Restricted larval index vs. SSB
ccf_boot(
  x = ss_dln3_combined_obs$larvae_index,
  y = ss_dln3_combined_obs$ssb,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. SSB (Delta-Lognormal)")
# First-differenced
ccf_boot(
  x = diff(ss_dln3_combined_obs$larvae_index),
  y = diff(ss_dln3_combined_obs$ssb),
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Δ Larval Index vs. Δ SSB (Delta-Lognormal)")
# And only differenced SSB on raw larvae
ccf_boot(
  x = ss_dln3_combined_obs$larvae_index,
  y = diff(ss_dln3_combined_obs$ssb),
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Larval Index vs. Δ SSB (Delta-Lognormal)")
# Join differenced SSB back to index
ssb_diff_df_dln3 <- ss_dln3_combined_obs %>%
  arrange(year) %>%
  mutate(ssb_diff = c(NA, diff(ssb))) %>%
  filter(!is.na(ssb_diff))
lm_lag0_dln3 <- lm(ssb_diff ~ larvae_index, data = ssb_diff_df_dln3)
summary(lm_lag0_dln3)
# Visual
ssb_diff_df_dln3 %>%
  ggplot(aes(x = larvae_index, y = ssb_diff, label = year)) +
  geom_point() +
  geom_text(nudge_y = 50, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index", y = "ΔSSB (mt)",
       title = "Contemporaneous larval index vs. change in SSB (Delta-Lognormal)") +
  theme_minimal()
# Assuming ecologically that SSB would have the strongest relationship to larval abundance, we'll plot those two time series visually for a qualitative look
scale_factor_ssb_dln3 <- max(ss_dln3_combined_obs$larvae_index, na.rm = TRUE) /
  max(ss_dln3_combined_obs$ssb, na.rm = TRUE)
ggplot(ss_dln3_combined_obs %>% filter(year >= 1984, year <= 2015),
       aes(x = year)) +
  # SSB scaled UP to primary axis units
  geom_ribbon(aes(ymin = ssb_lo * scale_factor_ssb_dln3,
                  ymax = ssb_hi * scale_factor_ssb_dln3),
              fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = ssb * scale_factor_ssb_dln3), color = "steelblue", linewidth = 0.8) +
  # Larvae index on primary axis
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "coral", alpha = 0.3) +
  geom_line(aes(y = larvae_index), color = "coral", linewidth = 0.8) +
  scale_y_continuous(
    name = "Estimated Larval Abundance Index; Delta-Lognormal AR1 (encounter)",
    sec.axis = sec_axis(~ . / scale_factor_ssb_dln3,
                        name = "SSB (mt)")
  ) +
  scale_x_continuous(breaks = seq(1984, 2015, by = 4)) +
  labs(title = "SoCal Cabezon: Larval Index vs. SSB (1984–2015), Delta-Lognormal",
       x = "Year") +
  theme_bw()
```

## 5.2. ss_cab_fit01 — STAR Comparisons & CCFs

```{r 5.2. ss_cab_fit01 STAR comparisons full extrapolation, cache = T}
#---- 5.2. ss_cab_fit_dln4 STAR comparisons and CCF
# Create one data frame to house all STAR + larval index data
# Note! For these comparisons, we move forward with the constrained 4.2.f. abundance index, assuming it's more stable.
ss_dln_fy_combined <- ss_dln_fy_cab_index %>%
  dplyr::select(year, est, lwr, upr) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
ss_dln_fy_combined_obs <- ss_dln_fy_combined %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf_boot(
  x = ss_dln_fy_combined_obs$larvae_index,
  y = ss_dln_fy_combined_obs$age0,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Age-0 Recruits (Delta-Lognormal Fixed Year)")

# CCF: Restricted larval index vs. recruitment deviations
ccf_boot(
  x = ss_dln_fy_combined_obs$larvae_index,
  y = ss_dln_fy_combined_obs$rec_dev,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Recruitment Deviations (Delta-Lognormal Fixed Year)")

# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.

# CCF: Restricted larval index vs. SSB
ccf_boot(
  x = ss_dln_fy_combined_obs$larvae_index,
  y = ss_dln_fy_combined_obs$ssb,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. SSB (Delta-Lognormal Fixed Year)")

# First-differenced
ccf_boot(
  x = diff(ss_dln_fy_combined_obs$larvae_index),
  y = diff(ss_dln_fy_combined_obs$ssb),
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Δ Larval Index vs. Δ SSB (Delta-Lognormal Fixed Year)")

# Assuming ecologically that SSB would have the strongest relationship to larval abundance, we'll plot those two time series visually for a qualitative look
dln_fy_scale_factor_ssb <- max(ss_dln_fy_combined_obs$larvae_index, na.rm = TRUE) /
  max(ss_dln_fy_combined_obs$ssb, na.rm = TRUE)

ggplot(ss_dln_fy_combined_obs %>% filter(year >= 1984, year <= 2015),
       aes(x = year)) +
  # SSB scaled UP to primary axis units
  geom_ribbon(aes(ymin = ssb_lo * dln_fy_scale_factor_ssb,
                  ymax = ssb_hi * dln_fy_scale_factor_ssb),
              fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = ssb * dln_fy_scale_factor_ssb), color = "steelblue", linewidth = 0.8) +
  # Larvae index on primary axis
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "coral", alpha = 0.3) +
  geom_line(aes(y = larvae_index), color = "coral", linewidth = 0.8) +
  scale_y_continuous(
    name = "Estimated Larval Abundance Index; Delta-Lognormal Fixed Year",
    sec.axis = sec_axis(~ . / dln_fy_scale_factor_ssb,
                        name = "SSB (mt)")
  ) +
  scale_x_continuous(breaks = seq(1984, 2015, by = 4)) +
  labs(title = "SoCal Cabezon: Larval Index vs. SSB (1984–2015), Delta-Lognormal Fixed Year",
       x = "Year") +
  theme_bw()
```




```{r 5.2.2. ss_cab_fit01 STAR comparisons station extrapolation, cache = T}
ss_dln_fy_combined <- ss_dln_fy_cab_index_obs %>%
  dplyr::select(year, est, lwr, upr) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
ss_dln_fy_combined_obs <- ss_dln_fy_combined %>%
  filter(!is.na(larvae_index))

# CCF: Restricted larval index vs. age-0 recruits
ccf_boot(
  x = ss_dln_fy_combined_obs$larvae_index,
  y = ss_dln_fy_combined_obs$age0,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Age-0 Recruits (Delta-Lognormal Fixed Year)")

# CCF: Restricted larval index vs. recruitment deviations
ccf_boot(
  x = ss_dln_fy_combined_obs$larvae_index,
  y = ss_dln_fy_combined_obs$rec_dev,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Recruitment Deviations (Delta-Lognormal Fixed Year)")

# Nothing really :( when we try against SSB, we get some strong signal, but this is probably due to a shared declining trend. We'll try to parse that.

# CCF: Restricted larval index vs. SSB
ccf_boot(
  x = ss_dln_fy_combined_obs$larvae_index,
  y = ss_dln_fy_combined_obs$ssb,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. SSB (Delta-Lognormal Fixed Year)")

# First-differenced
ccf_boot(
  x = diff(ss_dln_fy_combined_obs$larvae_index),
  y = diff(ss_dln_fy_combined_obs$ssb),
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Δ Larval Index vs. Δ SSB (Delta-Lognormal Fixed Year)")

# Assuming ecologically that SSB would have the strongest relationship to larval abundance, we'll plot those two time series visually for a qualitative look
dln_fy_scale_factor_ssb <- max(ss_dln_fy_combined_obs$larvae_index, na.rm = TRUE) /
  max(ss_dln_fy_combined_obs$ssb, na.rm = TRUE)

ggplot(ss_dln_fy_combined_obs %>% filter(year >= 1984, year <= 2015),
       aes(x = year)) +
  # SSB scaled UP to primary axis units
  geom_ribbon(aes(ymin = ssb_lo * dln_fy_scale_factor_ssb,
                  ymax = ssb_hi * dln_fy_scale_factor_ssb),
              fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = ssb * dln_fy_scale_factor_ssb), color = "steelblue", linewidth = 0.8) +
  # Larvae index on primary axis
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "coral", alpha = 0.3) +
  geom_line(aes(y = larvae_index), color = "coral", linewidth = 0.8) +
  scale_y_continuous(
    name = "Estimated Larval Abundance Index; Delta-Lognormal Fixed Year",
    sec.axis = sec_axis(~ . / dln_fy_scale_factor_ssb,
                        name = "SSB (mt)")
  ) +
  scale_x_continuous(breaks = seq(1984, 2015, by = 4)) +
  labs(title = "SoCal Cabezon: Larval Index vs. SSB (1984–2015), Delta-Lognormal Fixed Year",
       x = "Year") +
  theme_bw()
```

# 6. REGRESSION ANALYSES (CONSTRAINED)

## 6.1. ss_cab_fit1 — Regression Analysis

```{r 6.1. ss_cab_fit1 correlation testing, cache = T}
# RQ residuals for delta-lognormal AR1 (encounter) spatiotemporal model
ss_dln3_pred_obs %>%
  group_by(year) %>%
  mutate(med_resid = median(rq_resid)) %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method  = "loess", span = 0.4,
    se      = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Year",
    y     = "RQ Residual",
    title = "Annual RQ Residual Distributions — ss_cab_fit_dln3"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

ss_dln3_pred_obs %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x     = "Year",
    y     = "Cumulative Sum of Median RQ Residual",
    title = "CUSUM — ss_cab_fit_dln3"
  ) +
  theme_minimal()

# Is the underprediction concentrated in one delta component?
# Separate RQ residuals by model component
set.seed(my.seed)
ss_dln3_pred_obs <- ss_dln3_pred_obs %>%
  mutate(
    rq_resid1 = residuals(ss_cab_fit_dln3, model = 1, type = "mle-mvn"),
    rq_resid2 = residuals(ss_cab_fit_dln3, model = 2, type = "mle-mvn")
  )

# CUSUM for each component separately
ss_dln3_pred_obs %>%
  group_by(year) %>%
  summarise(
    med_resid1 = median(rq_resid1, na.rm = TRUE),
    med_resid2 = median(rq_resid2, na.rm = TRUE),  # positive tows only
    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    cusum1 = cumsum(replace_na(med_resid1, 0)),
    cusum2 = cumsum(replace_na(med_resid2, 0))
  ) %>%
  pivot_longer(cols = c(cusum1, cusum2),
               names_to = "component",
               values_to = "cusum") %>%
  ggplot(aes(x = year, y = cusum, color = component)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("cusum1" = "steelblue", "cusum2" = "firebrick"),
                     labels = c("Delta (binomial)", "Lognormal")) +
  labs(x = "Year", y = "CUSUM",
       title = "CUSUM by Delta Model Component — ss_cab_fit_dln3") +
  theme_minimal()

# Is the underprediction spatially structured?
ss_dln3_pred_obs %>%
  group_by(station) %>%
  summarise(mean_resid = mean(rq_resid), .groups = "drop") %>%
  ggplot(aes(x = station, y = mean_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = TRUE) +
  labs(x = "Station", y = "Mean RQ Residual",
       title = "Mean Residual by Station — ss_cab_fit_dln3") +
  theme_minimal()


#---- Simple lagged regression at -1 to check

ss_dln3_combined_lagged <- ss_dln3_combined_obs %>%
  arrange(year) %>%
  left_join(
    ss_dln3_combined_obs %>%
      dplyr::select(year, age0_lead1 = age0, rec_dev_lead1 = rec_dev) %>%
      mutate(year = year - 1),  # shift target year back by 1
    by = "year"
  )

ss_dln3_lm_age0   <- lm(age0_lead1    ~ larvae_index, data = ss_dln3_combined_lagged)
ss_dln3_lm_recdev <- lm(rec_dev_lead1 ~ larvae_index, data = ss_dln3_combined_lagged)

summary(ss_dln3_lm_age0)
summary(ss_dln3_lm_recdev)

ss_dln3_combined_lagged %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)",
       title = "Lagged Larval Index vs. Age-0 Recruits — ss_cab_fit_dln3") +
  theme_minimal()

ss_dln3_combined_lagged_no83 <- ss_dln3_combined_lagged %>% filter(year != 1983)

ss_dln3_lm_age0_no83 <- lm(age0_lead1 ~ larvae_index, data = ss_dln3_combined_lagged_no83)
summary(ss_dln3_lm_age0_no83)

ss_dln3_lm_age0_detrend   <- lm(age0_lead1    ~ larvae_index + year, data = ss_dln3_combined_lagged)
ss_dln3_lm_recdev_detrend <- lm(rec_dev_lead1 ~ larvae_index + year, data = ss_dln3_combined_lagged)

summary(ss_dln3_lm_age0_detrend)
summary(ss_dln3_lm_recdev_detrend)

plot(ss_dln3_lm_age0, which = 4)  # Cook's distance plot

# Remove the three most anomalous years
ss_dln3_combined_lagged_trimmed <- ss_dln3_combined_lagged %>%
  filter(!year %in% c(1983, 1986, 1998))

ss_dln3_combined_lagged_trimmed %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)",
       title = "Lagged Larval Index vs. Age-0 Recruits (trimmed) — ss_cab_fit_dln3") +
  theme_minimal()

ss_dln3_lm_age0_trimmed <- lm(age0_lead1 ~ larvae_index, data = ss_dln3_combined_lagged_trimmed)
summary(ss_dln3_lm_age0_trimmed)

ss_dln3_lm_age0_trimmed_detrend <- lm(age0_lead1 ~ larvae_index + year, data = ss_dln3_combined_lagged_trimmed)
summary(ss_dln3_lm_age0_trimmed_detrend)

```

## 6.2. ss_cab_fit01 — Regression Analysis

```{r 6.2. ss_cab_fit01 correlation testing, cache = T}
# RQ residuals for delta-lognormal fixed year effect model
ss_dln_fy_pred_obs %>%
  group_by(year) %>%
  mutate(med_resid = median(rq_resid)) %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method  = "loess", span = 0.4,
    se      = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Year",
    y     = "RQ Residual",
    title = "Annual RQ Residual Distributions — ss_cab_fit_dln4"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

ss_dln_fy_pred_obs %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x     = "Year",
    y     = "Cumulative Sum of Median RQ Residual",
    title = "CUSUM — ss_cab_fit_dln4"
  ) +
  theme_minimal()

# 1. Is the underprediction concentrated in one delta component?
set.seed(my.seed)
ss_dln_fy_pred_obs <- ss_dln_fy_pred_obs %>%
  mutate(
    rq_resid1 = residuals(ss_cab_fit_dln4, model = 1, type = "mle-mvn"),
    rq_resid2 = residuals(ss_cab_fit_dln4, model = 2, type = "mle-mvn")
  )

ss_dln_fy_pred_obs %>%
  group_by(year) %>%
  summarise(
    med_resid1 = median(rq_resid1, na.rm = TRUE),
    med_resid2 = median(rq_resid2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    cusum1 = cumsum(replace_na(med_resid1, 0)),
    cusum2 = cumsum(replace_na(med_resid2, 0))
  ) %>%
  pivot_longer(cols = c(cusum1, cusum2),
               names_to  = "component",
               values_to = "cusum") %>%
  ggplot(aes(x = year, y = cusum, color = component)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("cusum1" = "steelblue", "cusum2" = "firebrick"),
                     labels = c("cusum1" = "Delta (binomial)", "cusum2" = "Lognormal")) +
  labs(x = "Year", y = "CUSUM",
       title = "CUSUM by Delta Model Component — ss_cab_fit_dln4") +
  theme_minimal()

# 2. Is the underprediction spatially structured?
ss_dln_fy_pred_obs %>%
  group_by(station) %>%
  summarise(mean_resid = mean(rq_resid), .groups = "drop") %>%
  ggplot(aes(x = station, y = mean_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = TRUE) +
  labs(x = "Station", y = "Mean RQ Residual",
       title = "Mean Residual by Station — ss_cab_fit_dln4") +
  theme_minimal()

#---- Simple lagged regression at lag -1

ss_dln_fy_combined_lagged <- ss_dln_fy_combined_obs %>%
  arrange(year) %>%
  left_join(
    ss_dln_fy_combined_obs %>%
      dplyr::select(year, age0_lead1 = age0, rec_dev_lead1 = rec_dev) %>%
      mutate(year = year -1),
    by = "year"
  )

ss_dln_fy_lm_age0   <- lm(age0_lead1    ~ larvae_index, data = ss_dln_fy_combined_lagged)
ss_dln_fy_lm_recdev <- lm(rec_dev_lead1 ~ larvae_index, data = ss_dln_fy_combined_lagged)

summary(ss_dln_fy_lm_age0)
summary(ss_dln_fy_lm_recdev)

ss_dln_fy_combined_lagged %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)",
       title = "Lagged Larval Index vs. Age-0 Recruits — ss_cab_fit_dln4") +
  theme_minimal()

# Sensitivity: remove 1983
ss_dln_fy_combined_lagged_no83 <- ss_dln_fy_combined_lagged %>% filter(year != 1983)
ss_dln_fy_lm_age0_no83 <- lm(age0_lead1 ~ larvae_index, data = ss_dln_fy_combined_lagged_no83)
summary(ss_dln_fy_lm_age0_no83)

# Detrended regressions
ss_dln_fy_lm_age0_detrend   <- lm(age0_lead1    ~ larvae_index + year, data = ss_dln_fy_combined_lagged)
ss_dln_fy_lm_recdev_detrend <- lm(rec_dev_lead1 ~ larvae_index + year, data = ss_dln_fy_combined_lagged)
summary(ss_dln_fy_lm_age0_detrend)
summary(ss_dln_fy_lm_recdev_detrend)

# Cook's distance
plot(ss_dln_fy_lm_age0, which = 4)

# Trimmed analysis: remove three most anomalous years
ss_dln_fy_combined_lagged_trimmed <- ss_dln_fy_combined_lagged %>%
  filter(!year %in% c(1983, 1986, 1998))

ss_dln_fy_combined_lagged_trimmed %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)",
       title = "Lagged Larval Index vs. Age-0 Recruits (trimmed) — ss_cab_fit_dln4") +
  theme_minimal()

ss_dln_fy_lm_age0_trimmed <- lm(age0_lead1 ~ larvae_index, data = ss_dln_fy_combined_lagged_trimmed)
summary(ss_dln_fy_lm_age0_trimmed)

ss_dln_fy_lm_age0_trimmed_detrend <- lm(age0_lead1 ~ larvae_index + year, data = ss_dln_fy_combined_lagged_trimmed)
summary(ss_dln_fy_lm_age0_trimmed_detrend)

```


# Notes

**Check** if CDFW or another source has data that could proxy better than STAR for SSB. For example, taking Erica's approach, is there useful landings data for cabezon?

**Take** annual values for our significant covariates (temp and O2) and run them against your indices. Is there any visual relationship? Qualitative similarity between the time-series?

# Heatmaps for winning models

```{r}
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

utm_crs <- 32611

ss_map_pred <- predict(
  ss_cab_fit_dln3,
  newdata = ss_dln3_pred_grid,
  type = "response"
)

pred_plot <- ss_map_pred |>
  mutate(
    expected = est1 * est2,
    X_m = X * 1000,
    Y_m = Y * 1000
  )

ca_coast <- ne_states(
  country = "United States of America",
  returnclass = "sf"
) |>
  dplyr::filter(name == "California") |>
  st_transform(utm_crs)

calcofi_stations <- cabezon_shelf_spawn |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000
  ) |>
  distinct(line, station, X_m, Y_m) |>
  st_as_sf(coords = c("X_m", "Y_m"), crs = utm_crs)

calcofi_lines <- calcofi_stations |>
  arrange(line, station) |>
  group_by(line) |>
  summarize(do_union = FALSE, .groups = "drop") |>
  st_cast("LINESTRING")

pred_plot <- ss_map_pred |>
  mutate(X_m = X * 1000, Y_m = Y * 1000) |>
  summarise(
    est = mean(est),
    est1 = mean(est1),
    est2 = mean(est2),
    .by = c(X_m, Y_m))


# Plot showing larval observations by station over time
obs_plot <- cabezon_shelf_spawn |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000,
    larvae_positive = larvae_100m3 > 0
  )

x_range <- range(pred_plot$X_m, na.rm = TRUE)
y_range <- range(pred_plot$Y_m, na.rm = TRUE)

x_buffer <- diff(x_range) * 0.05
y_buffer <- diff(y_range) * 0.05

xlim_padded <- c(x_range[1] - x_buffer, x_range[2] + x_buffer)
ylim_padded <- c(y_range[1] - y_buffer, y_range[2] + y_buffer)

ggplot() +
  geom_sf(
    data = ca_coast,
    inherit.aes = FALSE,
    fill = "grey80",
    color = "grey40"
  ) +
  geom_sf(
    data = calcofi_lines,
    inherit.aes = FALSE,
    color = "grey95",
    linewidth = 0.6
  ) +
  geom_point(
    data = obs_plot |> filter(!larvae_positive),
    aes(x = X_m, y = Y_m),
    size = 0.4,
    alpha = 0.25
  ) +
  geom_point(
    data = obs_plot |> filter(larvae_positive),
    aes(x = X_m, y = Y_m, size = larvae_100m3),
    alpha = 0.7
  ) +
  coord_sf(
  crs = st_crs(utm_crs),
  xlim = xlim_padded,
  ylim = ylim_padded,
  expand = FALSE
  ) +
  facet_wrap(~ year) +
  scale_size_continuous(
    name = "Observed larvae / 100 m³",
    range = c(1, 7)
  ) +
  labs(
    title = "Observed cabezon larval abundance by CalCOFI station",
    x = NULL,
    y = NULL
  ) +
  theme_minimal()

# Plot showing only delta (distribution)
ggplot() +
  geom_raster(
    data = pred_plot,
    aes(x = X_m, y = Y_m, fill = log1p(est))
  ) +
  geom_sf(
    data = ca_coast,
    inherit.aes = FALSE,
    fill = "beige",
    color = "black",
    linewidth = 0.2
  ) +
#  geom_sf(
#    data = calcofi_lines,
#    inherit.aes = FALSE,
#    color = "white",
#    linewidth = 1.2,
#    alpha = 0.35
#  ) +
#  geom_point(
#    data = obs_aggregate %>% filter(ever_positive),
#    aes(x = X_m, y = Y_m),
#    inherit.aes = FALSE,
#    size = 2.5,
#    color = "goldenrod",
#    alpha = 1
#  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(pred_plot$X_m, na.rm = TRUE),
    ylim = range(pred_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  scale_fill_viridis_c(
  option = "mako",
  name = "Predicted larval abundance"
  ) +
  labs(
    title = "Persistent spatial structure in modeled larval abundance",
    subtitle = "Log-transformed expected abundance from delta-lognormal index model",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.border = element_rect(color = "grey65", fill = NA, linewidth = 0.6),
    plot.title = element_text(face = "bold", size = 17),
    plot.subtitle = element_text(size = 12, color = "grey35"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    panel.grid = element_line(color = "grey90", linewidth = 0.2)
  )

# Plot showing hurdle (abundance); same across years
ggplot() +
  geom_raster(
    data = pred_plot,
    aes(x = X_m, y = Y_m, fill = est2)
  ) +
  geom_sf(data = ca_coast, inherit.aes = FALSE, fill = "grey80") +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(pred_plot$X_m, na.rm = TRUE),
    ylim = range(pred_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  scale_fill_viridis_c(option = "mako",
                       name = "Predicted larvae") +
  theme_minimal()

pred_plot_yearly <- ss_map_pred |>
  mutate(X_m = X * 1000, Y_m = Y * 1000)

ggplot() +
  geom_raster(
    data = pred_plot_yearly,
    aes(x = X_m, y = Y_m, fill = log(est))
  ) +
  geom_sf(data = ca_coast, inherit.aes = FALSE, fill = "grey80") +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(pred_plot$X_m, na.rm = TRUE),
    ylim = range(pred_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  facet_wrap(~ year) +
  scale_fill_viridis_c(name = "Predicted larvae ln(abun)") +
  theme_minimal()

```

# Figures

```{r CCF publication figures}
library(dplyr)
library(ggplot2)
library(funtimes)

# ── 1. EXTRACT CCF RESULTS ──────────────────────────────────────────────
# Compute LOESS residuals first
loess_df <- ss_dln3_combined_obs %>%
  arrange(year) %>%
  mutate(
    loess_index = predict(loess(larvae_index ~ year, span = 0.5)),
    loess_ssb   = predict(loess(ssb ~ year, span = 0.5)),
    resid_index = larvae_index - loess_index,
    resid_ssb   = ssb - loess_ssb
  )

pdf(NULL)
ccf_age0   <- ccf_boot(ss_dln3_combined_obs$larvae_index,
                       ss_dln3_combined_obs$age0,
                       lag.max = 10, plot = "Spearman", B = 1000)
ccf_recdev <- ccf_boot(ss_dln3_combined_obs$larvae_index,
                       ss_dln3_combined_obs$rec_dev,
                       lag.max = 10, plot = "Spearman", B = 1000)
ccf_ssb    <- ccf_boot(ss_dln3_combined_obs$larvae_index,
                       ss_dln3_combined_obs$ssb,
                       lag.max = 10, plot = "Spearman", B = 1000)
ccf_dssb   <- ccf_boot(diff(ss_dln3_combined_obs$larvae_index),
                       diff(ss_dln3_combined_obs$ssb),
                       lag.max = 10, plot = "Spearman", B = 1000)


# Add: LOESS-detrended instead of first-differenced
ccf_dssl   <- ccf_boot(loess_df$resid_index,
                       loess_df$resid_ssb,
                       lag.max = 10, plot = "Spearman", B = 1000)
dev.off()

# ── 2. ASSEMBLE TIDY DATA FRAME ─────────────────────────────────────────
tidy_ccf <- function(df, panel_label) {
  df |>
    transmute(
      lag    = Lag,
      r      = r_S,
      lower  = lower_S,
      upper  = upper_S,
      sig    = r < lower | r > upper,
      panel  = panel_label
    )
}

ccf_df <- bind_rows(
  tidy_ccf(ccf_age0,   "Age-0 recruits"),
  tidy_ccf(ccf_recdev, "Recruitment deviations"),
  tidy_ccf(ccf_ssb,    "SSB (raw)"),
  tidy_ccf(ccf_dssl,   "SSB (LOESS-detrended)")
)

ccf_df$panel <- factor(
  ccf_df$panel,
  levels = c("Age-0 recruits", "Recruitment deviations",
             "SSB (raw)", "SSB (LOESS-detrended)")
)

# ── 3. PUBLICATION FIGURE ───────────────────────────────────────────────
p_ccf <- ggplot(ccf_df, aes(x = lag, y = r)) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "skyblue", alpha = 0.6) +
  geom_segment(aes(xend = lag, yend = 0), linewidth = 0.5,
               colour = "grey30") +
  geom_point(aes(fill = sig), shape = 21, size = 2.2,
             colour = "grey20", stroke = 0.4) +
  scale_fill_manual(values = c(`FALSE` = "white", `TRUE` = "black"),
                    labels = c("ns", "outside 95% CI"),
                    name = NULL) +
  facet_wrap(~ panel, ncol = 2) +
  scale_x_continuous(breaks = seq(-10, 10, 2)) +
  labs(
    x        = "Lag (years)",
    y        = "Spearman cross-correlation",
    title    = "Cross-correlations: restricted larval index vs. stock assessment outputs",
    subtitle = "Delta-lognormal IID model; blue band = bootstrap 95% CI (B = 1000)"
  ) +
  theme_bw(base_size = 11) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.background   = element_rect(fill = "grey95", colour = NA),
    strip.text         = element_text(face = "bold", size = 10),
    plot.title         = element_text(face = "bold", size = 12),
    plot.subtitle      = element_text(colour = "grey30", size = 9.5),
    legend.position    = "bottom",
    legend.margin      = margin(t = -5)
  )

p_ccf


ggsave("figures/ccf_panel.png", p_ccf,
       width = 8, height = 6.5, dpi = 300)
```

```{r CCF alternative figure}
library(patchwork)

# --- Compute LOESS smooths and residuals ---
loess_df <- ss_dln3_combined_obs %>%
  filter(year >= 1984, year <= 2015) %>%
  arrange(year) %>%
  mutate(
    loess_index = predict(loess(larvae_index ~ year, span = 0.5)),
    loess_ssb   = predict(loess(ssb ~ year, span = 0.5)),
    resid_index = larvae_index - loess_index,
    resid_ssb   = ssb - loess_ssb
  )

# Scale factor for dual axis
scale_factor <- max(loess_df$larvae_index, na.rm = TRUE) /
                max(loess_df$ssb, na.rm = TRUE)

# Rescale SSB residuals to index space for visual comparison
resid_scale  <- sd(loess_df$resid_index, na.rm = TRUE) /
                sd(loess_df$resid_ssb,   na.rm = TRUE)

# --- Panel A: Raw series + LOESS smooths ---
p_raw <- ggplot(loess_df, aes(x = year)) +
  # SSB ribbon + line
  geom_ribbon(aes(ymin = ssb_lo * scale_factor,
                  ymax = ssb_hi * scale_factor),
              fill = "steelblue", alpha = 0.2) +
  geom_line(aes(y = ssb * scale_factor),
            color = "steelblue", linewidth = 0.8, alpha = 0.5) +
  # SSB LOESS smooth
  geom_line(aes(y = loess_ssb * scale_factor),
            color = "steelblue", linewidth = 1.3) +
  # Larval index ribbon + line
  geom_ribbon(aes(ymin = lwr, ymax = upr),
              fill = "coral", alpha = 0.2) +
  geom_line(aes(y = larvae_index),
            color = "coral", linewidth = 0.8, alpha = 0.5) +
  # Index LOESS smooth
  geom_line(aes(y = loess_index),
            color = "coral", linewidth = 1.3) +
  scale_y_continuous(
    name = "Larval abundance index",
    sec.axis = sec_axis(~ . / scale_factor, name = "SSB (mt)")
  ) +
  scale_x_continuous(breaks = seq(1984, 2015, by = 4)) +
  labs(title = "A. Raw series with LOESS smooths (span = 0.5)",
       x = NULL) +
  theme_bw() +
  theme(axis.title.y.left  = element_text(color = "coral"),
        axis.title.y.right = element_text(color = "steelblue"))

# --- Panel B: Detrended residuals ---
p_resid <- ggplot(loess_df, aes(x = year)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(aes(y = resid_index),
            color = "coral", linewidth = 0.8) +
  geom_point(aes(y = resid_index),
             color = "coral", size = 1.5) +
  geom_line(aes(y = resid_ssb * resid_scale),
            color = "steelblue", linewidth = 0.8) +
  geom_point(aes(y = resid_ssb * resid_scale),
             color = "steelblue", size = 1.5) +
  scale_y_continuous(
    name = "Larval index residual",
    sec.axis = sec_axis(~ . / resid_scale, name = "SSB residual (mt)")
  ) +
  scale_x_continuous(breaks = seq(1984, 2015, by = 4)) +
  labs(title = "B. Detrended residuals (raw \u2212 LOESS smooth)",
       x = "Year") +
  theme_bw() +
  theme(axis.title.y.left  = element_text(color = "coral"),
        axis.title.y.right = element_text(color = "steelblue"))

# --- Combine ---
p_raw / p_resid

# --- CCF on LOESS-detrended residuals ---

# Compute residuals (reusing loess_df from the plot code)
loess_df <- ss_dln3_combined_obs %>%
  filter(year >= 1984, year <= 2015) %>%
  arrange(year) %>%
  mutate(
    loess_index = predict(loess(larvae_index ~ year, span = 0.5)),
    loess_ssb   = predict(loess(ssb ~ year, span = 0.5)),
    resid_index = larvae_index - loess_index,
    resid_ssb   = ssb - loess_ssb
  )

# CCF with bootstrap CIs on detrended residuals
ccf_boot(
  x       = loess_df$resid_index,
  y       = loess_df$resid_ssb,
  lag.max = 10,
  plot    = "Spearman",
  B       = 1000
); title(main = "CCF: Detrended Larval Index vs. Detrended SSB (LOESS residuals, span = 0.5)")


# --- Windowed CCF on LOESS residuals (1997-2012) ---

loess_window <- ss_dln3_combined_obs %>%
  filter(year >= 1999, year <= 2009) %>%
  arrange(year) %>%
  mutate(
    loess_index = predict(loess(larvae_index ~ year, span = 0.5)),
    loess_ssb   = predict(loess(ssb ~ year, span = 0.5)),
    resid_index = larvae_index - loess_index,
    resid_ssb   = ssb - loess_ssb
  )

# With only ~15 obs, pull lag.max in to 5 to avoid
# burning too many degrees of freedom at the outer lags
ccf_boot(
  x       = loess_window$resid_index,
  y       = loess_window$resid_ssb,
  lag.max = 5,
  plot    = "Spearman",
  B       = 1000
); title(main = "CCF: Detrended Larval Index vs. Detrended SSB\n1997-2012 window (LOESS residuals, span = 0.5)")
```


```{r sampling effort figures}
#ss_plot_data_year
#ss_plot_data_ms 

library(scales)

month_data <- ss_plot_data_ms %>%
  filter(group_var == "Month") %>%
  mutate(
    level = recode(as.character(level),
      "1" = "January", "2" = "February", "3" = "March"),
    level = factor(level, levels = c("January", "February", "March"))
  )

p_month <- ggplot(month_data, aes(x = level, y = count, fill = tow_type)) +
  geom_col(width = 0.6, color = "white", linewidth = 0.3) +
  shared_fill +
  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Cabezon Larvae: Positive vs. Total Tows",
    subtitle = "Sampling effort by month",
    y = "Number of Tows"
  ) +
  shared_bar_theme +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

# Refined fill palette
shared_fill <- scale_fill_manual(
  values = c("negative_tows" = "grey88", "positive_tows" = "#2C5F8D"),
  labels = c("negative_tows" = "Larvae Absent", "positive_tows" = "Larvae Present"),
  name = NULL
)

# Refined shared theme
shared_bar_theme <- theme_minimal(base_size = 12) +
  theme(
    strip.text         = element_text(face = "bold", size = 13, margin = margin(b = 8)),
    axis.text.x        = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y        = element_text(size = 10),
    axis.title.y       = element_text(size = 11, margin = margin(r = 10)),
    axis.title.x       = element_blank(),
    legend.position    = "bottom",
    legend.text        = element_text(size = 11),
    legend.key.size    = unit(0.5, "cm"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey92", linewidth = 0.4),
    plot.title         = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle      = element_text(color = "grey40", size = 10, margin = margin(b = 12)),
    plot.margin        = margin(15, 15, 10, 15)
  )

# For the year plot
p_year <- ggplot(ss_plot_data_year, aes(x = year, y = count, fill = tow_type)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.2) +
  shared_fill +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Cabezon Larvae: Positive vs. Total Tows",
    subtitle = "Annual sampling effort, 1981\u20132015",
    y = "Number of Tows",
    caption = "Stacked bars show larvae-present tows (blue) against total sampling effort (grey).\nElevated effort in some years (1981, 1984, 1991, 1994, 2009, 2012) reflects expanded grid coverage or event-specific surveys."
  ) +
  shared_bar_theme +
  theme(
    plot.caption = element_text(
      hjust = 0, size = 9, color = "grey30",
      margin = margin(t = 10), lineheight = 1.1
    )
  )

p_month
p_year

#ggsave("figures/tows-by-month.png", p_month,
#       width = 9.85, height = 6.8, dpi = 300)

#ggsave("figures/tows-by-year.png", p_year,
#       width = 9.85, height = 6.8, dpi = 300)


obs_aggregate <- obs_plot %>%
  group_by(station_id) %>%
  summarise(
    X_m = mean(X_m, na.rm = TRUE),
    Y_m = mean(Y_m, na.rm = TRUE),
    n_visits = n(),
    n_positive = sum(larvae_100m3 > 0, na.rm = TRUE),

    mean_abundance = mean(larvae_100m3, na.rm = TRUE),
    mean_positive_abundance = ifelse(
      sum(larvae_100m3 > 0, na.rm = TRUE) > 0,
      mean(larvae_100m3[larvae_100m3 > 0], na.rm = TRUE),
      NA_real_
    ),
    sum_abundance = sum(larvae_100m3, na.rm = TRUE),
    max_abundance = max(larvae_100m3, na.rm = TRUE),

    ever_positive = any(larvae_100m3 > 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    plot_abundance = mean_positive_abundance
  )

# Compute padded plot limits

library(rnaturalearth)

# Get both US and Mexico landmass
na_coast <- ne_countries(
  scale = "medium",  # or "large" if you have rnaturalearthhires
  country = c("United States of America", "Mexico"),
  returnclass = "sf"
)

x_range <- range(obs_aggregate$X_m, na.rm = TRUE)
y_range <- range(obs_aggregate$Y_m, na.rm = TRUE)
x_buffer <- diff(x_range) * 0.10
y_buffer <- diff(y_range) * 0.10

xlim_padded <- c(x_range[1] - x_buffer, x_range[2] + x_buffer)
ylim_padded <- c(y_range[1] - y_buffer, y_range[2] + y_buffer)
library(marmap)
library(ggplot2)
library(sf)
library(dplyr)

# Fetch bathymetry (you've already done this; included for completeness)
bathy <- getNOAA.bathy(
  lon1 = -122, lon2 = -116.5,
  lat1 = 31, lat2 = 35.5,
  resolution = 1
)
bathy_df <- fortify.bathy(bathy)

# Color palette
ocean_color <- "#CFE3F2"
land_color  <- "#E8DEC3"
contour_color <- "grey50"

# Region label coordinates ----
# Define in lon/lat, then transform to your UTM CRS
labels_sf <- st_as_sf(
  data.frame(
    name = c("California", "Mexico"),
    lon = c(-118, -117),
    lat = c(34.8, 32.4)
  ),
  coords = c("lon", "lat"),
  crs = 4326
) |>
  st_transform(st_crs(utm_crs))

label_coords <- st_coordinates(labels_sf)
X_california_label <- label_coords[1, "X"]
Y_california_label <- label_coords[1, "Y"]
X_mexico_label     <- label_coords[2, "X"]
Y_mexico_label     <- label_coords[2, "Y"]

# Transform bathymetry contour to UTM ----
# Extract just the 200m contour line as sf, transform to UTM
bathy_sf <- bathy_df |>
  filter(z >= -1000 & z <= 0) |>  # trim to relevant depth range for speed
  st_as_sf(coords = c("x", "y"), crs = 4326)

# Simpler approach: draw the contour in lon/lat space and let coord_sf reproject
# We do this by converting bathy to a stars/terra raster, extracting the contour
# as an sf LINESTRING, then transforming.

library(terra)
bathy_rast <- marmap::as.raster(bathy)
bathy_terra <- terra::rast(bathy_rast)
contour_levels <- c(-200, -1000, -2000)
contour_lines <- terra::as.contour(bathy_terra, levels = contour_levels) |>
  sf::st_as_sf() |>
  sf::st_set_crs(4326) |>
  sf::st_transform(st_crs(utm_crs))

# Plot ----
p_effort_grid <- ggplot() +
  # 200m isobath
  geom_sf(
    data = contour_lines,
    aes(alpha = factor(level)),
    color = contour_color,
    linewidth = 0.3,
    linetype = "dashed",
    show.legend = FALSE
  ) +
  # Land
  geom_sf(
    data = na_coast,
    inherit.aes = FALSE,
    fill = land_color,
    color = "grey40",
    linewidth = 0.3
  ) +
  # CalCOFI grid lines
  geom_sf(
    data = calcofi_lines,
    inherit.aes = FALSE,
    color = "grey75",
    linewidth = 0.25
  ) +
  # Negative tows (small open circles for full sampling footprint)
  geom_point(
    data = obs_aggregate %>% filter(!ever_positive),
    aes(x = X_m, y = Y_m),
    size = 0.6,
    alpha = 0.5,
    color = "grey40",
    shape = 1
  ) +
  # Positive tows (sized filled bubbles)
  geom_point(
    data = obs_aggregate %>% filter(ever_positive),
    aes(x = X_m, y = Y_m, size = plot_abundance),
    alpha = 0.75,
    color = "#2C5F8D"
  ) +
  # Region labels
  annotate("text",
    x = X_california_label, y = Y_california_label,
    label = "California",
    color = "grey25", size = 4, fontface = "italic"
  ) +
  annotate("text",
    x = X_mexico_label, y = Y_mexico_label,
    label = "Mexico",
    color = "grey25", size = 4, fontface = "italic"
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = xlim_padded,
    ylim = ylim_padded,
    expand = FALSE
  ) +
  scale_size_continuous(
    name = "Mean larvae / 100 m\u00b3\n(positive tows)",
    range = c(1, 8)
  ) +
  labs(
    title = "Observed cabezon larval abundance by CalCOFI station",
    subtitle = "Mean density from positive tows, aggregated by station, 1981\u20132015",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.background = element_rect(fill = ocean_color, color = NA),
    panel.grid.major = element_line(color = "white", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "grey40", size = 10, margin = margin(b = 10)),
    legend.position = "right"
  )

ggsave(
  "C:/Users/machs/SynologyDrive/Dan's Files/SIO/0_capstone/6_figures/core-grid-study-region-effort.png",
  p_effort_grid,
  width = 16,
  height = 11,
  units = "in",
  dpi = 300
)

```

```{r preferred larval index figure}
library(scales)  # for comma formatting

p_index <- ggplot(ss_dln3_cab_index, aes(x = year, y = est)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), 
              fill = "#2C5F8D", alpha = 0.2) +
  geom_line(color = "#1A3E5C", linewidth = 0.7) +
  geom_point(color = "#1A3E5C", size = 1.5) +
  scale_x_continuous(breaks = seq(1980, 2015, by = 5)) +
  scale_y_continuous(labels = comma, 
                     limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Estimated cabezon larval abundance index, 1981–2015",
    subtitle = "DLn-IID: Delta-lognormal model, IID encounter spatiotemporal field",
    x = NULL,
    y = "Estimated abundance index",
    caption = "Shaded band: 95% CI."
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(color = "grey40", margin = margin(b = 15)),
    plot.caption = element_text(color = "grey50", size = 11, hjust = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "grey92"),
    axis.title.y = element_text(size = 16, margin = margin(r = 10)),
    axis.text = element_text(size = 13),
    plot.margin = margin(15, 20, 15, 15)
  ) 

ggsave(
  "C:/Users/machs/SynologyDrive/Dan's Files/SIO/0_capstone/6_figures/preferred_index.png",
  plot = p_index,
  width = 12.5,
  height = 5.5,
  units = "in",
  dpi = 300
)
```

```{r study region figure}
# =============================================================
# Cabezon abundance across CalCOFI core grid
# Adds US + Mexico land, removes coastal border artifacts,
# moves Mexico label, and enlarges CalCOFI station markers
# =============================================================

library(sf)
library(dplyr)
library(stringr)
library(ggplot2)
library(rnaturalearth)
library(tibble)
library(scales)

# ---- 0. Map settings ---------------------------------------------------

kml_path <- "CalCOFI_75StandardStations.kml"  # adjust if needed

shelf_station_max <- 60
concave_ratio     <- 0.4
zoom_pad          <- 0.18

# Increase this if red border lines still hug the coast.
# Decrease if too much of the desired border disappears.
coast_clip_km <- 6

# Remove tiny leftover red line fragments after clipping.
# Increase if you still see small hanging red pieces.
min_border_segment_km <- 8

ocean_col  <- "#CFE3F2"
land_col   <- "#E8DEC3"
grid_col   <- "grey55"
border_col <- "#B23A48"
bubble_col <- "#2C5F8D"

core_lines <- c(76.7, 80.0, 81.8, 83.3, 86.7, 90.0, 93.3)

# ---- 1. Build land layer: US + Mexico ---------------------------------
# This fixes the issue where Baja/Mexico reads as water.

# ---- Label coordinates (lon/lat → UTM) --------------------------------

# Helper: convert a lon/lat point to UTM (matching utm_crs)
to_utm <- function(lon, lat, crs) {
  pt <- st_sfc(st_point(c(lon, lat)), crs = 4326) %>%
    st_transform(crs)
  coords <- st_coordinates(pt)
  list(X_m = coords[1], Y_m = coords[2])
}

ca  <- to_utm(-118, 34.8, utm_crs)   # central coast / LA area
mex <- to_utm(-117, 32.4, utm_crs)   # northern Baja

california_xy <- tibble(
  X_m   = ca$X_m,
  Y_m   = ca$Y_m,
  label = "California"
)

mexico_xy <- tibble(
  X_m   = mex$X_m,
  Y_m   = mex$Y_m,
  label = "Mexico"
)

land_ne <- ne_countries(
  country = c("United States of America", "Mexico"),
  scale = 10,
  returnclass = "sf"
) %>%
  st_make_valid() %>%
  st_transform(utm_crs)

# If you want to keep the old object name:
ca_coast <- land_ne

# ---- 2. Parse + clean CalCOFI KML --------------------------------------

stations <- st_read(kml_path, quiet = TRUE) %>%
  st_zm(drop = TRUE) %>%
  mutate(
    nm   = as.character(Name),
    line = suppressWarnings(as.numeric(str_extract(nm, "^[0-9]+\\.[0-9]+"))),
    stn  = suppressWarnings(as.numeric(str_extract(nm, "(?<=\\s)[0-9]+\\.[0-9]+")))
  ) %>%
  filter(!is.na(line), !is.na(stn)) %>%
  distinct(nm, .keep_all = TRUE) %>%
  filter(line %in% core_lines, stn >= 26.7, stn <= 120) %>%
  st_transform(utm_crs)

# ---- 3. CalCOFI transect lines ----------------------------------------

transects <- stations %>%
  group_by(line) %>%
  filter(n() >= 2) %>%
  arrange(stn, .by_group = TRUE) %>%
  summarise(do_union = FALSE, .groups = "drop") %>%
  st_cast("LINESTRING")

# ---- 4. Study-region outline: open C-shape -----------------------------

outline_spec <- tibble::tribble(
  ~order, ~line_target, ~stn_target,
  1,      76.7,         49.0,
  2,      76.7,         60.0,
  3,      93.3,         60.0,
  4,      93.3,         26.7
)

outline_nodes <- purrr::pmap_dfr(
  outline_spec,
  function(order, line_target, stn_target) {
    stations %>%
      filter(
        dplyr::near(line, line_target, tol = 0.05),
        dplyr::near(stn,  stn_target,  tol = 0.15)
      ) %>%
      slice(1) %>%
      mutate(order = order)
  }
) %>%
  arrange(order)

if (nrow(outline_nodes) != 4) {
  print(outline_nodes %>% st_drop_geometry())
  stop("Did not find exactly 4 outline stations. Check line/station values or tolerances.")
}

# IMPORTANT: do NOT close the loop
outline_coords <- st_coordinates(outline_nodes)

region_outline <- st_sf(
  geometry = st_sfc(
    st_linestring(outline_coords[, 1:2]),
    crs = st_crs(utm_crs)
  )
)

# ---- 5. Map extent from station bounding box ---------------------------

bbox <- st_bbox(stations)

x_range <- bbox["xmax"] - bbox["xmin"]
y_range <- bbox["ymax"] - bbox["ymin"]

xlim <- c(bbox["xmin"] - zoom_pad * x_range, bbox["xmax"] + zoom_pad * x_range)
ylim <- c(bbox["ymin"] - zoom_pad * y_range, bbox["ymax"] + zoom_pad * y_range)

p_grid <- ggplot() +
  geom_sf(
    data = ca_coast,
    inherit.aes = FALSE,
    fill = land_col,
    color = "grey55",
    linewidth = 0.45
  ) +

  # CalCOFI transect lines
  geom_sf(
    data = transects,
    inherit.aes = FALSE,
    color = "grey50",
    linewidth = 1.2,
    alpha = 0.9
  ) +

  # CalCOFI stations
  geom_sf(
    data = stations,
    inherit.aes = FALSE,
    shape = 21,
    size = 7,
    stroke = 0.65,
    fill = "white",
    color = "grey10"
  ) +

  # Shelf-restricted study-region outline: open C-shape
  geom_sf(
    data = region_outline,
    inherit.aes = FALSE,
    color = border_col,
    linewidth = 2,
    lineend = "round",
    linejoin = "round"
  ) +

  geom_text(
    data = california_xy,
    aes(x = X_m, y = Y_m, label = label),
    size = 7.0,
    fontface = "italic",
    color = "grey15",
    hjust = 0
  ) +

  geom_text(
    data = mexico_xy,
    aes(x = X_m, y = Y_m, label = label),
    size = 7,
    fontface = "italic",
    color = "grey15",
    hjust = 0
  ) +

  coord_sf(
    crs = st_crs(utm_crs),
    xlim = xlim,
    ylim = ylim,
    expand = FALSE
  ) +

  labs(
    title = "CalCOFI core grid and SCB-restricted study region",
    subtitle = "Core CalCOFI stations shown; study region restricted to nearshore-associated stations",
    x = NULL,
    y = NULL
  ) +

  theme_minimal(base_size = 18) +
  theme(
    panel.background = element_rect(fill = ocean_col, color = NA),
    panel.grid.major = element_line(color = "grey82", linewidth = 0.45),
    panel.grid.minor = element_blank(),

    panel.border = element_rect(
      color = "grey75",
      fill = NA,
      linewidth = 1.2
    ),
    
    axis.text = element_text(size = 17, color = "grey20"),
    axis.title = element_blank(),

    legend.position = "none",

    plot.title = element_text(
      face = "bold",
      size = 28,
      color = "black",
      margin = margin(b = 8)
    ),
    plot.subtitle = element_text(
      color = "grey30",
      size = 20,
      margin = margin(b = 14)
    ),
    plot.margin = margin(16, 18, 16, 18)
  )

p_grid

ggsave(
  "figures/calcofi-core-grid-study-region-poster.png",
  p_grid,
  width = 16,
  height = 11,
  units = "in",
  dpi = 300
)

```

```{r index vs SSB}
library(dplyr)
library(ggplot2)
library(scales)

plot_df <- ss_dln3_combined_obs %>%
  filter(year >= 1981, year <= 2015)

scale_factor_ssb_dln3 <- max(plot_df$larvae_index, na.rm = TRUE) /
  max(plot_df$ssb, na.rm = TRUE)

p_index_ssb <- ggplot(plot_df, aes(x = year)) +
  # SSB ribbon + line (scaled to primary axis)
  geom_ribbon(
    aes(
      ymin = ssb_lo * scale_factor_ssb_dln3,
      ymax = ssb_hi * scale_factor_ssb_dln3
    ),
    fill = "coral4",
    alpha = 0.18
  ) +
  geom_line(
    aes(y = ssb * scale_factor_ssb_dln3),
    color = "coral4",
    linewidth = 1.0
  ) +

  # Larval index ribbon + line
  geom_ribbon(
    aes(ymin = lwr, ymax = upr),
    fill = "dodgerblue4",
    alpha = 0.20
  ) +
  geom_line(
    aes(y = larvae_index),
    color = "dodgerblue4",
    linewidth = 1.0
  ) +

  scale_x_continuous(
    breaks = seq(1984, 2015, by = 4),
    expand = expansion(mult = c(0.01, 0.03))
  ) +
  scale_y_continuous(
    name = "Larval abundance index",
    labels = comma,
    sec.axis = sec_axis(
      ~ . / scale_factor_ssb_dln3,
      name = "Spawning stock biomass (mt)",
      labels = comma
    )
  ) +
  labs(
    title = "Estimated larval abundance index and spawning stock biomass",
    subtitle = "Southern California cabezon, 1984–2015; ribbons show 95% uncertainty intervals",
    x = NULL
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 18,
      lineheight = 0.95,
      margin = margin(b = 8)
    ),
    plot.subtitle = element_text(
      size = 17,
      color = "grey30",
      linheight = 1.0,
      margin = margin(b = 14)
    ),
    axis.title.y = element_text(
      size = 19,
      face = "bold",
      color = "dodgerblue4",
      margin = margin(r = 12)
    ),
    axis.title.y.right = element_text(
      size = 19,
      face = "bold",
      color = "coral4",
      margin = margin(l = 12)
    ),

    axis.text.x = element_text(size = 16, color = "grey25"),
    axis.text.y = element_text(size = 16, color = "dodgerblue4"),
    axis.text.y.right = element_text(size = 16, color = "coral4"),

    panel.grid.major = element_line(color = "grey86", linewidth = 0.45),
    panel.grid.minor = element_blank(),

    panel.border = element_rect(
      color = "grey78",
      fill = NA,
      linewidth = 0.45
    ),

    plot.margin = margin(16, 18, 16, 18)
  )

p_index_ssb

ggsave(
  "C:/Users/machs/SynologyDrive/Dan's Files/SIO/0_capstone/6_figures/idx-vs-ssb.png",
  p_index_ssb,
  width = 12.5,
  height = 6.5,
  units = "in",
  dpi = 300,
  bg = "white"
)
```

```{r index model comparison table}

library(flextable)

aic_table <- data.frame(
  Label = c("DLn0", "DLn1", "DLn2", "DLn3", "DLn4", "DLn5", "DLn6", "DLn7"),
  Intercept = c("Global", "Global", "Global", "Global",
                "Year FE", "Year FE", "Year FE", "Global"),
  Spatial = c("on", "on", "on", "on", "on", "on", "on", "on"),
  ST_binomial = c("off", "AR1", "AR1", "IID", "off", "IID", "IID", "AR1"),
  ST_lognormal = c("off", "AR1", "IID", "off", "off", "IID", "off", "off"),
  AIC = c(
    AIC(ss_cab_fit_dln0),
    AIC(ss_cab_fit_dln1),
    AIC(ss_cab_fit_dln2),
    AIC(ss_cab_fit_dln3),
    AIC(ss_cab_fit_dln4),
    AIC(ss_cab_fit_dln5),
    AIC(ss_cab_fit_dln6),
    AIC(ss_cab_fit_dln7)
  )
)

# Compute ΔAIC and sort
aic_table$delta_AIC <- aic_table$AIC - min(aic_table$AIC)
aic_table <- aic_table[order(aic_table$AIC), ]

# Round for display
aic_table$AIC <- round(aic_table$AIC, 1)
aic_table$delta_AIC <- round(aic_table$delta_AIC, 1)

# Build the flextable
ft <- flextable(aic_table) |>
  set_header_labels(
    Label = "Model",
    Intercept = "Intercept",
    Spatial = "Spatial RF",
    ST_binomial = "ST RF (binomial)",
    ST_lognormal = "ST RF (lognormal)",
    AIC = "AIC",
    delta_AIC = "\u0394AIC"
  ) |>
  align(j = c("AIC", "delta_AIC"), align = "right", part = "all") |>
  align(j = c("Intercept", "Spatial", "ST_binomial", "ST_lognormal"),
        align = "center", part = "all") |>
  bold(i = 1) |>  # bold the top (best) row
  add_footer_lines(
    values = c(
      "Model: candidate model identifier. Intercept: global intercept (\"Global\") or fixed year effects (\"Year FE\"). Spatial RF: spatial random field (always on both components). ST RF: spatiotemporal random field structure on the binomial (presence) and lognormal (positive density) components. \u0394AIC: difference from best-fit (lowest-AIC) model. All models fit using sdmTMB (Anderson et al., 2024) with delta-lognormal family."
    )
  ) |>
  autofit() |>
  fontsize(size = 10, part = "all") |>
  fontsize(size = 9, part = "footer")

ft

save_as_docx(ft, path = "C:/Users/machs/SynologyDrive/Dan's Files/SIO/0_capstone/6_figures/model_selection_table.docx")


```


# Additional Models

```{r power analysis on ccfs}
# =====================================================================
# Power analysis for the sieve-bootstrap CCF (funtimes::ccf_boot)
# Plants a known lagged correlation into surrogate series matched to the
# AR structure of YOUR real series, then asks how often ccf_boot detects it.
#
# Fixes vs. first draft (all verified against funtimes source):
#   * `level` is the confidence-level arg (NOT `cl`; `cl` = number of cores).
#   * Result columns are r_S / lower_S / upper_S (Spearman, matching the
#     main analysis) -- not rP/lower/upper.
#   * ccf_boot calls ARest() internally with default ar.method = "HVK",
#     which can return NON-STATIONARY coefficients on short surrogate draws,
#     crashing arima.sim. ar.method = "yw" (Yule-Walker) is always stationary
#     and fixes the crash. This is a minor, conservative deviation from the
#     HVK default; see note (B) at the bottom.
#   * cl = 1L forces sequential bootstrap so ccf_boot does NOT spawn/destroy
#     a parallel cluster on every call (which would be ruinously slow here).
# =====================================================================
library(funtimes)

index_ts  <- ss_dln3_combined_obs$larvae_index
assess_ts <- ss_dln3_combined_obs$rec_dev

# ---- 1. Fit AR models so surrogates inherit each series' autocorrelation ----
fit_ar <- function(x) {
  x <- as.numeric(scale(x))
  m <- ar(x, order.max = 3, method = "yule-walker", aic = TRUE)
  list(phi = as.numeric(m$ar), order = m$order)
}
sim_ar <- function(phi, n, burn = 200) {
  if (length(phi) == 0) return(rnorm(n))
  as.numeric(arima.sim(list(ar = phi), n = n, n.start = burn))
}

# ---- 2. One simulated dataset with a planted lag-L correlation r ----
make_pair <- function(r, n, lag, fitS, fitI) {
  S <- as.numeric(scale(sim_ar(fitS$phi, n)))
  e <- as.numeric(scale(sim_ar(fitI$phi, n)))
  I <- numeric(n)
  if (lag >= 0) {
    idx <- (lag + 1):n
    I[idx] <- r * S[1:(n - lag)] + sqrt(1 - r^2) * e[idx]
    if (lag > 0) I[1:lag] <- e[1:lag]
  } else {
    L <- -lag; idx <- 1:(n - L)
    I[idx] <- r * S[(L + 1):n] + sqrt(1 - r^2) * e[idx]
    I[(n - L + 1):n] <- e[(n - L + 1):n]
  }
  data.frame(I = I, S = S)
}

# ---- 3. Detection via YOUR procedure: ccf_boot, check the lag-L band ----
detect <- function(pair, lag, B = 300, level = 0.95, ar.method = "yw") {
  cc <- tryCatch(
    ccf_boot(pair$I, pair$S,
             lag.max   = max(5, abs(lag)),
             plot      = "none",
             level     = level,
             B         = B,
             cl        = 1L,            # sequential: no cluster spawned per call
             ar.method = ar.method),    # forwarded to ARest(); "yw" = stationary
    error = function(e) NULL)
  if (is.null(cc)) return(NA)
  row <- which(cc$Lag == lag)
  obs <- cc$r_S[row]                    # Spearman, matching the main analysis
  (obs < cc$lower_S[row]) || (obs > cc$upper_S[row])
}

# ---- 4. Power at a given r (redraws on the rare residual failure) ----
power_at_r <- function(r, n, lag, fitS, fitI, M = 200, B = 300,
                       level = 0.95, ar.method = "yw", max_tries = 5) {
  hits <- 0L; done <- 0L; fails <- 0L
  while (done < M) {
    d <- NA; tries <- 0L
    while (is.na(d) && tries < max_tries) {
      d <- detect(make_pair(r, n, lag, fitS, fitI), lag,
                  B = B, level = level, ar.method = ar.method)
      tries <- tries + 1L
      if (is.na(d)) fails <- fails + 1L
    }
    if (is.na(d)) next
    hits <- hits + d; done <- done + 1L
  }
  if (fails > 0) message(sprintf("  (r=%.2f: %d bootstrap refit failures, redrawn)", r, fails))
  hits / M
}

# ---- 5. Sweep r, report power curve and minimum detectable r ----
run_sweep <- function(index_ts, assess_ts, lag = 0,
                      r_grid = seq(0.3, 0.8, 0.05),
                      M = 200, B = 300, level = 0.95, ar.method = "yw") {
  n    <- min(length(index_ts), length(assess_ts))
  fitI <- fit_ar(index_ts)
  fitS <- fit_ar(assess_ts)
  cat(sprintf("n = %d | index AR(%d) | assessment AR(%d), phi1 = %.2f | ar.method = %s\n",
              n, fitI$order, fitS$order,
              if (length(fitS$phi)) fitS$phi[1] else 0, ar.method))
  pw  <- sapply(r_grid, power_at_r, n = n, lag = lag,
                fitS = fitS, fitI = fitI, M = M, B = B,
                level = level, ar.method = ar.method)
  mdr <- if (any(pw >= 0.8) && any(pw < 0.8)) approx(pw, r_grid, xout = 0.8)$y else NA
  print(data.frame(r = r_grid, power = round(pw, 2)))
  cat(sprintf("Minimum detectable |r| at 80%% power: %.2f\n", mdr))
  invisible(list(grid = r_grid, power = pw, mdr = mdr))
}

# ---- 6. Run it ----
set.seed(my.seed)
run_sweep(index_ts, assess_ts, lag = 0)    # repeat per assessment quantity, and for lag = 3, etc.
```


## 10.6. Tweedie IID heatmap & index

```{r}

tw_iid_pred_grid <- expand.grid(
  X = seq(min(cabezon_shelf_spawn$X), max(cabezon_shelf_spawn$X), by = 5),
  Y = seq(min(cabezon_shelf_spawn$Y), max(cabezon_shelf_spawn$Y), by = 5)
) %>%
  tidyr::crossing(year = as.integer(sort(unique(cabezon_shelf_spawn$year)))) %>%
  mutate(fyear = as.factor(year))

tw_iid_pred_map <- predict(tw_iid_fit, newdata = tw_iid_pred_grid, return_tmb_object = TRUE)
tw_iid_pred_index <- get_index(tw_iid_pred_map, area = 1, bias_correct = TRUE)

library(dplyr)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)

utm_crs <- 32611

# Predict over full grid x year
tw_iid_pred_map <- predict(
  tw_iid_fit,
  newdata = tw_iid_pred_grid,
  type = "response"
)

# Average predictions across years at each spatial grid cell
tw_iid_avg_pred_plot <- tw_iid_pred_map |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000,
    pred = est
  ) |>
  group_by(X, Y, X_m, Y_m) |>
  summarize(
    avg_pred = mean(pred, na.rm = TRUE),
    .groups = "drop"
  )

# California coastline
ca_coast <- ne_states(
  country = "United States of America",
  returnclass = "sf"
) |>
  dplyr::filter(name == "California") |>
  st_transform(utm_crs)

# Observed-data bounds for study-region crop
obs_plot <- cabezon_shelf_spawn |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000
  )

# Average prediction heatmap across all years
ggplot() +
  geom_raster(
    data = tw_iid_avg_pred_plot,
    aes(x = X_m, y = Y_m, fill = log1p(avg_pred))
  ) +
  geom_sf(
    data = ca_coast,
    inherit.aes = FALSE,
    fill = "grey80",
    color = "grey40"
  ) +
  geom_sf(
    data = calcofi_lines,
    inherit.aes = FALSE,
    color = "grey95",
    linewidth = 0.25
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(obs_plot$X_m, na.rm = TRUE),
    ylim = range(obs_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  scale_fill_viridis_c(
    name = "Mean predicted\nlarvae / 100 m³"
  ) +
  labs(
    title = "Average predicted cabezon larval abundance across years",
    subtitle = "Predictions averaged across years at each spatial grid cell",
    x = NULL,
    y = NULL
  ) +
  theme_minimal()

ggplot(tw_iid_pred_index, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Tweedie on Global Intercept with IID (encounter) Spatiotemporal Field")

df <- dplyr::inner_join(ss_dln3_cab_index, tw_iid_pred_index, by = "year")  # est_dln, est_twd

cor(df$est.x, df$est.y, method = "spearman")        # primary: rank, scale- and outlier-robust
cor(log(df$est.x), log(df$est.y))                   # secondary: Pearson on log scale tames the spike

ccf_boot(
  x = tw_iid_pred_index$est,
  y = ss_dln3_cab_index$est,
  lag.max = 10,
  plot = "Spearman",
  B = 1000
); title(main = "CCF: Restricted Larval Index vs. Age-0 Recruits (Delta-Lognormal)")

funtimes::ccf_boot(df$est.x, df$est.y,              # consistent with your SSB methods
                   lag.max = 3, plot = "Spearman")    # read the lag-0 value + CI

# leverage check: the influential years
cor(df$est.x[df$year != 2004], df$est.y[df$year != 2004], method = "spearman")
```


##10.7. ss_cab_fit2a — Tweedie with iid spatiotemporal

```{r}
#---- 10.2.a. Calibration summaries (Tweedie; single component)

ss_fit2a_pred_obs <- predict(ss_cab_fit2a, type = "response")

# Overall mean calibration (no est1/est2 for Tweedie)
ss_fit2a_pred_obs %>%
  summarise(
    obs_mean  = mean(larvae_100m3, na.rm = TRUE),
    pred_mean = mean(est, na.rm = TRUE),
    obs_occ   = mean(larvae_100m3 > 0, na.rm = TRUE),
    # Tweedie has no explicit p(encounter); use threshold on est as a rough proxy
    pred_occ_proxy = mean(est > 0, na.rm = TRUE)
  )

#---- 10.2.b. Predicted vs. observed plots

# Raw scale
plot(
  x = ss_fit2a_pred_obs$est,
  y = cabezon_shelf_spawn$larvae_100m3,
  xlab = "Predicted",
  ylab = "Observed"
)
abline(0, 1, lty = 2)

# Log scale
plot(
  log1p(ss_fit2a_pred_obs$est),
  log1p(cabezon_shelf_spawn$larvae_100m3),
  xlab = "log(Predicted + 1)",
  ylab = "log(Observed + 1)"
)
abline(0, 1, lty = 2)

#---- 10.2.c. Randomized quantile residuals

ss_fit2a_rq_res <- residuals(ss_cab_fit2a, type = "mle-mvn")

# QQ plot
qqnorm(ss_fit2a_rq_res)
qqline(ss_fit2a_rq_res)

# Spatial residual map by year
ss_fit2a_pred_obs$rq_resid <- ss_fit2a_rq_res

ggplot(ss_fit2a_pred_obs, aes(x = longitude, y = latitude, color = rq_resid)) +
  geom_point() +
  scale_color_gradient2() +
  facet_wrap(~ year)

#---- 10.2.d. Temporal calibration

ss_fit2a_pred_obs %>%
  group_by(year) %>%
  summarize(
    obs  = mean(larvae_100m3),
    pred = mean(est)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = obs), color = "black") +
  geom_line(aes(y = pred), color = "blue", linetype = "dashed")

#---- 10.2.e. Abundance index: regular prediction grid
# fyear must be present and match levels in training data
# Note: 1982 is extra_time — fixed effect for that level is interpolated by sdmTMB

ss_fit2a_pred_grid <- expand.grid(
  X = seq(min(cabezon_shelf_spawn$X), max(cabezon_shelf_spawn$X), by = 5),
  Y = seq(min(cabezon_shelf_spawn$Y), max(cabezon_shelf_spawn$Y), by = 5)
) %>%
  tidyr::crossing(year = as.integer(sort(unique(cabezon_shelf_spawn$year)))) %>%
  mutate(fyear = as.factor(year))

ss_fit2a_cab_map <- predict(ss_cab_fit2a,
                            newdata   = ss_fit2a_pred_grid,
                            return_tmb_object = TRUE)

ss_fit2a_cab_index <- get_index(ss_fit2a_cab_map, area = 1, bias_correct = TRUE)

ggplot(ss_fit2a_cab_index, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Tweedie IID (fixed year effects, regular grid)")

#---- 10.2.f. Abundance index: observed station grid (preferred)

ss_fit2a_pred_grid_obs <- cabezon_shelf_spawn %>%
  dplyr::select(X, Y, year) %>%
  distinct() %>%
  mutate(fyear = as.factor(year))

ss_fit2a_cab_map_obs <- predict(ss_cab_fit2a,
                                newdata = ss_fit2a_pred_grid_obs,
                                return_tmb_object = TRUE)

ss_fit2a_cab_index_obs <- get_index(ss_fit2a_cab_map_obs,
                                    area = 1,
                                    bias_correct = TRUE)

ggplot(ss_fit2a_cab_index_obs, aes(x = year, y = est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
  labs(y = "Estimated abundance index", x = "Year",
       title = "Abundance Index: Tweedie IID (fixed year effects, observed grid)")


# CUSUM
ss_full_pred_obs %>%
  group_by(year) %>%
  summarise(
    med_resid1 = median(rq_resid1, na.rm = TRUE),
    med_resid2 = median(rq_resid2, na.rm = TRUE),  # positive tows only
    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    cusum1 = cumsum(replace_na(med_resid1, 0)),
    cusum2 = cumsum(replace_na(med_resid2, 0))
  ) %>%
  pivot_longer(cols = c(cusum1, cusum2),
               names_to = "component",
               values_to = "cusum") %>%
  ggplot(aes(x = year, y = cusum, color = component)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("cusum1" = "steelblue", "cusum2" = "firebrick"),
                     labels = c("Delta (binomial)", "Gamma")) +
  labs(x = "Year", y = "CUSUM", title = "CUSUM by Delta Model Component — ss_cab_fit1") +
  theme_minimal()


library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

utm_crs <- 32611

ca_coast <- ne_states(
  country = "United States of America",
  returnclass = "sf"
) |>
  dplyr::filter(name == "California") |>
  st_transform(utm_crs)

calcofi_stations <- cabezon_shelf_spawn |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000
  ) |>
  distinct(line, station, X_m, Y_m) |>
  st_as_sf(coords = c("X_m", "Y_m"), crs = utm_crs)

calcofi_lines <- calcofi_stations |>
  arrange(line, station) |>
  group_by(line) |>
  summarize(do_union = FALSE, .groups = "drop") |>
  st_cast("LINESTRING")

pred_plot <- ss_map_pred |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000
  )

# Presence + Density
ggplot() +
  geom_raster(
    data = pred_plot_tw,
    aes(x = X_m, y = Y_m, fill = est)
  ) +
  geom_sf(
    data = ca_coast,
    inherit.aes = FALSE,
    fill = "grey80",
    color = "grey40"
  ) +
  geom_sf(
    data = calcofi_lines,
    inherit.aes = FALSE,
    color = "white",
    linewidth = 0.3
  ) +
  geom_sf(
    data = calcofi_stations,
    inherit.aes = FALSE,
    size = 0.5
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(pred_plot$X_m, na.rm = TRUE),
    ylim = range(pred_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "Predicted larvae") +
  theme_minimal()

# Density component
ggplot() +
  geom_raster(
    data = pred_plot_tw,
    aes(x = X_m, y = Y_m, fill = est2)
  ) +
  geom_sf(
    data = ca_coast,
    inherit.aes = FALSE,
    fill = "grey80",
    color = "grey40"
  ) +
  geom_sf(
    data = calcofi_lines,
    inherit.aes = FALSE,
    color = "white",
    linewidth = 0.3
  ) +
  geom_sf(
    data = calcofi_stations,
    inherit.aes = FALSE,
    size = 0.5
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(pred_plot$X_m, na.rm = TRUE),
    ylim = range(pred_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  scale_fill_viridis_c(name = "Predicted larvae") +
  theme_minimal()

# Tweedie maps
ss_map_pred_tw <- predict(
  ss_cab_fit2,
  newdata = ss_fit2a_pred_grid,
  type = "response"
)

pred_plot_tw <- ss_map_pred_tw |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000
  )

# Plot showing larval observations by station over time
obs_plot_tw <- cabezon_shelf_spawn |>
  mutate(
    X_m = X * 1000,
    Y_m = Y * 1000,
    larvae_positive = larvae_100m3 > 0
  )

ggplot() +
  geom_sf(
    data = ca_coast,
    inherit.aes = FALSE,
    fill = "grey80",
    color = "grey40"
  ) +
  geom_sf(
    data = calcofi_lines,
    inherit.aes = FALSE,
    color = "grey95",
    linewidth = 0.25
  ) +
  geom_point(
    data = obs_plot_tw |> filter(!larvae_positive),
    aes(x = X_m, y = Y_m),
    size = 0.4,
    alpha = 0.25
  ) +
  geom_point(
    data = obs_plot_tw |> filter(larvae_positive),
    aes(x = X_m, y = Y_m, size = larvae_100m3),
    alpha = 0.7
  ) +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(obs_plot$X_m, na.rm = TRUE),
    ylim = range(obs_plot$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  facet_wrap(~ year) +
  scale_size_continuous(
    name = "Observed larvae / 100 m³",
    range = c(1, 7)
  ) +
  labs(
    title = "Observed cabezon larval abundance by CalCOFI station",
    x = NULL,
    y = NULL
  ) +
  theme_minimal()

# Plot showing delta + hurdle components (distribution and abundance)
ggplot() +
  geom_raster(
    data = pred_plot_tw,
    aes(x = X_m, y = Y_m, fill = est)
  ) +
  geom_sf(data = ca_coast, inherit.aes = FALSE, fill = "grey80") +
  coord_sf(
    crs = st_crs(utm_crs),
    xlim = range(pred_plot_tw$X_m, na.rm = TRUE),
    ylim = range(pred_plot_tw$Y_m, na.rm = TRUE),
    expand = FALSE
  ) +
  facet_wrap(~ year) +
  scale_fill_viridis_c(name = "Predicted larvae") +
  theme_minimal()
```

## 10.8. ss_cab_fit2a — STAR correlation analysis

```{r}
#---- 10.8. ss_cab_fit2a STAR comparisons and CCF

# Create one data frame to house all STAR + larval index data
# Note: using observed station grid index (10.2.f) as preferred basis
tw_fy_combined <- ss_fit2a_cab_index_obs %>%
  dplyr::select(year, est) %>%
  rename(larvae_index = est) %>%
  left_join(
    STAR_SSB %>% rename(ssb = value, ssb_lo = lo, ssb_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_recdevs %>% rename(rec_dev = value, rec_dev_lo = lo, rec_dev_hi = hi),
    by = "year"
  ) %>%
  left_join(
    STAR_age0 %>% rename(age0 = value, age0_lo = lo, age0_hi = hi),
    by = "year"
  )

# Filter to observed years
tw_fy_combined_obs <- tw_fy_combined %>%
  filter(!is.na(larvae_index))

# CCF: Tweedie IID larval index vs. age-0 recruits
ccf(tw_fy_combined_obs$larvae_index, tw_fy_combined_obs$age0,
    lag.max = 10,
    main = "CCF: Tweedie IID Larval Index vs. Age-0 Recruits")

# CCF: Tweedie IID larval index vs. recruitment deviations
ccf(tw_fy_combined_obs$larvae_index, tw_fy_combined_obs$rec_dev,
    lag.max = 10,
    main = "CCF: Tweedie IID Larval Index vs. Recruitment Deviations")

# CCF: Tweedie IID larval index vs. SSB (raw)
# Note: shared declining trend may inflate signal — detrended version below is preferred
ccf(tw_fy_combined_obs$larvae_index, tw_fy_combined_obs$ssb,
    lag.max = 10,
    main = "CCF: Tweedie IID Larval Index vs. Spawning Stock Biomass (raw)")

# Detrend both series via loess before comparing against SSB
tw_fy_larvae_detrended <- residuals(loess(larvae_index ~ year, data = tw_fy_combined_obs, span = 0.75))
tw_fy_ssb_detrended    <- residuals(loess(ssb ~ year,          data = tw_fy_combined_obs, span = 0.75))

ccf(tw_fy_larvae_detrended, tw_fy_ssb_detrended,
    lag.max = 10,
    main = "CCF: Tweedie IID Larval Index vs. SSB (detrended)")
```

## 10.9. ss_cab_fit2a — Correlation analysis

```{r}
#---- 10.9. ss_cab_fit2a diagnostics and lagged regression

# RQ residuals: annual distributions
ss_fit2a_pred_obs %>%
  ggplot(aes(x = factor(year), y = rq_resid)) +
  geom_boxplot(outlier.size = 0.6, fill = "steelblue", alpha = 0.4) +
  geom_smooth(
    aes(x = as.numeric(factor(year)), y = rq_resid),
    method    = "loess", span = 0.4,
    se        = TRUE, color = "firebrick", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Year",
    y     = "RQ Residual",
    title = "Annual RQ Residual Distributions — ss_cab_fit2a"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

# CUSUM of median RQ residuals (single component; Tweedie)
ss_fit2a_pred_obs %>%
  group_by(year) %>%
  summarise(med_resid = median(rq_resid), .groups = "drop") %>%
  arrange(year) %>%
  mutate(cusum = cumsum(med_resid)) %>%
  ggplot(aes(x = year, y = cusum)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x     = "Year",
    y     = "Cumulative Sum of Median RQ Residual",
    title = "CUSUM — ss_cab_fit2a"
  ) +
  theme_minimal()

#---- Simple lagged regression at lag -1

tw_fy_combined_lagged <- tw_fy_combined_obs %>%
  arrange(year) %>%
  mutate(
    age0_lead1    = lead(age0, 1),
    rec_dev_lead1 = lead(rec_dev, 1)
  )

tw_fy_lm_age0   <- lm(age0_lead1    ~ larvae_index, data = tw_fy_combined_lagged)
tw_fy_lm_recdev <- lm(rec_dev_lead1 ~ larvae_index, data = tw_fy_combined_lagged)

summary(tw_fy_lm_age0)
summary(tw_fy_lm_recdev)

tw_fy_combined_lagged %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

# Sensitivity: remove 1983
tw_fy_combined_lagged_no83 <- tw_fy_combined_lagged %>% filter(year != 1983)
tw_fy_lm_age0_no83 <- lm(age0_lead1 ~ larvae_index, data = tw_fy_combined_lagged_no83)
summary(tw_fy_lm_age0_no83)

# Detrended regressions
tw_fy_lm_age0_detrend   <- lm(age0_lead1    ~ larvae_index + year, data = tw_fy_combined_lagged)
tw_fy_lm_recdev_detrend <- lm(rec_dev_lead1 ~ larvae_index + year, data = tw_fy_combined_lagged)
summary(tw_fy_lm_age0_detrend)
summary(tw_fy_lm_recdev_detrend)

# Cook's distance
plot(tw_fy_lm_age0, which = 4)

# Trimmed analysis: remove three most anomalous years
tw_fy_combined_lagged_trimmed <- tw_fy_combined_lagged %>%
  filter(!year %in% c(1983, 1986, 1998))

tw_fy_combined_lagged_trimmed %>%
  filter(!is.na(age0_lead1)) %>%
  ggplot(aes(x = larvae_index, y = age0_lead1, label = year)) +
  geom_point() +
  geom_text(nudge_y = 8, size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Larval Index (year t)", y = "Age-0 Recruits (year t+1)") +
  theme_minimal()

tw_fy_lm_age0_trimmed <- lm(age0_lead1 ~ larvae_index, data = tw_fy_combined_lagged_trimmed)
summary(tw_fy_lm_age0_trimmed)

tw_fy_lm_age0_trimmed_detrend <- lm(age0_lead1 ~ larvae_index + year, data = tw_fy_combined_lagged_trimmed)
summary(tw_fy_lm_age0_trimmed_detrend)

# Is underprediction spatially structured?
ss_fit2a_pred_obs %>%
  group_by(station) %>%
  summarise(mean_resid = mean(rq_resid), .groups = "drop") %>%
  ggplot(aes(x = station, y = mean_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = TRUE) +
  labs(x = "Station", y = "Mean RQ Residual",
       title = "Mean Residual by Station — ss_cab_fit2a") +
  theme_minimal()

# AIC comparison: Tweedie variants
AIC(ss_cab_fit2, ss_cab_fit2a)
```


